---
title: "F04 association — pathways, snapshot view"
subtitle: "Which whole pathways track the training response"
author: "A_HRvLR_2026"
format:
  html:
    toc: true
    embed-resources: true
    df-print: paged
execute:
  eval: true
  echo: false
  warning: false
  message: false
---

## In plain terms

Boil each person down to one number — their pathway activity at baseline,
or how much it changed after training — and see which whole pathways line
up with who gained the most muscle.

## Method

Proteins are first collapsed to one row per gene (the highest-mean row) and
scored per sample against a shared pathway collection with `singscore`, a
rank-based single-sample method. Each pathway score is then collapsed to
one value per subject in each of three windows — baseline (T1), training
change (T2 − T1), and acute-bout change (T3 − T2) — and regressed on each
continuous training response with `limma::lmFit` and `eBayes` moderation,
exactly as in the protein-level snapshot. Every pathway-by-trait-by-phase
cell is a separate test, and p-values are corrected with
Benjamini-Hochberg across pathways within each trait/phase family.
Descriptive, exploratory, n = 16.

## Results

```{r}
#| label: load-results
pacman::p_load(here, openxlsx, dplyr)

assoc <- read.xlsx(
  here(
    "04_Figures", "F04_association", "limma", "pathways", "c_data",
    "pathways_source_data.xlsx"
  ),
  "pathway_association"
)

n_features <- n_distinct(assoc$feature)
n_traits <- n_distinct(assoc$trait)
n_phases <- n_distinct(assoc$phase)
n_tests <- nrow(assoc)
n_hits <- sum(assoc$bh < 0.05, na.rm = TRUE)
min_bh <- min(assoc$bh, na.rm = TRUE)
```

`r n_features` pathways were tested against `r n_traits` training-response
traits across `r n_phases` phases (baseline, training, acute), for
`r n_tests` pathway-by-trait-by-phase cells in total. `r n_hits` of those
cells pass BH < 0.05; the smallest BH-adjusted q anywhere in the table is
`r sprintf("%.3f", min_bh)`. No pathway's snapshot score survives
correction for tracking any training response.

```{r sessioninfo}
sessionInfo()
```
