---
title: "HRvLR 03 — Differential Expression"
subtitle: "Design, contrasts, model fit, and the Pi-score, on the non-imputed matrix"
author: "A_HRvLR_2026"
date: today
bibliography: references.bib
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
    embed-resources: true
    df-print: paged
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(here)
knitr::opts_knit$set(root.dir = here::here())
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(readr)
})
```

This document reproduces the primary DEP stage and shows the result of each step.
It runs the same proteoDA workflow as `03_DEP/a_non_imputed/a_script/01_run_dep.R`:
build the design, define the contrasts, fit limma with `duplicateCorrelation`, and
score each protein with the Pi-score. The committed results are read back at the
points where a step needs the fitted object.

The central question of the study is whether high responders (HR) and low
responders (LR) differ in their proteomic response to training. The design is a
2×3 factorial — responder (HR/LR) × timepoint (T1 baseline, T2 trained, T3 acute)
— with repeated measures on subject.

# Why limma on the non-imputed matrix {#sec-why}

`limma` fits a per-protein linear model and stabilizes each protein's variance by
borrowing strength across all proteins (empirical Bayes moderation) [@ritchie2015].
That shrinkage matters here, where each cell has only seven or eight subjects —
too few for a stable per-protein variance on its own.

Peng et al. (2024) show that limma applied directly to data with missing values,
fitting each per-protein model on the observations that exist, outperforms
imputing first [@peng2024]. Imputation adds distributional assumptions that can
distort variance and inflate precision. So the **non-imputed** normalized matrix
is the primary input; the imputed matrices are run separately as a comparison
(`b_imputed/a_script/01_run_dep_imputed.R`).

# Load data and metadata {#sec-load}

The normalized matrix comes from stage 02; the canonical metadata comes from the
normalization `DAList` rather than from parsing sample names, so the factor levels
for `Group`, `Timepoint`, and `Group_Time` match every other stage.

```{r}
#| label: show-metadata
library(proteoDA)
dal_norm <- readRDS(here("02_Normalization", "c_data", "DAList_normalized.rds"))
table(dal_norm$metadata$Group, dal_norm$metadata$Timepoint)
cat(sprintf("Proteins: %d | Samples: %d\n", nrow(dal_norm$data), ncol(dal_norm$data)))
```

Each cell holds seven or eight subjects; a cell below eight reflects an outlier
removed in stage 01. The runner rebuilds a tidy metadata frame with `group`
(= `Group_Time`) and `subject` (= `Subject_ID`) and guards that the matrix columns
match the sample sheet before any model is fit:

```{r}
#| label: dep-load
#| eval: false
df <- read_csv(here("02_Normalization", "c_data", "normalized.csv"), show_col_types = FALSE)
ann_cols <- c("uniprot_id", "protein", "gene", "description")
mat <- as.matrix(df[, setdiff(names(df), ann_cols)])
rownames(mat) <- df$uniprot_id

dal_meta <- as.data.frame(dal_norm$metadata)
meta <- tibble(
  sample_id = dal_meta$Col_ID,
  responder = factor(dal_meta$Group, levels = c("HR", "LR")),
  time = factor(dal_meta$Timepoint, levels = c("T1", "T2", "T3")),
  group = factor(dal_meta$Group_Time,
    levels = c("HR_T1", "HR_T2", "HR_T3", "LR_T1", "LR_T2", "LR_T3")
  ),
  subject = dal_meta$Subject_ID
)
stopifnot(setequal(colnames(mat), meta$sample_id))
```

# Statistical design {#sec-design}

## Cell-means parametrization

The model uses a cell-means formula (`~ 0 + group`): no intercept, one coefficient
per `Group_Time` cell. Every planned contrast is then a plain arithmetic
expression over the six cell means, which keeps the contrasts easy to read and
audit.

## Within-subject correlation

Each subject contributes three observations (T1, T2, T3). Treating them as
independent would overstate the degrees of freedom and produce anti-conservative
p-values. `duplicateCorrelation` [@smyth2005] estimates one average within-subject
correlation, `rho`, by pooling across all proteins, and uses it as a fixed
blocking correlation in the per-protein fits. Pooling keeps `rho` stable at small
n; combined with `eBayes` moderation it is more powerful than a naive
repeated-measures ANOVA here.

```{r}
#| label: dep-design
#| eval: false
dal <- add_design(dal, "~ 0 + group + (1 | subject)")
colnames(dal$design$design_matrix) <- gsub("^group", "", colnames(dal$design$design_matrix))
```

```{r}
#| label: show-within-cor
dal_dep <- readRDS(here("03_DEP", "a_non_imputed", "c_data", "01_limma_DAList.rds"))
within_cor <- dal_dep$eBayes_fit$correlation
if (is.null(within_cor)) within_cor <- NA_real_
cat(sprintf("Within-subject correlation (rho): %.3f\n", within_cor))
```

A positive `rho` means subjects are consistent across timepoints, and modelling
that structure buys power. A `rho` near zero would mean the repeated-measures
structure adds little.

# Contrasts {#sec-contrasts}

Nine contrasts are pre-specified. The two interaction contrasts are the analytic
core: they ask whether the response to training (or to the acute bout) differs
between HR and LR.

| Contrast | Formula | Question |
|:--|:--|:--|
| Training_HR | HR_T2 − HR_T1 | Do HR change after training? |
| Training_LR | LR_T2 − LR_T1 | Do LR change after training? |
| Acute_HR | HR_T3 − HR_T2 | Do HR respond to the acute bout? |
| Acute_LR | LR_T3 − LR_T2 | Do LR respond to the acute bout? |
| Baseline_HRvLR | HR_T1 − LR_T1 | Do HR and LR differ at baseline? |
| Trained_HRvLR | HR_T2 − LR_T2 | …after training? |
| Acute_HRvLR | HR_T3 − LR_T3 | …after the acute bout? |
| Training_Interaction | (HR_T2 − HR_T1) − (LR_T2 − LR_T1) | Does the training response differ by group? |
| Acute_Interaction | (HR_T3 − HR_T2) − (LR_T3 − LR_T2) | Does the acute response differ by group? |

```{r}
#| label: dep-contrasts
#| eval: false
source(here("03_DEP", "contrasts.R"))
dal <- add_contrasts(dal, contrasts_vector = HRVLR_CONTRASTS)
```

Each interaction is a difference of differences: it subtracts the LR change from
the HR change, so a non-zero value means the two groups moved by different
amounts.

# Fit and extract {#sec-fit}

`fit_limma_model()` chains `lmFit` (per-protein fit with the blocking `rho`),
`contrasts.fit` (reparametrize into contrast space), and `eBayes` with
`robust = TRUE` (outlier-resistant variance moderation) [@ritchie2015]. `extract_DA_results()` applies Benjamini-Hochberg
across proteins [@benjamini1995]. The FDR threshold is **0.10**: at n ≈ 8 per
cell, strict 0.05 control would discard genuinely altered proteins with modest
support; the Pi-score adds a fold-change requirement on top.

```{r}
#| label: dep-fit
#| eval: false
dal <- fit_limma_model(dal)
dal <- extract_DA_results(dal, pval_thresh = 0.10, lfc_thresh = 0, adj_method = "BH")
```

```{r}
#| label: show-da-summary
da_sum <- read_csv(here("03_DEP", "a_non_imputed", "c_data", "02_DA_summary.csv"),
  show_col_types = FALSE
)
da_sum |>
  filter(type != "nonsig") |>
  select(contrast, type, any_of(c("sig.Pi", "sig.FDR.05", "sig.FDR.10")))
```

Where the FDR and Pi counts are low, that is expected for a small-n DIA study: the
analysis is powered for large fold changes, and the interaction contrasts test a
subtle hypothesis. The figure layer summarizes effect sizes (bootstrap CIs, fry,
RRHO2) that stay interpretable regardless of per-protein significance.

# Pi-score {#sec-pi}

The Pi-score [@xiao2014] folds significance and magnitude into one value:

$$\text{Pi} = p^{|\log_2 \text{FC}|}$$

A large fold change drives Pi below the raw p-value; a small one pushes it above.
A protein must show both a real change and statistical support to clear Pi < 0.05.

```{r}
#| label: pi-score
#| eval: false
res <- res |>
  mutate(
    pi_score = P.Value^abs(logFC),
    sig_pi = case_when(
      pi_score < 0.05 & logFC > 0 ~ 1L,
      pi_score < 0.05 & logFC < 0 ~ -1L,
      TRUE ~ 0L
    )
  )
```

```{r}
#| label: show-per-contrast
source(here("03_DEP", "contrasts.R"))
f <- here("03_DEP", "a_non_imputed", "c_data", "04_per_contrast_results", "Training_HR.csv")
if (file.exists(f)) {
  read_csv(f, show_col_types = FALSE) |>
    mutate(
      pi_score = P.Value^abs(logFC),
      sig_pi = case_when(
        pi_score < PI_THRESH & logFC > 0 ~ 1L,
        pi_score < PI_THRESH & logFC < 0 ~ -1L,
        TRUE ~ 0L
      )
    ) |>
    arrange(pi_score) |>
    select(gene, logFC, P.Value, adj.P.Val, pi_score, sig_pi) |>
    head(5)
}
```

# Combined results {#sec-combined}

All nine contrasts are assembled into one wide table, one row per protein with
contrast-suffixed columns (`logFC_Training_HR`, `adj.P.Val_Training_HR`, …). This
is the primary hand-off to the figure layer.

```{r}
#| label: show-combined
f <- here("03_DEP", "a_non_imputed", "c_data", "03_combined_results.csv")
if (file.exists(f)) {
  comb <- read_csv(f, show_col_types = FALSE)
  cat(sprintf("Combined results: %d proteins x %d columns\n", nrow(comb), ncol(comb)))
  comb |>
    select(
      uniprot_id, gene, starts_with("logFC_Training_HR"),
      starts_with("pi_score_Training_HR")
    ) |>
    head(4)
}
```

# Imputed comparison and figures {#sec-imputed}

`b_imputed/a_script/01_run_dep_imputed.R` repeats this exact fit on the three
imputed matrices (imp4p, MsCoreUtils, missForest) and reports the Spearman
concordance of each arm's logFC against the primary, non-imputed fit. High
concordance confirms the missing-value strategy does not reshape the effects
[@peng2024]. The HR-vs-LR response comparison and the per-contrast volcanoes live
in the figure layer (F03–F05); proteoDA also writes its own volcano and MA plots
into `a_non_imputed/b_reports/` during the primary run.

# Stage outputs {#sec-outputs}

| File | Description |
|:--|:--|
| `c_data/01_limma_DAList.rds` | Fitted `proteoDA` DAList with the eBayes model |
| `c_data/02_DA_summary.csv` | Significance counts per contrast, direction, and criterion |
| `c_data/03_combined_results.csv` | Primary output: nine contrasts, wide format |
| `c_data/04_per_contrast_results/` | One CSV per contrast (logFC, t, P.Value, adj.P.Val, pi_score, sig_pi) |
| `c_data/05_results.xlsx` | Workbook, one sheet per contrast plus the summary |
| `../b_imputed/c_data/` | Imputed-arm results and `logfc_concordance.csv` |

Stage 04 reads `03_combined_results.csv` and the fitted DAList for pathway
enrichment and figures.

# Session info {#sec-session}

```{r}
#| label: session-info
sessionInfo()
```

# References {#sec-refs}

::: {#refs}
:::
