---
title: "High vs Low Responders (HRvLR) Skeletal-Muscle Proteomics"
subtitle: "A ground-up walkthrough of the proteoDA-native pipeline: filtering → normalization → differential expression → figures"
author: "A_HRvLR_2026"
date: today
bibliography: 04_Figures/shared/references.bib
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: true
    code-tools: true
    theme: cosmo
    fig-cap-location: bottom
    embed-resources: true
execute:
  eval: false      # documentation build: code is shown, not re-run; outputs are frozen
  echo: true
  warning: false
  message: false
---

> **How to read this document.** Every code block is the *actual* code from the
> pipeline scripts, shown for provenance but **not re-executed** (`eval: false`).
> The π counts in @sec-pi-warning are the exception: they are read live from the
> committed DEP results, because when they were frozen by hand they drifted away
> from the fit and understated the problem. To reproduce, run the stage scripts in
> order. Paths are relative to the project root (`A_HRvLR_2026/`). Each stage also
> ships a live, self-verifying tutorial (`01_filtering.qmd`,
> `02_normalization.qmd`, `03_dep.qmd`).

# Overview {#sec-overview}

Resistance training produces a wide range of hypertrophic outcomes across people.
This pipeline asks whether that range is visible in the skeletal-muscle proteome:
do **high responders (HR)** and **low responders (LR)** differ in how their
proteomes move with training, and with an acute bout?

The design is a 2×3 factorial — responder (HR/LR) × timepoint — with repeated
measures on subject. Each subject is sampled three times: T1 (baseline), T2
(trained, 72 h post a training block), and T3 (acute, 1 h post a final bout). The
grouping factor `Group_Time` carries the six cells.

| `Group_Time` | Meaning | n samples |
|---|---|---|
| `HR_T1` … `HR_T3` | High responders, three timepoints | 7 each |
| `LR_T1` … `LR_T3` | Low responders, three timepoints | 8 each |

Nine contrasts are built on this scheme:

| Contrast | Definition | Question |
|---|---|---|
| `Training_HR` / `Training_LR` | `*_T2 − *_T1` | Training response within each group |
| `Acute_HR` / `Acute_LR` | `*_T3 − *_T2` | Acute-bout response within each group |
| `Baseline_HRvLR` | `HR_T1 − LR_T1` | Group difference at baseline |
| `Trained_HRvLR` / `Acute_HRvLR` | `HR_T2 − LR_T2` / `HR_T3 − LR_T3` | Group difference after training / acute |
| `Training_Interaction` | `(HR_T2 − HR_T1) − (LR_T2 − LR_T1)` | Does the training response differ by group? |
| `Acute_Interaction` | `(HR_T3 − HR_T2) − (LR_T3 − LR_T2)` | Does the acute response differ by group? |

The interaction contrasts are the analytic core: they ask the responder question
directly.

# What the pipeline found {#sec-found}

Read this before any figure. The headline is a **negative result**, and the figures
are easy to over-read without it.

**The phenotype is real.** HR and LR did the same training work and grew apart:
HR gained muscle and fibre area, LR did not, while strength gains matched (F01).
That is the premise the rest of the pipeline tries to explain.

**No global proteome difference survives.** F02 asks six independent ways — state
(PERMANOVA p = 0.62), response magnitude and direction (RRPP, all p ≥ 0.38),
heterogeneity (PERMDISP), design-explained variance (ω² ≈ 0), and even a *supervised*
axis built to separate the groups (CAP p ≈ 0.22, 40% leave-one-out error). All six
are null.

**The HR and LR responses are only weakly concordant, and no set-level test is
significant.** The pilot concordance is weak in both phases: Spearman ρ = 0.20
in training, ρ = 0.15 in the acute bout. The fry
rotation tests — the only properly blocked, error-rate-controlled set tests in the
pipeline — are null in every direction (best p = 0.087).

**Pathway enrichment is where the signal lives.** 691 of 14,088 pathway × contrast
tests are BH-significant (F03), concentrated acutely, and HR's acute response enriches
217 pathways against LR's 33. This is genuinely FDR-controlled and is the strongest
HR/LR asymmetry in the study. Note `Baseline_HRvLR` also returns 65 pathways, so the
groups are not identical at baseline.

**F05-F06 screen 945 cells and report a permutation p, not a q.** F05
classification (153 cells) and F06 prediction (792 cells) each report
a metric, a permutation p over B = 200, the screen size, and a leakage label —
there is no BH q in either, unlike F01-F03's limma BH-FDR, because a screen this
correlated (shared subjects, overlapping feature spaces, nested timepoint
configs) has no defensible multiple-comparison correction. A cell counts as a
**lead** only when `perm_p < .05` *and* the metric beats the trivial baseline
(`q2 > 0`, `auc > 0.5`); a collapsed permutation null can otherwise reach its p
floor while predicting worse than the group mean.

F05 classification is a complete null: 0 leads of 153. F06 prediction clears 32
leads of 792 (4.0%, below the 5% chance rate), yet 26 of the 32 concentrate on
`d_mcsa`. The module arm does not predict: all 8 module leads fall below
zero, median drop 0.461. The decomposition matters: 0.428 of that comes from
restricting the features to the two modules recoverable in every fold, and
0.000 from rebuilding the network in-fold. The leads depended on the nine
modules that do not reproduce under subsampling. HR and LR share module
architecture (preservation strong both directions, using cohort-defined labels,
so not itself free of circularity), and training- and acute-only contrast networks were tested and are
not viable at this cohort size (`shared/WGCNA`).

**So: the proteome does not separate high from low responders at this sample size.**
It is a null study with a strong phenotype and a pathway-level lead. Saying that
plainly is more defensible than any of the individual panels.

## One caveat that changes how you read almost every figure {#sec-pi-warning}

Most "significant" protein counts in F03_pathway/supp and F02 — the 55 divergent in training,
the 86 acute, the DEP-count bars — come from the **π gate**, `p^|logFC| < 0.05`
[@xiao2014], **not from BH-FDR**.

```{r}
#| label: pi-stats
#| eval: true
#| echo: false
# Read from the committed DEP results rather than hardcoding. These counts were quoted as
# 489 / 12 for months after the fit moved underneath them.
pi <- local({
  res <- readr::read_csv(
    here::here("03_DEP", "a_non_imputed", "c_data", "03_combined_results.csv"),
    show_col_types = FALSE
  )
  cn <- sub("^pi_score_", "", grep("^pi_score_", names(res), value = TRUE))
  calls <- purrr::map_dfr(cn, \(ct) tibble::tibble(
    pi = res[[paste0("pi_score_", ct)]],
    p = res[[paste0("P.Value_", ct)]],
    q = res[[paste0("adj.P.Val_", ct)]]
  ))
  sel <- calls |> dplyr::filter(!is.na(pi), pi < 0.05)
  list(
    n_pi = nrow(sel),
    n_pi_nonsig = sum(sel$p >= 0.05),
    pct_nonsig = round(100 * mean(sel$p >= 0.05), 1),
    worst_p = round(max(sel$p), 3),
    n_bh05 = sum(calls$q < 0.05, na.rm = TRUE),
    n_bh10 = sum(calls$q < 0.10, na.rm = TRUE)
  )
})
```

π controls no error rate. Because a large fold change drives the score down on its
own, fold change alone can buy admission: `r pi$n_pi_nonsig` of `r pi$n_pi` π-calls
(`r pi$pct_nonsig`%) have a raw p ≥ 0.05, the largest being **p = `r pi$worst_p`**.
Across the nine contrasts there are **`r pi$n_pi` π-calls against `r pi$n_bh05`
proteins at BH < 0.05** (`r pi$n_bh10` at BH < 0.10) — a
`r round(pi$n_pi / pi$n_bh05)`× inflation. The π set is not even a subset of the
nominally significant set.

Treat every π-derived count as a **ranked, hypothesis-generating list**, never as an
FDR-controlled discovery. The fgsea ring arcs in F03 and the GO counts in F02 *are*
BH-controlled and can be read as such.

# How to read this project {#sec-map}

Each stage and figure ships its own tutorial, written to be read on its own. Start at
the top and follow the chain, or jump to the one you need.

| Document | What it teaches |
|---|---|
| `00_input/a_script/00_inputs.qmd` | What DIA-MS is, what a protein group is, why a zero is not a zero |
| `01_Filtering/a_script/01_filtering.qmd` | MAR vs MNAR, the contaminant gate, the outlier consensus, and where the cutoffs came from |
| `02_Normalization/a_script/02_normalization.qmd` | cycloess, the four imputation arms, and why imputing before testing manufactures significance |
| `03_DEP/a_non_imputed/a_script/03_dep.qmd` | The cell-means design, `duplicateCorrelation`, eBayes, BH, and the π-score |
| `04_Figures/F03_pathway/supp/` | Pilot read-outs: HR-vs-LR concordance (training and acute) and the magnitude/concordance summary, before anything is finalised |
| `04_Figures/F01_phenotype/a_script/F01_phenotype.qmd` | The phenotype: matched work, divergent growth |
| `04_Figures/F02_proteome/a_script/F02_proteome.qmd` | The six global nulls (PERMANOVA, RRPP, PERMDISP, ω², CAP) |
| `04_Figures/F03_pathway/a_script/F03_pathway.qmd` | Volcano-in-ring, NES, fgsea, and the EnrichmentMap dedup |
| `04_Figures/shared/WGCNA/a_script/WGCNA.qmd` | WGCNA modules, `limma::fry`, and why the earlier module-phenotype positive was a subject-identity artifact; `loso_refit/`, `preservation/` and `contrast_networks/` test whether the modules generalize |
| `04_Figures/F05_classification/` | HR/LR classification screen: `<level>/<config>/HR_LR/<model>`, 153 cells, nested LOSO against a permutation null; 0 leads |
| `04_Figures/F06_prediction/` | Continuous-adaptation prediction screen: `<level>/<config>/<phenotype>/<model>`, 792 cells, nested LOSO against a permutation null; 32 leads, 26 on `d_mcsa`, all 8 module leads fall below zero on the two reproducible modules |

Known limitations are stated in each stage and figure page where the reader meets
them, rather than in a separate document: the π gate in @sec-pi-warning; the human-only
search space, which had no contaminant FASTA and no decoys, so reagent contaminants cannot be
detected at all (`01_filtering.qmd`); the 34 proteins the missingness filter admits and the
model then cannot test (`03_dep.qmd`); the module-prediction circularity that in-fold
refitting exposed in `shared/WGCNA`; and the transductive eigengenes `shared/WGCNA` shares
with F05-F06's `modules` level.

Architecture (shared house style): each stage is a numbered folder with
`a_script/` (code), `b_reports/` (proteoDA QC PDFs + figures), `c_data/` (RDS
hand-offs + a consolidated `.xlsx`). Stages are self-cleaning: a re-run wipes its
own outputs first.

```
00_input/         raw matrix, metadata, HPA reference
01_Filtering/     contaminant + missingness + outlier filtering  → DAList_filtered.rds
02_Normalization/ cycloess normalize; 4 imputation arms          → DAList_normalized.rds
03_DEP/
  a_non_imputed/  PRIMARY limma DE (NA-tolerant)                 → 03_combined_results.csv
  b_imputed/      imputed DE (concordance check)
04_Figures/       F03_pathway/supp pilot read-outs; F01-F03 publication composites;
                  shared/WGCNA module source; F05-F06 per-cell screens
```

# Stage 00 — Inputs {#sec-inputs}

| File | Contents |
|---|---|
| `00_input/HRvLR_raw.xlsx` | DIA-MS protein matrix, 2,400 protein groups × (5 annotation cols + 48 samples) |
| `00_input/HRvLR_meta.csv` | Sample sheet (`Col_ID`, `Subject_ID`, `Group`, `Timepoint`, `Group_Time`, plus phenotype columns) |
| `00_input/HPA_annotations_full.tsv` | Human Protein Atlas reference: secretome location, blood concentration, single-cell nCPM, protein class |

The matrix is aligned to the sample sheet by `Col_ID`; a mismatch errors
immediately.

# Stage 01 — Filtering {#sec-filtering}

The filter removes blood-plasma contaminants from a muscle biopsy without
discarding real muscle biology. A protein is flagged **remove** if it is
plasma-origin **and not** muscle-origin, sourced entirely from the Human Protein
Atlas:

- **plasma-origin** = HPA "Secreted to blood", an immunoglobulin gene, or a high
  erythrocyte single-cell signal (`Erythrocytes ≥ 5000 nCPM`, the hemoglobins);
- **muscle rescue** = high skeletal-myonuclei signal (`Myonuclei ≥ 50 nCPM`),
  unless the measured blood concentration identifies it as a true plasma protein
  (`blood_conc ≥ 1e9`).

```{r}
#| label: blood-classifier
bl <- readr::read_tsv(here::here("00_input", "HPA_annotations_full.tsv"), show_col_types = FALSE) |>
  dplyr::transmute(
    gene = Gene, protein_class = `Protein class`, secretome = `Secretome location`,
    blood_conc = suppressWarnings(as.numeric(`Blood concentration - Conc. blood MS [pg/L]`)),
    ery = suppressWarnings(as.numeric(`Single Cell Type RNA - Erythrocytes [nCPM]`)),
    myo = suppressWarnings(as.numeric(`Single Cell Type RNA - Myonuclei [nCPM]`))
  ) |>
  dplyr::mutate(
    secreted_blood = secretome == "Secreted to blood" & !is.na(secretome),
    is_ig = stringr::str_detect(dplyr::coalesce(protein_class, ""), "Immunoglobulin genes"),
    is_erythrocyte = !is.na(ery) & ery >= 5000,
    muscle_keep = !is.na(myo) & myo >= 50 & (is.na(blood_conc) | blood_conc < 1e9),
    verdict = dplyr::if_else(!muscle_keep & (secreted_blood | is_ig | is_erythrocyte),
      "remove", "keep"
    )
  )
```

After dedup (one row per accession), HPA presence, blood removal, and a group-wise
missingness filter (`filter_proteins_by_group`, ≥5 of 8 detected in ≥1 cell), the
matrix is screened for outlier samples by a four-method consensus — missingness
(with a within-subject arm), PCA Mahalanobis distance, a Hampel rule on median
intensity, and median inter-sample correlation — dropping a sample only when ≥3 of
4 agree.

```{r}
#| label: filter-and-outliers
dal <- zero_to_missing(DAList(data = int_mat, annotation = annot_df, metadata = metadata))
dal <- filter_proteins_by_group(dal, min_reps = 5, min_groups = 1, grouping_column = "Group_Time")
# 4-method consensus -> filter_samples(dal, !(Col_ID %in% outlier_ids))
```

Stage 01 turns the 2,400 × 48 spreadsheet into a clean **1,896 × 45** `DAList`
(three samples — `S29_T1`, `S28_T2`, `S28_T3` — trip ≥3 flags), saved to
`01_Filtering/c_data/DAList_filtered.rds`.

# Stage 02 — Normalization & imputation {#sec-norm}

Cyclic-loess normalization levels per-sample loading differences with a single
native call; `write_norm_report()` compares candidate methods first, and
`write_qc_report()` audits before and after.

```{r}
#| label: normalize
dal <- readRDS(here::here("01_Filtering", "c_data", "DAList_filtered.rds"))
dal <- normalize_data(dal, norm_method = "cycloess") # limma defaults: span 0.7
saveRDS(dal, here::here("02_Normalization", "c_data", "DAList_normalized.rds"))
```

Normalization rescales but never fills values; the normalized, **non-imputed**
matrix is the primary input to the DE test, because limma handles per-protein NAs
and imputing first inflates false positives. The stage also records a per-protein
eta-squared (variance explained by `Group_Time`) for the figure layer.

Four imputation arms supply a complete matrix for methods that need one (PCA,
WGCNA, rotation tests). **missForest is canonical** (figures and WGCNA read it);
imp4p, the MsCoreUtils kNN/QRILC hybrid, and the Perseus MNAR draw are kept as
comparisons.

```{r}
#| label: impute
set.seed(42)
mat <- as.matrix(dal$data)
ord <- order(rownames(mat)) # deterministic order before the stochastic fit
mf <- missForest::missForest(t(mat[ord, ]), maxiter = 10, ntree = 100) # canonical arm
imp <- t(mf$ximp)[rownames(mat), ] # proteins back to rows
```

# Stage 03 — Differential expression {#sec-dep}

The primary DE fits limma with `duplicateCorrelation` on the non-imputed matrix.
The cell-means design gives one coefficient per `Group_Time` cell; the
`(1 | subject)` block pools a single within-subject correlation across all
proteins [@smyth2005], and `eBayes` moderates each protein's variance
[@ritchie2015].

```{r}
#| label: dep-fit
dal <- add_design(dal, "~ 0 + group + (1 | subject)")
dal <- add_contrasts(dal, contrasts_vector = c(
  "Training_HR = HR_T2 - HR_T1", "Training_LR = LR_T2 - LR_T1",
  "Acute_HR = HR_T3 - HR_T2", "Acute_LR = LR_T3 - LR_T2",
  "Baseline_HRvLR = HR_T1 - LR_T1", "Trained_HRvLR = HR_T2 - LR_T2",
  "Acute_HRvLR = HR_T3 - LR_T3",
  "Training_Interaction = (HR_T2 - HR_T1) - (LR_T2 - LR_T1)",
  "Acute_Interaction = (HR_T3 - HR_T2) - (LR_T3 - LR_T2)"
))
dal <- fit_limma_model(dal)
dal <- extract_DA_results(dal, pval_thresh = 0.10, lfc_thresh = 0, adj_method = "BH")
```

The reported gate is the π-score [@xiao2014], `π = p^|logFC| < 0.05`, alongside a
BH-FDR threshold of 0.10 [@benjamini1995].

**These are not two levels of the same filter.** BH controls the false-discovery rate;
π controls nothing, and π is *not* a stricter version of p < 0.05 — it is a different
screen that admits proteins the p-value would reject (see @sec-pi-warning). Read the
BH counts as discoveries and the π counts as a ranking. The comparison arm
(`b_imputed/a_script/01_run_dep_imputed.R`) repeats the fit on the four imputed
matrices and reports Spearman concordance of each arm's logFC against the primary;
high concordance confirms imputation does not reshape the effects [@peng2024].

# Stage 04 — Figures {#sec-figures}

| Figure | Content |
|---|---|
| F01 | Phenotype atlas: HR vs LR growth and strength despite matched training |
| F02 | Proteome overview and QC: PCA, DEP counts, effect sizes, set overlaps, eta-squared |
| F03_pathway | enrichVolcano ring-volcanoes per contrast |
| F03_pathway/supp | Pilot read-outs: HR-vs-LR concordance (training and acute — quadrant ORA, NES scatter, fry, RRHO2, bootstrap) and the magnitude/concordance summary |
| shared/WGCNA | WGCNA modules on the missForest-imputed proteome, linked to phenotype; the eigengene source for the `modules` level of F05-F06; `loso_refit`, `preservation`, `contrast_networks` test generalization |
| F05_classification | HR/LR classification screen over `<level>/<config>/HR_LR/<model>`: 153 cells, nested LOSO against a permutation null; 0 leads |
| F06_prediction | Continuous-adaptation prediction screen over `<level>/<config>/<phenotype>/<model>`: 792 cells, nested LOSO against a permutation null; 32 leads, 26 on `d_mcsa`, module leads fall below zero on the two reproducible modules |

The HR-vs-LR response question is first sighted in F03_pathway/supp, where the two groups'
contrasts are compared directly with rotation tests and rank-rank overlap, then taken
up module-wise in `shared/WGCNA` and by classification and prediction in
F05_classification and F06_prediction.

# Reproducing {#sec-repro}

Run the stage scripts in order from the project root:

```bash
Rscript 00_input/a_script/01_build_phenotype.R
Rscript 00_input/a_script/02_build_blood_list.R
Rscript 01_Filtering/a_script/01_run_filtering.R
Rscript 02_Normalization/a_script/01_run_normalization.R
Rscript 02_Normalization/imputation/a_script/impute_missforest.R
Rscript 02_Normalization/imputation/a_script/impute_imp4p.R
Rscript 02_Normalization/imputation/a_script/impute_mscoreutils.R
Rscript 02_Normalization/imputation/a_script/impute_perseus.R
Rscript 03_DEP/a_non_imputed/a_script/01_run_dep.R
Rscript 03_DEP/a_non_imputed/a_script/02_blood_filter_sensitivity.R
Rscript 03_DEP/a_non_imputed/a_script/03_untested_proteins.R
Rscript 03_DEP/b_imputed/a_script/01_run_dep_imputed.R
Rscript 03_DEP/b_imputed/a_script/02_imp4p_circularity.R

# imputation QC reads the b_imputed outputs, so it runs last, after stage 03:
Rscript 02_Normalization/imputation/a_script/imputation_qc.R
```

# References {#sec-refs}

::: {#refs}
:::
