---
title: "High vs Low Responders (HRvLR) Skeletal-Muscle Proteomics"
subtitle: "A ground-up walkthrough of the proteoDA-native pipeline: filtering → normalization → differential expression → figures"
author: "A_HRvLR_2026"
date: today
bibliography: 03_DEP/a_non_imputed/a_script/references.bib
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: true
    code-tools: true
    theme: cosmo
    fig-cap-location: bottom
    embed-resources: true
execute:
  eval: false      # documentation build: code is shown, not re-run; outputs are frozen
  echo: true
  warning: false
  message: false
---

> **How to read this document.** Every code block is the *actual* code from the
> pipeline scripts, shown for provenance but **not re-executed** (`eval: false`).
> Numbers are frozen from the committed run. To reproduce, run the stage scripts
> in order. Paths are relative to the project root (`A_HRvLR_2026/`). Each stage
> also ships a live, self-verifying tutorial (`01_filtering.qmd`,
> `02_normalization.qmd`, `03_dep.qmd`).

# Overview {#sec-overview}

Resistance training produces a wide range of hypertrophic outcomes across people.
This pipeline asks whether that range is visible in the skeletal-muscle proteome:
do **high responders (HR)** and **low responders (LR)** differ in how their
proteomes move with training, and with an acute bout?

The design is a 2×3 factorial — responder (HR/LR) × timepoint — with repeated
measures on subject. Each subject is sampled three times: T1 (baseline), T2
(trained, 72 h post a training block), and T3 (acute, 1 h post a final bout). The
grouping factor `Group_Time` carries the six cells.

| `Group_Time` | Meaning | n samples |
|---|---|---|
| `HR_T1` … `HR_T3` | High responders, three timepoints | 7 each |
| `LR_T1` … `LR_T3` | Low responders, three timepoints | 8 each |

Nine contrasts are built on this scheme:

| Contrast | Definition | Question |
|---|---|---|
| `Training_HR` / `Training_LR` | `*_T2 − *_T1` | Training response within each group |
| `Acute_HR` / `Acute_LR` | `*_T3 − *_T2` | Acute-bout response within each group |
| `Baseline_HRvLR` | `HR_T1 − LR_T1` | Group difference at baseline |
| `Trained_HRvLR` / `Acute_HRvLR` | `HR_T2 − LR_T2` / `HR_T3 − LR_T3` | Group difference after training / acute |
| `Training_Interaction` | `(HR_T2 − HR_T1) − (LR_T2 − LR_T1)` | Does the training response differ by group? |
| `Acute_Interaction` | `(HR_T3 − HR_T2) − (LR_T3 − LR_T2)` | Does the acute response differ by group? |

The interaction contrasts are the analytic core: they ask the responder question
directly.

Architecture (shared house style): each stage is a numbered folder with
`a_script/` (code), `b_reports/` (proteoDA QC PDFs + figures), `c_data/` (RDS
hand-offs + a consolidated `.xlsx`). Stages are self-cleaning: a re-run wipes its
own outputs first.

```
00_input/         raw matrix, metadata, HPA reference
01_Filtering/     contaminant + missingness + outlier filtering  → DAList_filtered.rds
02_Normalization/ cycloess + (missforest | imp4p | mscoreutils)  → DAList_normalized.rds
03_DEP/
  a_non_imputed/  PRIMARY limma DE (NA-tolerant)                 → 03_combined_results.csv
  b_imputed/      imputed DE (concordance check)
04_Figures/       F01–F06 publication composites
```

# Stage 00 — Inputs {#sec-inputs}

| File | Contents |
|---|---|
| `00_input/HRvLR_raw.xlsx` | DIA-MS protein matrix, 2,400 protein groups × (5 annotation cols + 48 samples) |
| `00_input/HRvLR_meta.csv` | Sample sheet (`Col_ID`, `Subject_ID`, `Group`, `Timepoint`, `Group_Time`, plus phenotype columns) |
| `00_input/HPA_annotations.tsv` | Human Protein Atlas reference: secretome location, blood concentration, single-cell nCPM, protein class |

The matrix is aligned to the sample sheet by `Col_ID`; a mismatch errors
immediately.

# Stage 01 — Filtering {#sec-filtering}

The filter removes blood-plasma contaminants from a muscle biopsy without
discarding real muscle biology. A protein is flagged **remove** if it is
plasma-origin **and not** muscle-origin, sourced entirely from the Human Protein
Atlas:

- **plasma-origin** = HPA "Secreted to blood", an immunoglobulin gene, or a high
  erythrocyte single-cell signal (`Erythrocytes ≥ 5000 nCPM`, the hemoglobins);
- **muscle rescue** = high skeletal-myonuclei signal (`Myonuclei ≥ 50 nCPM`),
  unless the measured blood concentration identifies it as a true plasma protein
  (`blood_conc ≥ 1e9`).

```{r}
#| label: blood-classifier
bl <- readr::read_tsv(here::here("00_input", "HPA_annotations.tsv"), show_col_types = FALSE) |>
  dplyr::transmute(
    gene = Gene, protein_class = `Protein class`, secretome = `Secretome location`,
    blood_conc = suppressWarnings(as.numeric(`Blood concentration - Conc. blood MS [pg/L]`)),
    ery = suppressWarnings(as.numeric(`Single Cell Type RNA - Erythrocytes [nCPM]`)),
    myo = suppressWarnings(as.numeric(`Single Cell Type RNA - Myonuclei [nCPM]`))
  ) |>
  dplyr::mutate(
    secreted_blood = secretome == "Secreted to blood" & !is.na(secretome),
    is_ig = stringr::str_detect(dplyr::coalesce(protein_class, ""), "Immunoglobulin genes"),
    is_erythrocyte = !is.na(ery) & ery >= 5000,
    muscle_keep = !is.na(myo) & myo >= 50 & (is.na(blood_conc) | blood_conc < 1e9),
    verdict = dplyr::if_else(!muscle_keep & (secreted_blood | is_ig | is_erythrocyte),
      "remove", "keep"
    )
  )
```

After dedup (one row per accession), HPA presence, blood removal, and a group-wise
missingness filter (`filter_proteins_by_group`, ≥5 of 8 detected in ≥1 cell), the
matrix is screened for outlier samples by a four-method consensus — missingness
(with a within-subject arm), PCA Mahalanobis distance, a Hampel rule on median
intensity, and median inter-sample correlation — dropping a sample only when ≥3 of
4 agree.

```{r}
#| label: filter-and-outliers
dal <- zero_to_missing(DAList(data = int_mat, annotation = annot_df, metadata = metadata))
dal <- filter_proteins_by_group(dal, min_reps = 5, min_groups = 1, grouping_column = "Group_Time")
# 4-method consensus -> filter_samples(dal, !(Col_ID %in% outlier_ids))
```

Stage 01 turns the 2,400 × 48 spreadsheet into a clean **1,885 × 45** `DAList`
(three samples — `S29_T1`, `S28_T2`, `S28_T3` — trip ≥3 flags), saved to
`01_Filtering/c_data/DAList_filtered.rds`.

# Stage 02 — Normalization & imputation {#sec-norm}

Cyclic-loess normalization levels per-sample loading differences with a single
native call; `write_norm_report()` compares candidate methods first, and
`write_qc_report()` audits before and after.

```{r}
#| label: normalize
dal <- readRDS(here::here("01_Filtering", "c_data", "DAList_filtered.rds"))
dal <- normalize_data(dal, norm_method = "cycloess") # limma defaults: span 0.7
saveRDS(dal, here::here("02_Normalization", "c_data", "DAList_normalized.rds"))
```

Normalization rescales but never fills values; the normalized, **non-imputed**
matrix is the primary input to the DE test, because limma handles per-protein NAs
and imputing first inflates false positives. The stage also records a per-protein
eta-squared (variance explained by `Group_Time`) for the figure layer.

Four imputation arms supply a complete matrix for methods that need one (PCA,
WGCNA, rotation tests). **missForest is canonical** (figures and WGCNA read it);
imp4p, the MsCoreUtils kNN/QRILC hybrid, and the Perseus MNAR draw are kept as
comparisons.

```{r}
#| label: impute
set.seed(42)
mat <- as.matrix(dal$data)
ord <- order(rownames(mat)) # deterministic order before the stochastic fit
mf <- missForest::missForest(t(mat[ord, ]), maxiter = 10, ntree = 100) # canonical arm
imp <- t(mf$ximp)[rownames(mat), ] # proteins back to rows
```

# Stage 03 — Differential expression {#sec-dep}

The primary DE fits limma with `duplicateCorrelation` on the non-imputed matrix.
The cell-means design gives one coefficient per `Group_Time` cell; the
`(1 | subject)` block pools a single within-subject correlation across all
proteins [@smyth2005], and `eBayes` moderates each protein's variance
[@ritchie2015].

```{r}
#| label: dep-fit
dal <- add_design(dal, "~ 0 + group + (1 | subject)")
dal <- add_contrasts(dal, contrasts_vector = c(
  "Training_HR = HR_T2 - HR_T1", "Training_LR = LR_T2 - LR_T1",
  "Acute_HR = HR_T3 - HR_T2", "Acute_LR = LR_T3 - LR_T2",
  "Baseline_HRvLR = HR_T1 - LR_T1", "Trained_HRvLR = HR_T2 - LR_T2",
  "Acute_HRvLR = HR_T3 - LR_T3",
  "Training_Interaction = (HR_T2 - HR_T1) - (LR_T2 - LR_T1)",
  "Acute_Interaction = (HR_T3 - HR_T2) - (LR_T3 - LR_T2)"
))
dal <- fit_limma_model(dal)
dal <- extract_DA_results(dal, pval_thresh = 0.10, lfc_thresh = 0, adj_method = "BH")
```

Significance uses the Pi-score [@xiao2014], `Pi = p^|logFC| < 0.05`, on top of a
BH-FDR threshold of 0.10 [@benjamini1995] — appropriate for a small-n DIA study
powered for large effects. The comparison arm
(`b_imputed/a_script/01_run_dep_imputed.R`) repeats the fit on the four imputed
matrices and reports Spearman concordance of each arm's logFC against the primary;
high concordance confirms imputation does not reshape the effects [@peng2024].

# Stage 04 — Figures {#sec-figures}

| Figure | Content |
|---|---|
| F01 | Phenotype atlas: HR vs LR growth and strength despite matched training |
| F02 | Proteome overview and QC: PCA, DEP counts, effect sizes, set overlaps, eta-squared |
| F03 | enrichVolcano ring-volcanoes per contrast |
| F04 | Training concordance: HR vs LR (quadrant ORA, NES scatter, fry, RRHO2, bootstrap) |
| F05 | Acute concordance: HR vs LR |
| F06 | WGCNA modules on the missForest-imputed proteome, linked to phenotype |

The HR-vs-LR response question is answered in F04/F05, where the two groups'
contrasts are compared directly with rotation tests and rank-rank overlap.

# Reproducing {#sec-repro}

Run the stage scripts in order from the project root:

```bash
Rscript 01_Filtering/a_script/01_run_filtering.R
Rscript 02_Normalization/a_script/01_run_normalization.R
Rscript 02_Normalization/imputation/a_script/a_imp4p.R
Rscript 02_Normalization/imputation/a_script/b_mscoreutils.R
Rscript 02_Normalization/imputation/a_script/c_missforest.R
Rscript 02_Normalization/imputation/a_script/d_perseus.R
Rscript 03_DEP/a_non_imputed/a_script/01_run_dep.R
Rscript 03_DEP/b_imputed/a_script/01_run_dep_imputed.R
```

# References {#sec-refs}

::: {#refs}
:::
