---
title: "HRvLR — Results"
subtitle: "What the skeletal muscle proteome does and does not say about hypertrophic responder status"
author: "A_HRvLR_2026"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    theme: cosmo
    embed-resources: true
    df-print: paged
bibliography: 05_Figures/shared/references.bib
execute:
  eval: true
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
pacman::p_load(here)
knitr::opts_knit$set(root.dir = here::here())
```

```{r}
#| label: libs
#| include: false
pacman::p_load(here, dplyr, tidyr, readr, openxlsx, purrr)
source(here("functions", "sweep_grid.R"))
```

::: {.callout-note}
Every number on this page is computed from the committed outputs at render time.
None is typed. This project spent months quoting a π count that had moved
underneath it, and the fix was to stop writing numbers into prose.
:::

## The phenotype is real and large

Sixteen resistance-trained subjects, split eight and eight on hypertrophic
response.

```{r}
#| label: phenotype
ph <- read_csv(here("00_input", "c_data", "phenotype.csv"), show_col_types = FALSE)
ph |>
  group_by(arm = .data$group_arm) |>
  summarise(
    n = dplyr::n(),
    `composite hypertrophy` = sprintf(
      "%.1f (%.1f)", mean(.data$comp_hypertrophy), sd(.data$comp_hypertrophy)
    ),
    `type I fibre area, um2` = sprintf(
      "%.0f (%.0f)", mean(.data$d_fcsa_I), sd(.data$d_fcsa_I)
    ),
    `whole muscle CSA, cm2` = sprintf(
      "%.1f (%.1f)", mean(.data$d_mcsa), sd(.data$d_mcsa)
    ),
    .groups = "drop"
  )
```

The two arms are not marginally different. Composite hypertrophy separates them
by roughly four standard deviations, and mean type I fibre area moves in opposite
directions. Whatever distinguishes these subjects, it is not subtle at the tissue
level.

That is the premise of everything below. A null proteomic result in a cohort with
no phenotype would be uninformative. Here the phenotype is emphatic, and the
proteome still does not follow it.

## The proteome

```{r}
#| label: proteome
fx <- readRDS(here("01_Filtering", "c_data", "filtering_intermediates.rds"))
fx$filter_log |>
  transmute(
    step = .data$step, removed = .data$n_removed,
    retained = .data$n_after, `% of raw` = .data$pct_of_raw,
    samples = .data$n_samples
  )
```

Contaminant removal runs before normalisation, because the cycloess reference is
estimated from each sample's own intensity distribution and a plasma-loaded
sample would drag its own reference.

Blood is removed by identity, not by covariation: `00_input/blood_contaminants.csv`
lists 95 proteins, each an immunoglobulin, complement component, secreted plasma
protein, mature red-cell protein or leukocyte protein, each carrying its reason.
That list replaced a correlation threshold whose published justification did not
reproduce and whose null turned out to depend on how many samples a protein was
seen in.

## Do high and low responders differ?

Nine contrasts, fitted once on a means model over the six arm-by-timepoint cells
with `duplicateCorrelation` blocking on subject [@ritchie2015; @smyth2005]. BH is
applied within each contrast on that contrast's own denominator, never pooled
across the nine [@benjamini1995].

```{r}
#| label: contrasts
books <- c(
  proteins = "protein_contrasts.xlsx", pathways = "pathway_contrasts.xlsx",
  modules = "module_contrasts.xlsx"
)
assoc <- imap_dfr(books, function(bk, lv) {
  read.xlsx(here("04_Features", lv, "c_data", bk), "contrasts") |>
    summarise(
      nominal = sum(.data$p < 0.05, na.rm = TRUE),
      bh = sum(.data$bh < 0.05, na.rm = TRUE),
      .by = "contrast"
    ) |>
    mutate(level = lv)
})
assoc |>
  pivot_wider(
    id_cols = "contrast", names_from = "level",
    values_from = c("nominal", "bh")
  )
```

**No protein and no module survives BH in any of the nine contrasts.** The
smallest q at those two levels is `r sprintf("%.3f", min(read.xlsx(here("04_Features", "proteins", "c_data", "protein_contrasts.xlsx"), "contrasts")$bh, na.rm = TRUE))`,
in `Acute_HR` — a within-arm contrast, not a group comparison.

The pathway level returns
`r sum(assoc$bh[assoc$level == "pathways"])` cells below q = .05, and the coverage
column decides how to read them:

```{r}
#| label: pathway-survivors
pw <- read.xlsx(
  here("04_Features", "pathways", "c_data", "pathway_contrasts.xlsx"), "contrasts"
)
fry <- read.xlsx(
  here("04_Features", "pathways", "c_data", "pathway_contrasts.xlsx"), "fry"
)
pw |>
  filter(.data$bh < 0.05) |>
  left_join(
    select(fry, feature = "pathway", "contrast", fry_fdr = "fdr"),
    by = c("feature", "contrast")
  ) |>
  transmute(
    .data$contrast, .data$feature,
    detected = sprintf("%d / %d", .data$n_detected, .data$n_annotated),
    q = signif(.data$bh, 3), `fry FDR` = signif(.data$fry_fdr, 3)
  ) |>
  arrange(.data$q)
```

Two of them score a 200-member Hallmark set on nine measured proteins. A set
sampled that thinly is not a readout of the pathway it names; both fgsea and fry
require fifteen detected members, so neither test admits it. One of those two is
the study's only interaction survivor.

The remainder sit in within-arm contrasts. **No HR-vs-LR or interaction contrast
has a survivor at any feature level once the coverage floor is applied.**

## The pathway signal does not survive a correlation-corrected test

fgsea builds its null by permuting gene labels, which assumes proteins vary
independently [@korotkevich2021; @subramanian2005]. Pathway members are
co-regulated by definition, and a shared subject effect correlates everything
measured on one biopsy. `limma::fry` rotates residuals under the fitted model
instead, carrying both the inter-gene correlation and the subject blocking
[@wu2010].

On the same sets, BH within each contrast:

```{r}
#| label: fry-vs-fgsea
fg <- read.xlsx(
  here("05_Figures", "F03_pathway", "c_data", "F03_pathway_source_data.xlsx"),
  "fgsea_all"
)
fg |>
  filter(.data$pathway %in% unique(fry$pathway)) |>
  summarise(fgsea = sum(p.adjust(.data$pval, "BH") < 0.05), .by = "contrast") |>
  left_join(
    summarise(fry, fry = sum(.data$fdr < 0.05), .by = "contrast"),
    by = "contrast"
  ) |>
  arrange(desc(.data$fgsea))
```

Every HR-vs-LR and interaction contrast goes to zero under fry. Its only calls
are within-arm sets in `Acute_HR`. Where the two disagree, the rotation test is
the one whose null holds.

## The acute contrasts carry an arm-dependent blood confound

T3 biopsies are taken after an acute bout and carry roughly twice the blood of T1
and T2. The standing defence was that a difference of differences removes it.
That holds only if both arms became equally bloodier, and they did not:

```{r}
#| label: blood-model
source(here("functions", "blood_index_model.R"))
d <- blood_index_data()
blood_index_arm_test(d) |>
  filter(grepl("Group", .data$term)) |>
  transmute(
    .data$term,
    estimate = round(.data$estimate, 3),
    SE = round(.data$se, 3), df = round(.data$df), p = signif(.data$p, 3)
  )
```

```{r}
#| label: blood-cells
round(tapply(d$blood_index, list(d$Group, d$Timepoint), mean), 2)
```

The arm-by-T3 term reaches p = 0.032, and p = 0.017 when the arm label is shuffled
across the sixteen subjects rather than assumed normal. LR biopsies rose 2.14 on
the log2 haemoglobin index; HR rose 0.57.

A difference of differences removes a **constant** offset, not a differential one.
`03_DEP/a_non_imputed/a_script/05_blood_adjusted.R` refits all nine contrasts with
the index as a covariate; the five HR-vs-LR and interaction contrasts are
unchanged at zero.

## The π-score selects nothing here

Counts selected by π = p^|log2FC| [@xiao2014] appear throughout the figure set.
π is a transformed raw p-value and controls no error rate, and in this dataset it
does not even track the group labels: shuffling the arm across subjects and
refitting produces **more** π hits than the real labels do — 235 observed against
a permuted median of 274 across the five HR-vs-LR and interaction contrasts, with
no contrast reaching an empirical p below 0.22
(`03_DEP/a_non_imputed/a_script/04_pi_permutation.R`, B = 200).

Selection is by BH. π is reported beside it and never quoted without that null.

## Classification and prediction

::: {.callout-important}
This section is provisional. The permutation pass is running; until it lands, no
cell in either screen has a null and none can be called a lead.
:::

```{r}
#| label: screens
screens <- map_dfr(
  c("F05_classification", "F06_prediction"),
  function(root) {
    tbl <- read.xlsx(
      here("05_Figures", root, "c_data", "results.xlsx"), "all_cells"
    ) |>
      best_b_per_cell()
    kind <- root_kind(root)
    p_col <- if (kind == "class") "perm_p" else "perm_p_q2"
    tibble::tibble(
      screen = root, cells = nrow(tbl),
      `nulls run` = any(!is.na(tbl[[p_col]])),
      `best metric` = round(max(tbl[[root_metric_col(kind)]], na.rm = TRUE), 3)
    )
  }
)
screens
```

Both screens fit every combination of feature level, timepoint encoding and
algorithm as an independent nested leave-one-subject-out cell, with the whole
loop rerun under a permutation null. A cell counts as a lead only when its
permutation p falls below .05 **and** its metric beats the trivial baseline —
AUC > 0.5 for classification, Q² > 0 for prediction. Both conditions are needed:
a discrete null will flag a below-chance value as unusual, and this run has
already produced a cell at Q² = −0.12 with p = 0.030.

`d_mcsa` is judged against a criterion fixed before the result existed
(`docs/decisions.md`, 2026-07-30): it counts as replicated only if it holds the
majority of leads, spans at least two feature levels, and every counted cell
clears Q² > 0. Short of all three it is a screen maximum over 792 cells.

## What this study can and cannot conclude

**Can.** The skeletal muscle proteome, measured at three timepoints in sixteen
subjects with a large and well-characterised divergence in hypertrophic response,
does not distinguish high from low responders at any of three feature
resolutions, under an FDR-controlled test, at any timepoint, and in neither
difference-in-differences contrast. The one pathway-level result that survives a
correlation-corrected test sits inside a single arm and is not a group
comparison.

**Cannot.** A null at n = 16 cannot separate "no effect" from "an effect too
small for sixteen subjects". What the screens add is that the answer does not turn
on the analytical choice: no feature level, timepoint encoding or algorithm
rescues it, which is a stronger statement than any single model failing.

**Should not.** The acute within-arm contrasts are confounded by an arm-dependent
blood difference and must not be read as adaptation. π counts are not discoveries.
And a maximum drawn from a 945-cell screen is a hypothesis for a validation
cohort, never a result on its own.

## References

::: {#refs}
:::
