---
title: "HRvLR 01 — Filtering"
subtitle: "Building the DAList: contaminant, missingness, and outlier removal"
author: "A_HRvLR_2026"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
    embed-resources: true
    df-print: paged
bibliography: ../../04_Figures/shared/references.bib
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
pacman::p_load(here)
knitr::opts_knit$set(root.dir = here::here())
```

## What stage 01 does

Stage 01 turns the raw DIA-MS spreadsheet into a filtered `DAList`, the proteoDA
object every later stage builds on. It collapses duplicate accessions, removes
blood-plasma contamination, drops proteins too sparsely detected to model, and
removes any sample that fails a consensus of outlier checks.

This page **reads what that script produced**; it does not recompute it.
`01_Filtering/a_script/01_run_filtering.R` decides every verdict once and writes them to
`c_data/`. Re-deriving them here would be a second implementation of the filter, free to drift
away from the one that actually made the data, and a reader would have no way to tell which
number was real. Every figure and table below comes from the committed artifacts.

The design is repeated measures: high responders (HR) and low responders (LR),
each sampled at T1 (baseline), T2 (trained), and T3 (acute). The factor
`Group_Time` carries the six cells (`HR_T1` … `LR_T3`) and is the grouping column
for the missingness filter.

## Reads and writes

The stage reads three files from `00_input/`: the protein matrix
`HRvLR_raw.xlsx`, the sample sheet `HRvLR_meta.csv`, and the contaminant evidence
`HPA_annotations_full.tsv`. The runnable script writes `DAList_filtered.rds` (the
input to stage 02), `filtering_intermediates.rds` (reused by the stage-02 report
and the figure layer), `filtering_report.xlsx`, `protein_calls.csv` (one verdict per
protein), `contamination_index.csv` (per-sample contaminant load), and the QC figure
`filtering_qc.png`. This document reads those artifacts and writes nothing.

## The filtered object {#sec-load}

`DAList_filtered.rds` is the finished product: a proteoDA `DAList` holding three
name-aligned tables. `filtering_intermediates.rds` carries everything the run script
learned on the way — the per-protein verdicts, the outlier diagnostics, the cascade, and
the thresholds it used — so this page can show its working without repeating it.

```{r}
#| label: load
pacman::p_load(proteoDA, readr, dplyr, tidyr, stringr)
source(here("01_Filtering", "a_script", "filter_config.R"))

dal <- readRDS(here("01_Filtering", "c_data", "DAList_filtered.rds"))
fx <- readRDS(here("01_Filtering", "c_data", "filtering_intermediates.rds"))
n_raw <- fx$n_raw

c(raw_proteins = n_raw, kept_proteins = nrow(dal$data), kept_samples = ncol(dal$data))
table(dal$metadata$Group, dal$metadata$Timepoint)
```

Duplicate UniProt accessions would be collapsed to the highest-mean-intensity row, but this
matrix has none, so that branch never fires: all `r n_raw` raw rows are distinct accessions.

## Remove contaminants and blood proteins {#sec-contaminants}

Two different things get removed here, and conflating them was the bug this stage
used to have.

**Handling contaminants** are proteins that were never in the muscle: skin keratins
shed from gloves and dust, and the haemoglobin chains. No annotation rule can catch
them, because nothing about a keratin marks it as blood. They are curated by hand in
`filter_config.R`, matched **by accession only** — cRAP collides with real muscle
proteins on gene symbol (`ALDOA` with rabbit `ALDOA_RABIT`; `GLUD1` with bovine
P00366, one digit from human P00367) and on description (a `/trypsin|casein|albumin/`
regex eats parvalbumin and the casein kinases). Only the cRAP dust/contact section is
used: its UPS1 section is a spike-in standard we never spiked, and it contains
myoglobin and creatine kinase M — ranks 1 and 2 of this proteome.

**Blood proteins** are a different problem. A muscle biopsy genuinely carries blood,
and plasma proteins are abundant enough to bury the muscle signal. A protein is
removed when HPA marks it blood-derived (secreted-to-blood, immunoglobulin, or a
strong erythrocyte marker) **and** it fails the muscle rescue: high skeletal-myonuclei
expression keeps it, unless its measured plasma concentration marks it as a genuine
plasma protein. The rescue matters because HPA tags *any* plasma-detectable protein as
"Secreted to blood" — which is precisely why creatine kinase and myoglobin are clinical
muscle-damage markers. Without it, GPI (a core glycolytic enzyme) is deleted.

Absence from HPA means **unknown**, never *remove*. The earlier version of this stage
dropped 127 measured proteins for failing to match a partial, muscle-biased export —
including CKMT1A and ATP1A3, whose myonuclei transcript count is zero even though the
instrument measured the protein. `00_input/HPA_annotations_full.tsv` is the complete
20,162-gene table (provenance in `filter_config.R`), so nothing now falls through
that gap.

The verdict for every protein is decided once, in `01_run_filtering.R`, and written to
`protein_calls.csv`. This document reads that table rather than re-deriving it, so the
two cannot drift apart.

```{r}
#| label: calls
calls <- read_csv(here("01_Filtering", "c_data", "protein_calls.csv"), show_col_types = FALSE)
count(calls, verdict)
```

The rescued proteins are the ones the rule would have thrown away: real muscle proteins
that HPA also sees in plasma.

```{r}
#| label: removed-rescued
calls |>
  filter(str_detect(verdict, "rescued")) |>
  select(gene, myo, blood_conc, secretome) |>
  arrange(desc(myo)) |>
  head(8)
```

## The DAList {#sec-build}

A `DAList` holds three name-aligned tables: `data` (the intensity matrix),
`annotation` (one row per protein), and `metadata` (one row per sample). The row
names of `data` match `annotation`, and the column names of `data` match the row
names of `metadata`.

`zero_to_missing()` sets any zero intensity to `NA`. On this dataset it converts nothing:
the DIA export already encodes non-detection as `NA` (24.3% of raw cells), and the smallest
positive intensity is 3,499. The call is a defensive no-op, kept in case a future export
writes zeros. What matters is that the gaps arrive labelled as gaps. Stage 02's "Why a gap is
not just a gap" section is where that assumption starts doing work, because every imputation
choice rests on it. Stage 01 does not impute; it filters.

```{r}
#| label: build-dalist
identical(rownames(dal$data), rownames(dal$annotation))
identical(colnames(dal$data), rownames(dal$metadata))
c(
  zeros_in_matrix = sum(dal$data == 0, na.rm = TRUE),
  pct_missing = round(100 * mean(is.na(dal$data)), 1)
)
```

## Remove outlier samples {#sec-outliers}

Whether a whole sample is untrustworthy is decided by four independent
heuristics: high missingness (with a within-subject arm comparing a subject's
three timepoints), off intensity level (Hampel 3×MAD), multivariate position in
PCA space, and low correlation with the other samples. Requiring at least three of
four to agree keeps any single noisy metric from dropping a good sample.

### The two statistics, in plain terms

**Mahalanobis distance** is how far a sample sits from the centre of the cloud,
measured in a way that respects the shape of the cloud. Plain Euclidean distance
treats the axes as independent and equally scaled, which is wrong when the axes
are correlated: a sample can look unremarkable on PC1 and unremarkable on PC2 and
still be somewhere no sample should be, because in this data PC1 and PC2 move
together and this one moves against the grain. Mahalanobis divides through by the
covariance, so distance is counted in units of how much the data actually vary in
that direction. Under multivariate normality it follows a chi-square distribution
with degrees of freedom equal to the number of dimensions, which is where the
`qchisq(1 - mahal_p, df = 3)` cutoff comes from.

Two honest caveats attach to it. The chi-square cutoff *assumes* PC1–PC3 are
approximately multivariate normal, and nobody has checked that here. And the
distance is computed on complete cases only — the PCA runs on proteins with no
missing values at all (`rowSums(is.na(...)) == 0`). That subset is not a random
sample of the proteome; it is the well-detected, abundant end of it. So the
outlier geometry is measured in a space that has already been selected, and a
sample whose problem lives entirely in the sparse proteins is a sample this test
cannot see.

**MAD and the Hampel rule.** The median absolute deviation is the median of the
absolute distances from the median — a robust stand-in for the standard deviation.
The reason to prefer it here is circular reasoning avoidance: an outlier inflates
the standard deviation, so an SD-based cutoff is widened by the very point it is
supposed to catch, and a bad enough sample hides itself. The median and the MAD
barely move when one value goes wild, so the yardstick stays honest. The Hampel
rule flags anything more than `mad_k` MADs from the median — here 3 — and it is
applied twice, to per-sample median intensity and to per-sample median correlation
with the other samples.

### Three samples were removed, and they are not missing biopsies

This is the part of the stage most easily misread. **48 samples were collected and 45 were
analysed.** The three that went are `S29_T1`, `S28_T2` and `S28_T3` — real biopsies that were
taken, run, and measured, detecting 32–56% of the proteome apiece. They were removed because
three or more of the four checks agreed they were untrustworthy, not because they were never
taken. Anywhere the design is described as "S28 is baseline-only, S29 has no baseline", that is
the *consequence* of this filter and should be stated as such.

The gap is clean: the three flagged samples score 3 or 4 flags and the next-worst sample in the
cohort scores 2. The consensus rule was not straining to make a call.

```{r}
#| label: outliers
fx$outlier_diag |>
  arrange(desc(n_flags)) |>
  select(Col_ID, pct_missing, miss_flag, pca_flag, mad_flag, cor_flag, n_flags, consensus_outlier) |>
  head(6)

fx$outlier_ids
```

## Drop barely-detected proteins {#sec-missingness}

The missingness filter runs **after** outlier removal, and the order is load-bearing.
Run it first and it counts detections contributed by samples that are about to be
discarded, so a protein can be retained on the strength of evidence that is no longer
in the matrix. Two proteins, CRYM and TSNAX, were surviving on exactly that.

`filter_proteins_by_group()` is proteoDA's native missingness filter. With
`min_reps = 5, min_groups = 1` it keeps a protein detected in at least 5 samples of at
least one `Group_Time` cell.

`min_groups = 1` is the deliberate part, and it is the setting most likely to look
like a mistake. The stricter alternative — require the protein to clear `min_reps`
in *every* cell — sounds like better quality control, and it is precisely what we
must not do. A protein that is switched on in one responder group and off in the
other is exactly the biology this study is hunting. Under an all-cells rule it
would be deleted for being absent from the cells where its absence is the finding.
Demanding presence everywhere pre-selects for proteins that do not differ between
groups, which is a filter aimed directly at the hypothesis. Requiring the protein
to hold up in one cell keeps that signal while still removing proteins that are
nowhere consistently detected and cannot support a model anywhere.

The full cascade, as the run script recorded it. `n_removed` counts **proteins**, so the
outlier step scores a legitimate zero there — dropping a sample removes no protein. The
`n_samples` column is what shows the three samples leaving.

```{r}
#| label: cascade
fx$filter_log
```

## Where the numbers came from {#sec-thresholds}

Six constants decide what this stage keeps, and a reader is entitled to ask where
each one came from. The short answer is that they are pragmatic thresholds chosen
against this dataset, not values derived from theory. They are defensible, and
they are not principled. Both things are true and both should be said.

`ery_cut <- 5000` (erythrocyte nCPM). A protein is called an erythrocyte marker
when its HPA single-cell RNA in red cells is at least 5,000 nCPM. The number is
set where it does the job it exists to do: it catches the haemoglobins, which sit
far above any plausible cutoff, without sweeping in proteins that merely happen to
be transcribed in red cells at ordinary levels. It is a cut placed in a gap, not a
cut derived from a null distribution.

`myo_cut <- 20` (myonuclei nCPM). The floor for the muscle rescue. A protein
flagged as blood-derived gets a second hearing if its skeletal-myonuclei RNA
clears 20 nCPM, on the reasoning that muscle genuinely transcribes it and its
presence in a muscle biopsy needs no blood to explain it. It is a modest bar,
chosen to be generous to muscle, because the cost of wrongly deleting a real
muscle protein is higher here than the cost of keeping a doubtful one. It was 50,
and at 50 the rescue failed to fire for GPI (myonuclei 43.1), ANXA2 (40.2) and PPIA
(21.1) — all real muscle proteins. See @sec-limitations: the move to 20 readmits no
contaminant, and it used HPA annotation only.

`blood_max <- 1e9` (pg/L). The veto on the rescue. Without it, a protein that is
both transcribed in muscle *and* genuinely abundant in plasma would be rescued on
the strength of its muscle transcription alone, and its intensity in our matrix
would still be mostly blood. Any protein whose measured blood concentration is at
or above 10⁹ pg/L is refused the rescue no matter how muscle-expressed it is. This
is the one threshold with a clear logical job: it stops the rescue rule from
undoing the contaminant rule.

`mahal_p <- 0.01`. The chi-square tail probability for the PCA flag. It is a
conventional 1% rather than a calibrated error rate, and, as above, it rests on a
normality assumption nobody has tested.

`mad_k <- 3`. The Hampel constant. Three MADs is the standard convention, roughly
comparable to 3 SD under normality, and it is used unchanged for both the
intensity flag and the correlation flag.

`outlier_k <- 3`. Three of the four checks must agree before a sample is removed.
This is the setting that carries the weight, and it is the reason the arbitrariness
of the other five is survivable. Every individual threshold above is a judgement
call, and any one of them could be wrong at the margin. But a sample is only
dropped when three independent heuristics — built on different statistics, looking
at different failure modes — all point at it. A single mis-set cutoff cannot
delete a good sample on its own; it would need two other methods to agree with it.
The consensus rule is the safeguard, and it is what lets the individual numbers be
pragmatic without being dangerous.

## Known limitations {#sec-limitations}

Three things this stage does not do. None of them is fixed here; they are disclosed
so that a reader knows what is in the matrix.

**Reagent contaminants cannot be detected at all.** The delivered matrix contains only
valid human UniProt accessions, and zero `CON__`, `REV__` or decoy entries. Porcine
trypsin, BSA, Lys-C and the caseins are all absent — but a human-only search space
*cannot report* them, so their absence is not evidence they were not there. Worse, an
unsearched contaminant peptide does not vanish; the engine assigns it to the
best-scoring human homolog, turning contamination into a false human identification.
No post-hoc row filter can undo that. Fixing it would require re-searching with a
contaminant FASTA appended. This is a stated limitation, not a solved problem.

**HBG2 was a protein-inference artifact, not fetal haemoglobin.** It is now on the
curated drop list, and the reasoning is worth keeping: HbF is α₂γ₂, and after HBA1 and
HBB were removed, HBG2 was the *only* globin left standing. A gamma chain with no alpha
chain is not a haemoglobin. HBG2 shares tryptic peptides with HBB and HBD, and HBG1 and
HBG2 differ at a single residue, so shared signal was being razor-assigned onto the last
entry standing. The same failure mode probably explains **MYH6** (cardiac α-myosin, in a
vastus lateralis biopsy) and the **ACACA/ACACB** inversion; both need a peptide-level
check before either is read as biology.

**The `myo_cut` threshold was chosen by inspection, and there is no clean gap behind it.**
Lowering it from 50 to 20 recovers GPI, ANXA2 and PPIA, and readmits no contaminant. That
choice was made after looking at where proteins fell, so it is a form of selection and it
should be declared. It is selection on *HPA annotation only* — no contrast, no phenotype and
no group label was consulted — so it cannot leak into the HR/LR result.

What it is not is a cut placed in a gap. Removed-as-plasma proteins span myonuclei 0–236.8
and rescued ones span 20.6–362.3; they overlap across the entire rescue range, and eight
removed proteins sit *above* `myo_cut` (BTD 236.8, A2M 105.1, C1S 89.0, PZP 74.7, ITIH2 44.0,
SERPINF1 28.3, GSN 26.0, ITIH4 21.4). `blood_max` is what separates them: those eight measure
2.3 × 10⁹ – 4.2 × 10¹¹ pg/L in plasma, against 10⁶–10⁸ for genuinely intracellular proteins.
The **measured** concentration does the work. The transcript floor only decides who gets a
second hearing, and it over-deletes: C1QBP (mitochondrial), HMGB2 (nuclear), PPIB, LGALS1 and
CTSB (an exercise myokine) all clear `blood_max` and fail on myonuclei alone.

**The null does not rest on this rule.** Refitting the nine contrasts with all 136
blood-tagged proteins readmitted returns **zero** BH hits in every HR-vs-LR and interaction
contrast (`03_DEP/a_non_imputed/b_reports/blood_filter_sensitivity.csv`). The same run shows
why the rule is kept: readmitting them takes `Acute_LR` from 19 hits to 97, **57 of them the
readmitted blood proteins** — the 2×-bloodier T3 biopsy arriving exactly where it was
predicted to.

## Result {#sec-recap}

proteoDA records every operation in the `tags` slot. Stage 01 reduces the raw
`r n_raw`-protein × 48-sample spreadsheet to a `r nrow(dal$data)` × `r ncol(dal$data)`
`DAList` (proteins × samples).

```{r}
#| label: verify
names(dal$tags)
dim(dal$data)
```

Two proteoDA habits carry through the rest of the pipeline. A `DAList` is three
name-aligned tables, so the native filters preserve alignment and any
hand-subsetting moves `data` and `annotation` together. DIA zeros are missing
values, not measured zeros. Stage 02 picks up `DAList_filtered.rds` and
normalizes it.

## Key terms

**`DAList`** — proteoDA's container [@thurman2023]: three name-aligned tables,
`data` (intensities), `annotation` (one row per protein), `metadata` (one row per
sample). The alignment is enforced, so a filter can never separate a protein from
its annotation.

**MAR (missing at random)** — the gap is unrelated to the protein's true
abundance; a random identification failure. The missingness carries no information
about the value.

**MNAR (missing not at random)** — the protein is missing *because* it fell below
the detection limit. The gap is informative: it means "low". Most DIA gaps are
MNAR-flavoured, which is why a zero must become `NA` and why imputation choice
matters downstream [@lazar2016; @webbrobertson2015].

**`min_groups = 1`** — a protein survives the missingness filter if it clears
`min_reps` in any one `Group_Time` cell. Requiring every cell would delete
proteins that are on in one responder group and off in the other, which is the
biology under investigation.

**Mahalanobis distance** — distance from the centroid measured in units of the
data's own covariance. It catches a sample that is unexceptional on each PC
separately but sits in an impossible place in the correlated space. Assumes
approximate multivariate normality of PC1–PC3, and here it is computed on complete
proteins only, which is itself a selected subset.

**MAD and the Hampel rule** — the median absolute deviation is a robust stand-in
for the standard deviation. Flagging at 3 × MAD catches outliers without letting
the outlier inflate the yardstick that is supposed to catch it, which is what
happens with an SD-based cutoff.

**Consensus outlier (`outlier_k = 3`)** — a sample is removed only when at least
three of the four independent checks agree. This is what keeps any single
pragmatic threshold from dropping a good sample by itself.

## Recap

- Stage 01 turns the raw spreadsheet into a filtered `DAList`: remove blood
  contamination and curated handling contaminants, drop consensus-outlier samples,
  then drop sparsely-detected proteins. 2,400 × 48 becomes 1,920 × 45.
- **45 samples is 48 minus three QC failures**, not a collection roster. `S29_T1`,
  `S28_T2` and `S28_T3` were taken and measured; they failed the 3-of-4 consensus.
- DIA zeros become `NA` because the gaps are mostly MNAR — "below detection", not
  "measured zero" — and that assumption is what every later imputation decision
  rests on.
- The missingness filter deliberately requires presence in only one `Group_Time`
  cell, because an all-cells rule would filter out the group-specific proteins the
  study exists to find.
- The six thresholds are pragmatic, tuned on this dataset, not derived from theory.
  The 3-of-4 consensus rule is what makes that acceptable for sample removal.
- Disclosed and unfixed: the search space was human-only with no contaminant FASTA
  and no decoys, so reagent contaminants **cannot be detected at all** and their
  absence is not evidence of absence. That one is not fixable by any row filter.

```{r sessioninfo}
sessionInfo()
```

## References

::: {#refs}
:::
