---
title: "HRvLR 01 — Filtering"
subtitle: "Building the DAList: contaminant, missingness, and outlier removal"
author: "A_HRvLR_2026"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
    embed-resources: true
    df-print: paged
bibliography: ../../shared/references.bib
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
pacman::p_load(here)
knitr::opts_knit$set(root.dir = here::here())
```

## What stage 01 does

Stage 01 turns the raw DIA-MS spreadsheet into a filtered `DAList`, the proteoDA
object every later stage builds on. It collapses duplicate accessions, removes
blood-plasma contamination, drops proteins too sparsely detected to model, and
removes any sample that fails a consensus of outlier checks.

This page runs that logic live and in memory, the same as
`01_Filtering/a_script/01_run_filtering.R`, and writes nothing. The last section
confirms the in-memory result matches the committed `DAList_filtered.rds`.

The design is repeated measures: high responders (HR) and low responders (LR),
each sampled at T1 (baseline), T2 (trained), and T3 (acute). The factor
`Group_Time` carries the six cells (`HR_T1` … `LR_T3`) and is the grouping column
for the missingness filter.

## Reads and writes

The stage reads three files from `00_input/`: the protein matrix
`HRvLR_raw.xlsx`, the sample sheet `HRvLR_meta.csv`, and the contaminant evidence
`HPA_annotations.tsv`. The runnable script writes `DAList_filtered.rds` (the
input to stage 02), `filtering_intermediates.rds` (reused by the stage-02 report
and the figure layer), `filtering_report.xlsx`, `hpa_absent_dropped.csv` (the
accessions with no HPA match), and a contaminant plot to `b_reports/`. This
document recomputes the object but saves none of it.

## Load the matrix and sample sheet {#sec-load}

The raw spreadsheet holds protein identity in leading annotation columns and one
intensity column per sample. The two blocks are split, and the intensity columns
are aligned to the sample sheet by `Col_ID`, which is what lets data and metadata
stay paired without a manual key.

```{r}
#| label: load
pacman::p_load(proteoDA, readxl, readr, dplyr, tidyr, stringr)

raw <- read_excel(here("00_input", "HRvLR_raw.xlsx"))
annot_cols <- c("uniprot_id", "protein", "gene", "description", "n_seq")
annotation <- raw[, annot_cols]
intensity <- raw[, setdiff(names(raw), annot_cols)]

metadata <- as.data.frame(read_csv(here("00_input", "HRvLR_meta.csv"), show_col_types = FALSE))
rownames(metadata) <- metadata$Col_ID
stopifnot(setequal(colnames(intensity), metadata$Col_ID))
intensity <- intensity[, metadata$Col_ID]
n_raw <- nrow(annotation)

c(proteins = n_raw, samples = ncol(intensity))
table(metadata$Group, metadata$Timepoint)
```

## Collapse duplicate accessions {#sec-dedup}

A few UniProt accessions appear on more than one row. Keeping the
highest-mean-intensity row leaves one measurement per protein, and picking by mean
intensity favours the better-quantified copy.

```{r}
#| label: dedup
if (any(duplicated(annotation$uniprot_id))) {
  rm_mean <- rowMeans(data.matrix(intensity), na.rm = TRUE)
  keep_idx <- tibble(i = seq_along(rm_mean), id = annotation$uniprot_id, m = rm_mean) |>
    group_by(id) |>
    slice_max(m, n = 1, with_ties = FALSE) |>
    pull(i)
  annotation <- annotation[keep_idx, ]
  intensity <- intensity[keep_idx, ]
}
nrow(annotation)
```

## Remove blood contaminants {#sec-contaminants}

A muscle biopsy carries blood, and plasma proteins (albumin, hemoglobins,
immunoglobulins) are abundant enough to bury the muscle signal. A protein is
flagged for removal when it looks blood-derived (HPA "Secreted to blood", an
immunoglobulin, or a strong erythrocyte marker) and does not clear the muscle
rescue: a high skeletal-myonuclei signal keeps it, unless its measured blood
concentration marks it as a genuine plasma protein. The rule is built from the
Human Protein Atlas table.

```{r}
#| label: classifier
ery_cut <- 5000
myo_cut <- 50
blood_max <- 1e9

bl <- read_tsv(here("00_input", "HPA_annotations.tsv"), show_col_types = FALSE) |>
  transmute(
    gene          = Gene,
    uniprot       = Uniprot,
    protein_class = `Protein class`,
    secretome     = `Secretome location`,
    blood_conc    = suppressWarnings(as.numeric(`Blood concentration - Conc. blood MS [pg/L]`)),
    ery           = suppressWarnings(as.numeric(`Single Cell Type RNA - Erythrocytes [nCPM]`)),
    myo           = suppressWarnings(as.numeric(`Single Cell Type RNA - Myonuclei [nCPM]`))
  ) |>
  mutate(
    secreted_blood = secretome == "Secreted to blood" & !is.na(secretome),
    is_ig = str_detect(coalesce(protein_class, ""), "Immunoglobulin genes"),
    is_erythrocyte = !is.na(ery) & ery >= ery_cut,
    muscle_keep = !is.na(myo) & myo >= myo_cut & (is.na(blood_conc) | blood_conc < blood_max),
    reason = case_when(
      muscle_keep & (secreted_blood | is_ig | is_erythrocyte) ~ "keep: muscle-expressed (rescued)",
      is_erythrocyte ~ "remove: erythrocyte (hemoglobin)",
      is_ig ~ "remove: immunoglobulin",
      secreted_blood ~ "remove: secreted-to-blood (plasma)",
      TRUE ~ "keep: not blood-associated"
    ),
    verdict = if_else(str_starts(reason, "remove"), "remove", "keep")
  )

table(bl$verdict)
```

The HPA table keys on UniProt accession, so the match is on accession, not gene
symbol. Each accession is stripped of its isoform suffix and HPA's comma-separated
multi-accession cells are split so every accession matches on its own. Removal is
plain subsetting applied to `annotation` and `intensity` together, so the two stay
aligned. The cascade keeps only proteins with an HPA accession match, logging the
dropped ones to `hpa_absent_dropped.csv`, then drops the blood-flagged ones.

```{r}
#| label: apply-contaminants
strip_iso <- function(x) sub("-\\d+$", "", x)
hpa_acc <- bl |>
  filter(!is.na(uniprot), uniprot != "") |>
  separate_longer_delim(uniprot, delim = ", ") |>
  mutate(acc = strip_iso(uniprot))
hpa_present <- unique(hpa_acc$acc)
remove_acc <- unique(hpa_acc$acc[hpa_acc$verdict == "remove"])

flog <- tibble(step = "Raw input", n_after = nrow(annotation), n_removed = NA_integer_)
acc <- strip_iso(annotation$uniprot_id)
keep <- acc %in% hpa_present
dropped_ids <- annotation$uniprot_id[!keep]
flog <- bind_rows(flog, tibble(step = "HPA presence", n_after = sum(keep), n_removed = sum(!keep)))
annotation <- annotation[keep, ]
intensity <- intensity[keep, ]
acc <- strip_iso(annotation$uniprot_id)
keep <- !(acc %in% remove_acc)
flog <- bind_rows(flog, tibble(step = "Blood contaminant removal", n_after = sum(keep), n_removed = sum(!keep)))
annotation <- annotation[keep, ]
intensity <- intensity[keep, ]

flog
```

The removed and rescued proteins, each with the HPA evidence behind the call:

```{r}
#| label: removed-rescued
bl |>
  filter(verdict == "remove") |>
  arrange(reason, desc(blood_conc)) |>
  select(gene, reason, blood_conc, ery, myo) |>
  head(8)

bl |>
  filter(str_detect(reason, "rescued")) |>
  arrange(desc(myo)) |>
  select(gene, reason, blood_conc, myo) |>
  head(6)
```

## Build the DAList {#sec-build}

A `DAList` holds three name-aligned tables: `data` (the intensity matrix),
`annotation` (one row per protein), and `metadata` (one row per sample). The row
names of `data` match `annotation`, and the column names of `data` match the row
names of `metadata`. `zero_to_missing()` converts DIA zeros to `NA`, because a
reported zero means not detected rather than a measured zero, and the downstream
filters and limma both treat absence as missing.

```{r}
#| label: build-dalist
int_mat <- as.data.frame(data.matrix(intensity))
rownames(int_mat) <- annotation$uniprot_id
annot_df <- as.data.frame(annotation)
rownames(annot_df) <- annotation$uniprot_id

dal <- zero_to_missing(DAList(data = int_mat, annotation = annot_df, metadata = metadata))

identical(rownames(dal$data), rownames(dal$annotation))
identical(colnames(dal$data), rownames(dal$metadata))
round(100 * mean(is.na(dal$data)), 1)
```

### Why a gap is not a zero: MAR and MNAR

That conversion deserves more than one line, because this is the stage where the
nature of the missingness matters most, and it is the assumption that every later
imputation choice rests on.

A value can be missing in two quite different ways [@lazar2016]. Under **MAR**
(missing at random) the gap has nothing to do with the protein's true abundance:
a peptide failed to be identified, a run hiccuped, the search engine did not
assign a spectrum. Whether the value is missing is unrelated to what the value
would have been, so the gap carries no information about the protein. Under
**MNAR** (missing not at random) the gap exists *because of* the value. The
protein was there, but too low to clear the instrument's detection limit, so it
was not reported. Here the gap is itself a measurement: it says "low".

In label-free DIA proteomics most gaps have an MNAR flavour, concentrated in the
low-intensity proteins where abundance and detection limit collide
[@webbrobertson2015]. Two consequences follow. First, `zero_to_missing()` is the
right move rather than a technicality: a reported zero would tell limma that we
measured this protein and it was nothing, which is a claim the instrument never
made, and it would drag the protein's mean down with a fabricated data point.
Missing is the honest encoding. Second, imputation downstream is not a neutral
housekeeping step. Filling MNAR gaps with a MAR-style estimate — a group mean, a
nearest-neighbour average — quietly invents a mid-range value for something whose
missingness meant "low", and it is choosing an answer, not recovering one. Stage
01 does not impute; it filters. It only makes sure the gaps are labelled honestly
so the later stages can decide what to do about them with their eyes open.

## Drop barely-detected proteins {#sec-missingness}

`filter_proteins_by_group()` is proteoDA's native missingness filter. With
`min_reps = 5, min_groups = 1` it keeps a protein detected in at least 5 of the 8
samples in at least one `Group_Time` cell.

`min_groups = 1` is the deliberate part, and it is the setting most likely to look
like a mistake. The stricter alternative — require the protein to clear `min_reps`
in *every* cell — sounds like better quality control, and it is precisely what we
must not do. A protein that is switched on in one responder group and off in the
other is exactly the biology this study is hunting. Under an all-cells rule it
would be deleted for being absent from the cells where its absence is the finding.
Demanding presence everywhere pre-selects for proteins that do not differ between
groups, which is a filter aimed directly at the hypothesis. Requiring the protein
to hold up in one cell keeps that signal while still removing proteins that are
nowhere consistently detected and cannot support a model anywhere.

```{r}
#| label: missingness
n0 <- nrow(dal$data)
dal <- filter_proteins_by_group(dal, min_reps = 5, min_groups = 1, grouping_column = "Group_Time")
flog <- bind_rows(flog, tibble(
  step = "Missingness (>=5 of 8 in >=1 group)",
  n_after = nrow(dal$data), n_removed = n0 - nrow(dal$data)
))
flog <- flog |> mutate(pct_of_raw = round(n_after / n_raw * 100, 1))
flog
```

## Remove outlier samples {#sec-outliers}

Whether a whole sample is untrustworthy is decided by four independent
heuristics: high missingness (with a within-subject arm comparing a subject's
three timepoints), off intensity level (Hampel 3×MAD), multivariate position in
PCA space, and low correlation with the other samples. Requiring at least three of
four to agree keeps any single noisy metric from dropping a good sample.

### The two statistics, in plain terms

**Mahalanobis distance** is how far a sample sits from the centre of the cloud,
measured in a way that respects the shape of the cloud. Plain Euclidean distance
treats the axes as independent and equally scaled, which is wrong when the axes
are correlated: a sample can look unremarkable on PC1 and unremarkable on PC2 and
still be somewhere no sample should be, because in this data PC1 and PC2 move
together and this one moves against the grain. Mahalanobis divides through by the
covariance, so distance is counted in units of how much the data actually vary in
that direction. Under multivariate normality it follows a chi-square distribution
with degrees of freedom equal to the number of dimensions, which is where the
`qchisq(1 - mahal_p, df = 3)` cutoff comes from.

Two honest caveats attach to it. The chi-square cutoff *assumes* PC1–PC3 are
approximately multivariate normal, and nobody has checked that here. And the
distance is computed on complete cases only — the PCA runs on proteins with no
missing values at all (`rowSums(is.na(...)) == 0`). That subset is not a random
sample of the proteome; it is the well-detected, abundant end of it. So the
outlier geometry is measured in a space that has already been selected, and a
sample whose problem lives entirely in the sparse proteins is a sample this test
cannot see.

**MAD and the Hampel rule.** The median absolute deviation is the median of the
absolute distances from the median — a robust stand-in for the standard deviation.
The reason to prefer it here is circular reasoning avoidance: an outlier inflates
the standard deviation, so an SD-based cutoff is widened by the very point it is
supposed to catch, and a bad enough sample hides itself. The median and the MAD
barely move when one value goes wild, so the yardstick stays honest. The Hampel
rule flags anything more than `mad_k` MADs from the median — here 3 — and it is
applied twice, to per-sample median intensity and to per-sample median correlation
with the other samples.

```{r}
#| label: outliers
mahal_p <- 0.01
mad_k <- 3
outlier_k <- 3

# missingness, with a within-subject spread across timepoints
pct_missing <- colMeans(is.na(dal$data)) * 100
miss_info <- dal$metadata |>
  select(Col_ID, Subject_ID, Group, Timepoint) |>
  mutate(pct_missing = pct_missing[Col_ID], prefix = str_remove(Col_ID, "_T[123]$")) |>
  group_by(prefix) |>
  mutate(delta_missing = sapply(pct_missing, \(x) max(abs(x - pct_missing)))) |>
  ungroup()
miss_thresh <- quantile(pct_missing, 0.75) + 1.5 * IQR(pct_missing)
delta_thresh <- quantile(miss_info$delta_missing, 0.75) + 1.5 * IQR(miss_info$delta_missing)
miss_info$miss_flag <- miss_info$pct_missing > miss_thresh | miss_info$delta_missing > delta_thresh

# PCA Mahalanobis distance on PC1-3 (complete proteins only)
complete_mat <- dal$data[rowSums(is.na(dal$data)) == 0, ]
pc3 <- prcomp(t(log2(complete_mat + 1)), center = TRUE, scale. = TRUE)$x[, 1:3]
mahal <- mahalanobis(pc3, colMeans(pc3), cov(pc3))
pca_flags <- tibble(Col_ID = colnames(dal$data), pca_flag = mahal > qchisq(1 - mahal_p, df = 3))

# Hampel rule on per-sample median intensity
samp_med <- apply(log2(dal$data + 1), 2, median, na.rm = TRUE)
mad_flags <- tibble(
  Col_ID = names(samp_med),
  mad_flag = abs(samp_med - median(samp_med)) > mad_k * mad(samp_med)
)

# low median correlation with the other samples
cor_mat <- cor(log2(dal$data + 1), use = "pairwise.complete.obs")
med_cor <- apply(cor_mat, 2, function(x) median(x[x < 1], na.rm = TRUE))
cor_flags <- tibble(
  Col_ID = names(med_cor),
  cor_flag = med_cor < median(med_cor) - mad_k * mad(med_cor)
)

outlier_diag <- miss_info |>
  left_join(pca_flags, by = "Col_ID") |>
  left_join(mad_flags, by = "Col_ID") |>
  left_join(cor_flags, by = "Col_ID") |>
  mutate(
    n_flags = miss_flag + pca_flag + mad_flag + cor_flag,
    consensus_outlier = n_flags >= outlier_k
  )
outlier_ids <- outlier_diag$Col_ID[outlier_diag$consensus_outlier]

head(outlier_diag[order(-outlier_diag$n_flags), c("Col_ID", "miss_flag", "pca_flag", "mad_flag", "cor_flag", "n_flags")], 6)
outlier_ids
```

```{r}
#| label: apply-outliers
if (length(outlier_ids)) dal <- filter_samples(dal, !(Col_ID %in% outlier_ids))
dim(dal$data)
```

## Where the numbers came from {#sec-thresholds}

Six constants decide what this stage keeps, and a reader is entitled to ask where
each one came from. The short answer is that they are pragmatic thresholds chosen
against this dataset, not values derived from theory. They are defensible, and
they are not principled. Both things are true and both should be said.

`ery_cut <- 5000` (erythrocyte nCPM). A protein is called an erythrocyte marker
when its HPA single-cell RNA in red cells is at least 5,000 nCPM. The number is
set where it does the job it exists to do: it catches the haemoglobins, which sit
far above any plausible cutoff, without sweeping in proteins that merely happen to
be transcribed in red cells at ordinary levels. It is a cut placed in a gap, not a
cut derived from a null distribution.

`myo_cut <- 50` (myonuclei nCPM). The floor for the muscle rescue. A protein
flagged as blood-derived gets a second hearing if its skeletal-myonuclei RNA
clears 50 nCPM, on the reasoning that muscle genuinely transcribes it and its
presence in a muscle biopsy needs no blood to explain it. Fifty is a modest bar,
chosen to be generous to muscle, because the cost of wrongly deleting a real
muscle protein is higher here than the cost of keeping a doubtful one.

`blood_max <- 1e9` (pg/L). The veto on the rescue. Without it, a protein that is
both transcribed in muscle *and* genuinely abundant in plasma would be rescued on
the strength of its muscle transcription alone, and its intensity in our matrix
would still be mostly blood. Any protein whose measured blood concentration is at
or above 10⁹ pg/L is refused the rescue no matter how muscle-expressed it is. This
is the one threshold with a clear logical job: it stops the rescue rule from
undoing the contaminant rule.

`mahal_p <- 0.01`. The chi-square tail probability for the PCA flag. It is a
conventional 1% rather than a calibrated error rate, and, as above, it rests on a
normality assumption nobody has tested.

`mad_k <- 3`. The Hampel constant. Three MADs is the standard convention, roughly
comparable to 3 SD under normality, and it is used unchanged for both the
intensity flag and the correlation flag.

`outlier_k <- 3`. Three of the four checks must agree before a sample is removed.
This is the setting that carries the weight, and it is the reason the arbitrariness
of the other five is survivable. Every individual threshold above is a judgement
call, and any one of them could be wrong at the margin. But a sample is only
dropped when three independent heuristics — built on different statistics, looking
at different failure modes — all point at it. A single mis-set cutoff cannot
delete a good sample on its own; it would need two other methods to agree with it.
The consensus rule is the safeguard, and it is what lets the individual numbers be
pragmatic without being dangerous.

## Known limitations {#sec-limitations}

Three things this stage does not do. None of them is fixed here; they are disclosed
so that a reader knows what is in the matrix.

**There is no contaminant-FASTA or decoy removal.** Stage 01 has no cRAP list, no
common-contaminant database, and no reverse-hit or decoy filter anywhere in it.
The only contaminant logic is the blood rule, and the blood rule asks exactly three
questions: is it secreted to blood, is it an immunoglobulin, is it high in
erythrocytes. Keratins are none of those, so they sail through. KRT1, KRT2, KRT10
and KRT8 are all present in the final 1,896-protein matrix, and KRT1, KRT2 and
KRT10 are the three canonical skin and handling contaminants that sit at the top of
every cRAP list ever written. Their abundance in a biopsy is a fact about who
handled the tube, not about muscle. They are being modelled as biology.

**HBG2 survives, and it should not.** Fetal haemoglobin gamma-2 has an HPA blood
concentration of 7 g/L, which makes it a plasma protein by any reasonable reading.
It survives because of how the erythrocyte test is built: the test keys on
*adult*-erythrocyte single-cell RNA, and adult red cells do not transcribe the
fetal chain, so HBG2's erythrocyte nCPM is 0 and the test never fires. HBB and
HBA1, whose adult transcription is enormous, were caught correctly. The gap is
that the rule reasons about transcription in a cell type rather than about the
protein's own abundance in plasma, and HBG2 is exactly the protein that falls
through it.

**The HPA-presence filter is load-bearing and un-provenanced.** The line
`keep <- acc %in% hpa_present` drops every protein with no matching accession in
`HPA_annotations.tsv`. That is 127 proteins removed, not for anything they did, but
for being absent from a reference file. And that file is a partial export with no
recorded version, no query string, and no retrieval date — there is no
`PROVENANCE.md` beside it. A filter that removes 127 real measurements on the
authority of an undocumented file is a reproducibility hole: a future export of the
HPA table would silently produce a different matrix, and nobody could say why. It
should be provenance-stamped.

## Result and verification {#sec-recap}

proteoDA records every operation in the `tags` slot. Stage 01 reduces the raw
`r n_raw`-protein spreadsheet to a `r nrow(dal$data)` × `r ncol(dal$data)`
`DAList` (proteins × samples). The in-memory result matches the committed
`DAList_filtered.rds`.

```{r}
#| label: verify
names(dal$tags)
dim(dal$data)

committed <- readRDS(here("01_Filtering", "c_data", "DAList_filtered.rds"))
c(
  same_dim = identical(dim(dal$data), dim(committed$data)),
  same_proteins = identical(sort(rownames(dal$data)), sort(rownames(committed$data))),
  same_samples = identical(sort(colnames(dal$data)), sort(colnames(committed$data)))
)
```

Two proteoDA habits carry through the rest of the pipeline. A `DAList` is three
name-aligned tables, so the native filters preserve alignment and any
hand-subsetting moves `data` and `annotation` together. DIA zeros are missing
values, not measured zeros. Stage 02 picks up `DAList_filtered.rds` and
normalizes it.

## Key terms

**`DAList`** — proteoDA's container [@thurman2023]: three name-aligned tables,
`data` (intensities), `annotation` (one row per protein), `metadata` (one row per
sample). The alignment is enforced, so a filter can never separate a protein from
its annotation.

**MAR (missing at random)** — the gap is unrelated to the protein's true
abundance; a random identification failure. The missingness carries no information
about the value.

**MNAR (missing not at random)** — the protein is missing *because* it fell below
the detection limit. The gap is informative: it means "low". Most DIA gaps are
MNAR-flavoured, which is why a zero must become `NA` and why imputation choice
matters downstream [@lazar2016; @webbrobertson2015].

**`min_groups = 1`** — a protein survives the missingness filter if it clears
`min_reps` in any one `Group_Time` cell. Requiring every cell would delete
proteins that are on in one responder group and off in the other, which is the
biology under investigation.

**Mahalanobis distance** — distance from the centroid measured in units of the
data's own covariance. It catches a sample that is unexceptional on each PC
separately but sits in an impossible place in the correlated space. Assumes
approximate multivariate normality of PC1–PC3, and here it is computed on complete
proteins only, which is itself a selected subset.

**MAD and the Hampel rule** — the median absolute deviation is a robust stand-in
for the standard deviation. Flagging at 3 × MAD catches outliers without letting
the outlier inflate the yardstick that is supposed to catch it, which is what
happens with an SD-based cutoff.

**Consensus outlier (`outlier_k = 3`)** — a sample is removed only when at least
three of the four independent checks agree. This is what keeps any single
pragmatic threshold from dropping a good sample by itself.

## Recap

- Stage 01 turns the raw spreadsheet into a filtered `DAList`: deduplicate
  accessions, remove blood contamination, drop sparsely-detected proteins, drop
  consensus-outlier samples.
- DIA zeros become `NA` because the gaps are mostly MNAR — "below detection", not
  "measured zero" — and that assumption is what every later imputation decision
  rests on.
- The missingness filter deliberately requires presence in only one `Group_Time`
  cell, because an all-cells rule would filter out the group-specific proteins the
  study exists to find.
- The six thresholds are pragmatic, tuned on this dataset, not derived from theory.
  The 3-of-4 consensus rule is what makes that acceptable for sample removal.
- Disclosed and unfixed: no cRAP or decoy removal, so KRT1/KRT2/KRT10 and KRT8
  remain; HBG2 escapes the erythrocyte test because that test keys on adult-red-cell
  RNA rather than plasma abundance; and the HPA-presence filter drops 127 proteins
  on the authority of an export with no recorded provenance.

```{r sessioninfo}
sessionInfo()
```

## References

::: {#refs}
:::
