---
title: "HRvLR 01 — Filtering"
subtitle: "Building the DAList: contaminant, missingness, and outlier removal"
author: "A_HRvLR_2026"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
    embed-resources: true
    df-print: paged
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
pacman::p_load(here)
knitr::opts_knit$set(root.dir = here::here())
```

## What stage 01 does

Stage 01 turns the raw DIA-MS spreadsheet into a filtered `DAList`, the proteoDA
object every later stage builds on. It collapses duplicate accessions, removes
blood-plasma contamination, drops proteins too sparsely detected to model, and
removes any sample that fails a consensus of outlier checks.

This page runs that logic live and in memory, the same as
`01_Filtering/a_script/01_run_filtering.R`, and writes nothing. The last section
confirms the in-memory result matches the committed `DAList_filtered.rds`.

The design is repeated measures: high responders (HR) and low responders (LR),
each sampled at T1 (baseline), T2 (trained), and T3 (acute). The factor
`Group_Time` carries the six cells (`HR_T1` … `LR_T3`) and is the grouping column
for the missingness filter.

## Reads and writes

The stage reads three files from `00_input/`: the protein matrix
`HRvLR_raw.xlsx`, the sample sheet `HRvLR_meta.csv`, and the contaminant evidence
`HPA_annotations.tsv`. The runnable script writes `DAList_filtered.rds` (the
input to stage 02), `filtering_intermediates.rds` (reused by the stage-02 report
and the figure layer), `filtering_report.xlsx`, `hpa_absent_dropped.csv` (the
accessions with no HPA match), and a contaminant plot to `b_reports/`. This
document recomputes the object but saves none of it.

## Load the matrix and sample sheet {#sec-load}

The raw spreadsheet holds protein identity in leading annotation columns and one
intensity column per sample. The two blocks are split, and the intensity columns
are aligned to the sample sheet by `Col_ID`, which is what lets data and metadata
stay paired without a manual key.

```{r}
#| label: load
pacman::p_load(proteoDA, readxl, readr, dplyr, tidyr, stringr)

raw <- read_excel(here("00_input", "HRvLR_raw.xlsx"))
annot_cols <- c("uniprot_id", "protein", "gene", "description", "n_seq")
annotation <- raw[, annot_cols]
intensity <- raw[, setdiff(names(raw), annot_cols)]

metadata <- as.data.frame(read_csv(here("00_input", "HRvLR_meta.csv"), show_col_types = FALSE))
rownames(metadata) <- metadata$Col_ID
stopifnot(setequal(colnames(intensity), metadata$Col_ID))
intensity <- intensity[, metadata$Col_ID]
n_raw <- nrow(annotation)

c(proteins = n_raw, samples = ncol(intensity))
table(metadata$Group, metadata$Timepoint)
```

## Collapse duplicate accessions {#sec-dedup}

A few UniProt accessions appear on more than one row. Keeping the
highest-mean-intensity row leaves one measurement per protein, and picking by mean
intensity favours the better-quantified copy.

```{r}
#| label: dedup
if (any(duplicated(annotation$uniprot_id))) {
  rm_mean <- rowMeans(data.matrix(intensity), na.rm = TRUE)
  keep_idx <- tibble(i = seq_along(rm_mean), id = annotation$uniprot_id, m = rm_mean) |>
    group_by(id) |>
    slice_max(m, n = 1, with_ties = FALSE) |>
    pull(i)
  annotation <- annotation[keep_idx, ]
  intensity <- intensity[keep_idx, ]
}
nrow(annotation)
```

## Remove blood contaminants {#sec-contaminants}

A muscle biopsy carries blood, and plasma proteins (albumin, hemoglobins,
immunoglobulins) are abundant enough to bury the muscle signal. A protein is
flagged for removal when it looks blood-derived (HPA "Secreted to blood", an
immunoglobulin, or a strong erythrocyte marker) and does not clear the muscle
rescue: a high skeletal-myonuclei signal keeps it, unless its measured blood
concentration marks it as a genuine plasma protein. The rule is built from the
Human Protein Atlas table.

```{r}
#| label: classifier
ery_cut <- 5000
myo_cut <- 50
blood_max <- 1e9

bl <- read_tsv(here("00_input", "HPA_annotations.tsv"), show_col_types = FALSE) |>
  transmute(
    gene          = Gene,
    uniprot       = Uniprot,
    protein_class = `Protein class`,
    secretome     = `Secretome location`,
    blood_conc    = suppressWarnings(as.numeric(`Blood concentration - Conc. blood MS [pg/L]`)),
    ery           = suppressWarnings(as.numeric(`Single Cell Type RNA - Erythrocytes [nCPM]`)),
    myo           = suppressWarnings(as.numeric(`Single Cell Type RNA - Myonuclei [nCPM]`))
  ) |>
  mutate(
    secreted_blood = secretome == "Secreted to blood" & !is.na(secretome),
    is_ig = str_detect(coalesce(protein_class, ""), "Immunoglobulin genes"),
    is_erythrocyte = !is.na(ery) & ery >= ery_cut,
    muscle_keep = !is.na(myo) & myo >= myo_cut & (is.na(blood_conc) | blood_conc < blood_max),
    reason = case_when(
      muscle_keep & (secreted_blood | is_ig | is_erythrocyte) ~ "keep: muscle-expressed (rescued)",
      is_erythrocyte ~ "remove: erythrocyte (hemoglobin)",
      is_ig ~ "remove: immunoglobulin",
      secreted_blood ~ "remove: secreted-to-blood (plasma)",
      TRUE ~ "keep: not blood-associated"
    ),
    verdict = if_else(str_starts(reason, "remove"), "remove", "keep")
  )

table(bl$verdict)
```

The HPA table keys on UniProt accession, so the match is on accession, not gene
symbol. Each accession is stripped of its isoform suffix and HPA's comma-separated
multi-accession cells are split so every accession matches on its own. Removal is
plain subsetting applied to `annotation` and `intensity` together, so the two stay
aligned. The cascade keeps only proteins with an HPA accession match, logging the
dropped ones to `hpa_absent_dropped.csv`, then drops the blood-flagged ones.

```{r}
#| label: apply-contaminants
strip_iso <- function(x) sub("-\\d+$", "", x)
hpa_acc <- bl |>
  filter(!is.na(uniprot), uniprot != "") |>
  separate_longer_delim(uniprot, delim = ", ") |>
  mutate(acc = strip_iso(uniprot))
hpa_present <- unique(hpa_acc$acc)
remove_acc <- unique(hpa_acc$acc[hpa_acc$verdict == "remove"])

flog <- tibble(step = "Raw input", n_after = nrow(annotation), n_removed = NA_integer_)
acc <- strip_iso(annotation$uniprot_id)
keep <- acc %in% hpa_present
dropped_ids <- annotation$uniprot_id[!keep]
flog <- bind_rows(flog, tibble(step = "HPA presence", n_after = sum(keep), n_removed = sum(!keep)))
annotation <- annotation[keep, ]
intensity <- intensity[keep, ]
acc <- strip_iso(annotation$uniprot_id)
keep <- !(acc %in% remove_acc)
flog <- bind_rows(flog, tibble(step = "Blood contaminant removal", n_after = sum(keep), n_removed = sum(!keep)))
annotation <- annotation[keep, ]
intensity <- intensity[keep, ]

flog
```

The removed and rescued proteins, each with the HPA evidence behind the call:

```{r}
#| label: removed-rescued
bl |>
  filter(verdict == "remove") |>
  arrange(reason, desc(blood_conc)) |>
  select(gene, reason, blood_conc, ery, myo) |>
  head(8)

bl |>
  filter(str_detect(reason, "rescued")) |>
  arrange(desc(myo)) |>
  select(gene, reason, blood_conc, myo) |>
  head(6)
```

## Build the DAList {#sec-build}

A `DAList` holds three name-aligned tables: `data` (the intensity matrix),
`annotation` (one row per protein), and `metadata` (one row per sample). The row
names of `data` match `annotation`, and the column names of `data` match the row
names of `metadata`. `zero_to_missing()` converts DIA zeros to `NA`, because a
reported zero means not detected rather than a measured zero, and the downstream
filters and limma both treat absence as missing.

```{r}
#| label: build-dalist
int_mat <- as.data.frame(data.matrix(intensity))
rownames(int_mat) <- annotation$uniprot_id
annot_df <- as.data.frame(annotation)
rownames(annot_df) <- annotation$uniprot_id

dal <- zero_to_missing(DAList(data = int_mat, annotation = annot_df, metadata = metadata))

identical(rownames(dal$data), rownames(dal$annotation))
identical(colnames(dal$data), rownames(dal$metadata))
round(100 * mean(is.na(dal$data)), 1)
```

## Drop barely-detected proteins {#sec-missingness}

`filter_proteins_by_group()` is proteoDA's native missingness filter. With
`min_reps = 5, min_groups = 1` it keeps a protein detected in at least 5 of the 8
samples in at least one `Group_Time` cell. Requiring presence in only one cell is
deliberate: a protein switched on in one responder group and off in the other
carries real signal and is kept rather than filtered as sparse.

```{r}
#| label: missingness
n0 <- nrow(dal$data)
dal <- filter_proteins_by_group(dal, min_reps = 5, min_groups = 1, grouping_column = "Group_Time")
flog <- bind_rows(flog, tibble(
  step = "Missingness (>=5 of 8 in >=1 group)",
  n_after = nrow(dal$data), n_removed = n0 - nrow(dal$data)
))
flog <- flog |> mutate(pct_of_raw = round(n_after / n_raw * 100, 1))
flog
```

## Remove outlier samples {#sec-outliers}

Whether a whole sample is untrustworthy is decided by four independent
heuristics: high missingness (with a within-subject arm comparing a subject's
three timepoints), off intensity level (Hampel 3×MAD), multivariate position in
PCA space, and low correlation with the other samples. Requiring at least three of
four to agree keeps any single noisy metric from dropping a good sample.

```{r}
#| label: outliers
mahal_p <- 0.01
mad_k <- 3
outlier_k <- 3

# missingness, with a within-subject spread across timepoints
pct_missing <- colMeans(is.na(dal$data)) * 100
miss_info <- dal$metadata |>
  select(Col_ID, Subject_ID, Group, Timepoint) |>
  mutate(pct_missing = pct_missing[Col_ID], prefix = str_remove(Col_ID, "_T[123]$")) |>
  group_by(prefix) |>
  mutate(delta_missing = sapply(pct_missing, \(x) max(abs(x - pct_missing)))) |>
  ungroup()
miss_thresh <- quantile(pct_missing, 0.75) + 1.5 * IQR(pct_missing)
delta_thresh <- quantile(miss_info$delta_missing, 0.75) + 1.5 * IQR(miss_info$delta_missing)
miss_info$miss_flag <- miss_info$pct_missing > miss_thresh | miss_info$delta_missing > delta_thresh

# PCA Mahalanobis distance on PC1-3 (complete proteins only)
complete_mat <- dal$data[rowSums(is.na(dal$data)) == 0, ]
pc3 <- prcomp(t(log2(complete_mat + 1)), center = TRUE, scale. = TRUE)$x[, 1:3]
mahal <- mahalanobis(pc3, colMeans(pc3), cov(pc3))
pca_flags <- tibble(Col_ID = colnames(dal$data), pca_flag = mahal > qchisq(1 - mahal_p, df = 3))

# Hampel rule on per-sample median intensity
samp_med <- apply(log2(dal$data + 1), 2, median, na.rm = TRUE)
mad_flags <- tibble(
  Col_ID = names(samp_med),
  mad_flag = abs(samp_med - median(samp_med)) > mad_k * mad(samp_med)
)

# low median correlation with the other samples
cor_mat <- cor(log2(dal$data + 1), use = "pairwise.complete.obs")
med_cor <- apply(cor_mat, 2, function(x) median(x[x < 1], na.rm = TRUE))
cor_flags <- tibble(
  Col_ID = names(med_cor),
  cor_flag = med_cor < median(med_cor) - mad_k * mad(med_cor)
)

outlier_diag <- miss_info |>
  left_join(pca_flags, by = "Col_ID") |>
  left_join(mad_flags, by = "Col_ID") |>
  left_join(cor_flags, by = "Col_ID") |>
  mutate(
    n_flags = miss_flag + pca_flag + mad_flag + cor_flag,
    consensus_outlier = n_flags >= outlier_k
  )
outlier_ids <- outlier_diag$Col_ID[outlier_diag$consensus_outlier]

head(outlier_diag[order(-outlier_diag$n_flags), c("Col_ID", "miss_flag", "pca_flag", "mad_flag", "cor_flag", "n_flags")], 6)
outlier_ids
```

```{r}
#| label: apply-outliers
if (length(outlier_ids)) dal <- filter_samples(dal, !(Col_ID %in% outlier_ids))
dim(dal$data)
```

## Result and verification {#sec-recap}

proteoDA records every operation in the `tags` slot. Stage 01 reduces the raw
`r n_raw`-protein spreadsheet to a `r nrow(dal$data)` × `r ncol(dal$data)`
`DAList` (proteins × samples). The in-memory result matches the committed
`DAList_filtered.rds`.

```{r}
#| label: verify
names(dal$tags)
dim(dal$data)

committed <- readRDS(here("01_Filtering", "c_data", "DAList_filtered.rds"))
c(
  same_dim = identical(dim(dal$data), dim(committed$data)),
  same_proteins = identical(sort(rownames(dal$data)), sort(rownames(committed$data))),
  same_samples = identical(sort(colnames(dal$data)), sort(colnames(committed$data)))
)
```

Two proteoDA habits carry through the rest of the pipeline. A `DAList` is three
name-aligned tables, so the native filters preserve alignment and any
hand-subsetting moves `data` and `annotation` together. DIA zeros are missing
values, not measured zeros. Stage 02 picks up `DAList_filtered.rds` and
normalizes it.
