---
title: "HRvLR 01 — Filtering"
subtitle: "The DAList object; contaminant, missingness, and outlier removal"
author: "A_HRvLR_2026"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
    embed-resources: true
    df-print: paged
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(here)
knitr::opts_knit$set(root.dir = here::here())
```

This document reproduces stage 01 of the HRvLR pipeline and shows the result of
each step. It runs the same logic as `01_Filtering/a_script/01_run_filtering.R`,
but entirely in memory: it computes the filtered object live and writes no files.
The final section confirms the live result matches the committed
`DAList_filtered.rds`.

The study is a 2×3 design: high responders (HR) and low responders (LR), each
sampled at three timepoints (T1 baseline, T2 trained, T3 acute). The grouping
factor `Group_Time` carries the six cells (`HR_T1` … `LR_T3`).

# Load the raw matrix and sample sheet {#sec-load}

Stage 01 begins with a raw DIA-MS spreadsheet (`00_input/HRvLR_raw.xlsx`). The
annotation columns (protein identity) are separated from the sample-intensity
columns. The sample sheet `00_input/HRvLR_meta.csv` supplies one row per sample,
including `Group_Time`. The matrix is aligned to the sheet by `Col_ID`.

```{r}
#| label: load
library(proteoDA)
library(readxl)
library(readr)
library(dplyr)
library(tidyr)
library(stringr)

raw <- read_excel(here("00_input", "HRvLR_raw.xlsx"))
annot_cols <- c("uniprot_id", "protein", "gene", "description", "n_seq")
annotation <- raw[, annot_cols]
intensity <- raw[, setdiff(names(raw), annot_cols)]

metadata <- as.data.frame(read_csv(here("00_input", "HRvLR_meta.csv"), show_col_types = FALSE))
rownames(metadata) <- metadata$Col_ID
stopifnot(setequal(colnames(intensity), metadata$Col_ID))
intensity <- intensity[, metadata$Col_ID] # errors if matrix/metadata IDs disagree
n_raw <- nrow(annotation)

c(proteins = n_raw, samples = ncol(intensity))
table(metadata$Group, metadata$Timepoint) # eight subjects per responder x timepoint cell
```

# Step 1: collapse duplicate accessions {#sec-dedup}

A few UniProt accessions appear on more than one row. The highest-mean-intensity
row is kept for each, so the matrix carries one measurement per protein:

```{r}
#| label: dedup
if (any(duplicated(annotation$uniprot_id))) {
  rm_mean <- rowMeans(data.matrix(intensity), na.rm = TRUE)
  keep_idx <- tibble(i = seq_along(rm_mean), id = annotation$uniprot_id, m = rm_mean) |>
    group_by(id) |>
    slice_max(m, n = 1, with_ties = FALSE) |>
    pull(i)
  annotation <- annotation[keep_idx, ]
  intensity <- intensity[keep_idx, ]
}
nrow(annotation)
```

# Step 2: remove blood contaminants {#sec-contaminants}

Muscle biopsies carry blood, and plasma proteins (albumin, hemoglobins,
immunoglobulins) are abundant enough to drown out the muscle signal. A protein is
flagged for removal if it looks like blood (HPA "Secreted to blood", an
immunoglobulin, or a strong erythrocyte signal) **and** does not look like muscle
(rescued by a strong skeletal-myonuclei signal, unless its measured blood
concentration marks it as a true plasma protein). The classifier is built from
the Human Protein Atlas annotations:

```{r}
#| label: classifier
ery_cut <- 5000
myo_cut <- 50
blood_max <- 1e9

bl <- read_tsv(here("00_input", "HPA_annotations.tsv"), show_col_types = FALSE) |>
  transmute(
    gene          = Gene,
    uniprot       = Uniprot,
    protein_class = `Protein class`,
    secretome     = `Secretome location`,
    blood_conc    = suppressWarnings(as.numeric(`Blood concentration - Conc. blood MS [pg/L]`)),
    ery           = suppressWarnings(as.numeric(`Single Cell Type RNA - Erythrocytes [nCPM]`)),
    myo           = suppressWarnings(as.numeric(`Single Cell Type RNA - Myonuclei [nCPM]`))
  ) |>
  mutate(
    secreted_blood = secretome == "Secreted to blood" & !is.na(secretome),
    is_ig = str_detect(coalesce(protein_class, ""), "Immunoglobulin genes"),
    is_erythrocyte = !is.na(ery) & ery >= ery_cut,
    muscle_keep = !is.na(myo) & myo >= myo_cut & (is.na(blood_conc) | blood_conc < blood_max),
    reason = case_when(
      muscle_keep & (secreted_blood | is_ig | is_erythrocyte) ~ "keep: muscle-expressed (rescued)",
      is_erythrocyte ~ "remove: erythrocyte (hemoglobin)",
      is_ig ~ "remove: immunoglobulin",
      secreted_blood ~ "remove: secreted-to-blood (plasma)",
      TRUE ~ "keep: not blood-associated"
    ),
    verdict = if_else(str_starts(reason, "remove"), "remove", "keep")
  )

table(bl$verdict) # how the HPA genes split
```

Removal is plain subsetting, applied to `annotation` and `intensity` together so
they stay aligned. The cascade keeps only proteins present in HPA, then drops the
blood-flagged ones; each step is logged:

```{r}
#| label: apply-contaminants
hpa_genes <- unique(bl$gene)
remove_genes <- bl$gene[bl$verdict == "remove"]

flog <- tibble(step = "Raw input", n_after = nrow(annotation), n_removed = NA_integer_)
keep <- annotation$gene %in% hpa_genes
flog <- bind_rows(flog, tibble(step = "HPA presence", n_after = sum(keep), n_removed = sum(!keep)))
annotation <- annotation[keep, ]
intensity <- intensity[keep, ]
keep <- !(annotation$gene %in% remove_genes)
flog <- bind_rows(flog, tibble(step = "Blood contaminant removal", n_after = sum(keep), n_removed = sum(!keep)))
annotation <- annotation[keep, ]
intensity <- intensity[keep, ]

flog
```

The removed and rescued proteins, with the HPA evidence behind each call:

```{r}
#| label: removed-rescued
bl |>
  filter(verdict == "remove") |>
  arrange(reason, desc(blood_conc)) |>
  select(gene, reason, blood_conc, ery, myo) |>
  head(8)

bl |>
  filter(str_detect(reason, "rescued")) |>
  arrange(desc(myo)) |>
  select(gene, reason, blood_conc, myo) |>
  head(6)
```

# Build the `DAList` {#sec-build}

proteoDA operates on a `DAList` — a named list holding three name-aligned tables:
`data` (intensity matrix), `annotation` (one row per protein), and `metadata`
(one row per sample). The row names of `data` equal the row names of
`annotation`, and the column names of `data` equal the row names of `metadata`.
`zero_to_missing()` turns DIA zeros into `NA`, since a reported zero means "not
detected" rather than a measured zero:

```{r}
#| label: build-dalist
int_mat <- as.data.frame(data.matrix(intensity))
rownames(int_mat) <- annotation$uniprot_id
annot_df <- as.data.frame(annotation)
rownames(annot_df) <- annotation$uniprot_id

dal <- zero_to_missing(DAList(data = int_mat, annotation = annot_df, metadata = metadata))

identical(rownames(dal$data), rownames(dal$annotation)) # proteins aligned
identical(colnames(dal$data), rownames(dal$metadata)) # samples aligned
round(100 * mean(is.na(dal$data)), 1) # percent missing after zero -> NA
```

# Step 3: drop barely-detected proteins {#sec-missingness}

`filter_proteins_by_group()` is the native proteoDA filter. With `min_reps = 5,
min_groups = 1` it keeps a protein detected in at least 5 of the 8 samples in **at
least one** `Group_Time` cell — a protein switched on in one responder group but
off in the other is kept, since detection is not required everywhere:

```{r}
#| label: missingness
n0 <- nrow(dal$data)
dal <- filter_proteins_by_group(dal, min_reps = 5, min_groups = 1, grouping_column = "Group_Time")
flog <- bind_rows(flog, tibble(
  step = "Missingness (>=5 of 8 in >=1 group)",
  n_after = nrow(dal$data), n_removed = n0 - nrow(dal$data)
))
flog <- flog |> mutate(pct_of_raw = round(n_after / n_raw * 100, 1))
flog # full cascade, raw -> final, with % of raw retained
```

# Step 4: detect and remove outlier samples {#sec-outliers}

Whether a whole sample is untrustworthy is decided by a consensus of four
independent heuristics — high missingness (with a within-subject arm comparing the
three timepoints of one subject), off intensity level (Hampel 3×MAD), multivariate
position in PCA space, and low correlation with the other samples. A sample is
dropped only if at least **three of four** agree:

```{r}
#| label: outliers
mahal_p <- 0.01
mad_k <- 3
outlier_k <- 3

# Method 1: sample missingness, plus a within-subject spread across timepoints
pct_missing <- colMeans(is.na(dal$data)) * 100
miss_info <- dal$metadata |>
  select(Col_ID, Subject_ID, Group, Timepoint) |>
  mutate(pct_missing = pct_missing[Col_ID], prefix = str_remove(Col_ID, "_T[123]$")) |>
  group_by(prefix) |>
  mutate(delta_missing = sapply(pct_missing, \(x) max(abs(x - pct_missing)))) |>
  ungroup()
miss_thresh <- quantile(pct_missing, 0.75) + 1.5 * IQR(pct_missing) # Tukey upper fence
delta_thresh <- quantile(miss_info$delta_missing, 0.75) + 1.5 * IQR(miss_info$delta_missing)
miss_info$miss_flag <- miss_info$pct_missing > miss_thresh | miss_info$delta_missing > delta_thresh

# Method 2: PCA Mahalanobis distance on PC1-3 (complete proteins only)
complete_mat <- dal$data[rowSums(is.na(dal$data)) == 0, ]
pc3 <- prcomp(t(log2(complete_mat + 1)), center = TRUE, scale. = TRUE)$x[, 1:3]
mahal <- mahalanobis(pc3, colMeans(pc3), cov(pc3))
pca_flags <- tibble(Col_ID = colnames(dal$data), pca_flag = mahal > qchisq(1 - mahal_p, df = 3))

# Method 3: Hampel rule on per-sample median intensity
samp_med <- apply(log2(dal$data + 1), 2, median, na.rm = TRUE)
mad_flags <- tibble(
  Col_ID = names(samp_med),
  mad_flag = abs(samp_med - median(samp_med)) > mad_k * mad(samp_med)
)

# Method 4: low median correlation with the other samples
cor_mat <- cor(log2(dal$data + 1), use = "pairwise.complete.obs")
med_cor <- apply(cor_mat, 2, function(x) median(x[x < 1], na.rm = TRUE)) # x<1 drops self
cor_flags <- tibble(
  Col_ID = names(med_cor),
  cor_flag = med_cor < median(med_cor) - mad_k * mad(med_cor)
)

outlier_diag <- miss_info |>
  left_join(pca_flags, by = "Col_ID") |>
  left_join(mad_flags, by = "Col_ID") |>
  left_join(cor_flags, by = "Col_ID") |>
  mutate(
    n_flags = miss_flag + pca_flag + mad_flag + cor_flag,
    consensus_outlier = n_flags >= outlier_k
  )
outlier_ids <- outlier_diag$Col_ID[outlier_diag$consensus_outlier]

head(outlier_diag[order(-outlier_diag$n_flags), c("Col_ID", "miss_flag", "pca_flag", "mad_flag", "cor_flag", "n_flags")], 6)
outlier_ids # samples tripping >= 3 of 4 flags
```

```{r}
#| label: apply-outliers
if (length(outlier_ids)) dal <- filter_samples(dal, !(Col_ID %in% outlier_ids))
dim(dal$data)
```

# Result and verification {#sec-recap}

proteoDA logs every operation in the `tags` slot. Stage 01 turns a 2,400 × 48
spreadsheet into a clean **1,885 × 45** `DAList`. The live result matches the
committed `DAList_filtered.rds` exactly:

```{r}
#| label: verify
names(dal$tags)
dim(dal$data)

committed <- readRDS(here("01_Filtering", "c_data", "DAList_filtered.rds"))
c(
  same_dim = identical(dim(dal$data), dim(committed$data)),
  same_proteins = identical(sort(rownames(dal$data)), sort(rownames(committed$data))),
  same_samples = identical(sort(colnames(dal$data)), sort(colnames(committed$data)))
)
```

The side-effecting script additionally writes `filtering_report.xlsx` (the
classifier verdicts, filter cascade, and outlier diagnostics shown above), a
contaminant plot to `b_reports/`, and a `filtering_intermediates.rds` that the
stage-02 report and the figure layer reuse, then saves `DAList_filtered.rds` for
stage 02. Key points:

- A `DAList` is three name-aligned tables; keeping them in step is what proteoDA
  relies on. The native filters preserve the alignment; hand-subsetting must move
  `data` and `annotation` together.
- `zero_to_missing()` reflects that DIA zeros are missing values.
- `filter_proteins_by_group()` and `filter_samples()` are the native filters.

Stage 02 picks up this file and normalizes it.
