---
title: "01 · Filtering"
subtitle: "Protein report to a filtered, unnormalised DAList"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    df-print: kable
    fig-width: 7
    fig-height: 5
    fig-dpi: 150
    fig-align: center
execute:
  message: false
---

The report arrives at protein level, one row per accession and one column per MS run. This
notebook removes contaminant proteins by identity, drops outlier samples by a four-method
consensus, and filters on detection last, on the samples that remain. Contaminants leave before
`02_Normalization` because cyclic loess estimates each sample's reference from that sample's own
intensity distribution.

```{r setup}
library(here)
library(proteoDA)
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(purrr)
library(readr)
library(readxl)
library(ggplot2)
library(forcats)
library(patchwork)
library(writexl)

out <- here("01_Preprocess", "01_Filtering", "c_data")
dir.create(out, recursive = TRUE, showWarnings = FALSE)

strip_iso <- function(x) sub("-[0-9]+$", "", x)
```

## Thresholds

Each value was set from annotation alone. No contrast, phenotype or group label was consulted.

```{r config}
cfg <- list(
  miss_min_reps = 5, # detected samples a Group_Time cell needs
  miss_min_groups = 1, # cells that must clear miss_min_reps
  cor_min_obs = 15, # samples a protein needs before blood_cor is computed
  outlier_k = 3, # outlier methods that must agree, of four
  mahal_p = 0.01, # PCA Mahalanobis chi-square tail
  mad_k = 3, # Hampel constant
  ery_cut = 5000, # erythrocyte nCPM at or above: red-cell protein
  myo_cut = 20, # myonuclei nCPM at or above: candidate for muscle rescue
  blood_max = 1e9, # rescue only below this plasma concentration, pg/L
  blood_anchor = c("HBB", "HBA1", "HBD", "HBG1", "HBG2")
)
```

`myo_cut` is 20 rather than 50. HPA tags every plasma-detectable protein "Secreted to blood",
which is why creatine kinase and myoglobin are clinical damage markers, and the rescue exists to
undo that tag. At 50 it missed GPI (myonuclei 43.1), ANXA2 (40.2) and PPIA (21.1). The
transcript floor does not separate plasma from muscle on its own: removed plasma proteins span
myonuclei 0 to 237 and rescued ones 21 to 362. `blood_max` does the separating. The eight
removed proteins above `myo_cut` all measure 2.3e9 to 4.2e11 pg/L in plasma; intracellular
proteins measure 1e6 to 1e8.

The floor over-deletes five real proteins: C1QBP, HMGB2, PPIB, LGALS1 and CTSB clear
`blood_max` and fail only on myonuclei. V1 refitted the nine contrasts with all 120
blood-tagged proteins readmitted and found no BH hit in any HR-versus-LR or interaction
contrast, so the null does not rest on this rule. Readmitting them took Acute_LR from 0 to 53
hits at BH < 0.10, 40 of them the readmitted blood proteins.

## Contaminant lists

Two sources, both matched by accession. cRAP collides with muscle proteins on gene symbol
(ALDOA against rabbit ALDOA_RABIT) and on description (a trypsin or albumin pattern takes
parvalbumin with it). Only the cRAP dust and contact section is used; the UPS1 spike-in section
holds myoglobin and creatine kinase M, ranks 1 and 2 of this proteome.

```{r contaminants}
contaminants <- tribble(
  ~uniprot_id, ~gene, ~class, ~reason,
  "P04264", "KRT1", "keratin", "cRAP dust/contact; cornified epidermal",
  "P35908", "KRT2", "keratin", "cRAP dust/contact; epidermal",
  "P13645", "KRT10", "keratin", "cRAP dust/contact; type-I partner of KRT1",
  "P35527", "KRT9", "keratin", "cRAP dust/contact; palmoplantar",
  "P08779", "KRT16", "keratin", "cRAP dust/contact",
  "P04259", "KRT6B", "keratin", "cRAP dust/contact",
  "P02533", "KRT14", "keratin", "cRAP dust/contact; basal epidermal",
  "P13647", "KRT5", "keratin", "cRAP dust/contact; basal epidermal",
  "P02538", "KRT6A", "keratin", "cRAP dust/contact",
  "Q04695", "KRT17", "keratin", "cRAP dust/contact",
  "Q7Z794", "KRT77", "keratin", "cRAP dust/contact",
  "Q8N1N4", "KRT78", "keratin", "cRAP dust/contact",
  "P05787", "KRT8", "keratin", "simple-epithelial; Spearman +0.78 with the KRT1/2/10 load",
  "P15924", "DSP", "keratin", "desmoplakin; skin desmosome",
  "Q02413", "DSG1", "keratin", "desmoglein-1; skin desmosome",
  "P81605", "DCD", "keratin", "dermcidin; sweat",
  "P31151", "S100A7", "keratin", "psoriasin; skin",
  "P01040", "CSTA", "keratin", "cystatin-A; cornified envelope",
  "P68871", "HBB", "globin", "red-cell carryover",
  "P69905", "HBA1", "globin", "red-cell carryover",
  "P02042", "HBD", "globin", "red-cell carryover",
  "P69891", "HBG1", "globin", "red-cell carryover",
  "P69892", "HBG2", "globin", "orphan gamma-globin; shares tryptic peptides with HBB/HBD",
  "P09105", "HBQ1", "globin", "red-cell carryover",
  "Q9NZD4", "AHSP", "globin", "erythroid-specific haemoglobin chaperone",
  "P01834", "IGKC", "immunoglobulin", "absent from HPA; constant region",
  "P01860", "IGHG3", "immunoglobulin", "absent from HPA; constant region",
  "P01880", "IGHD", "immunoglobulin", "absent from HPA; constant region",
  "P02746", "C1QB", "complement", "absent from HPA",
  "P00736", "C1R", "complement", "absent from HPA"
) |>
  bind_rows(
    read_csv(here("00_Input", "blood_contaminants.csv"), col_types = cols(.default = "c")) |>
      select(uniprot_id, gene, class, reason)
  )

# Share of each sample's summed intensity, measured before removal. These describe the
# biopsies; they filter nothing.
qc_panels <- list(
  keratin = c(
    "KRT1", "KRT2", "KRT10", "KRT8", "KRT9", "KRT16", "KRT14", "KRT5", "KRT6A",
    "KRT6B", "DCD"
  ),
  blood = c(
    "HBB", "HBA1", "HBD", "HBG1", "HBG2", "HBQ1", "AHSP", "ALB", "CA1", "SLC4A1",
    "SPTA1"
  ),
  leukocyte = c("MPO", "CTSG", "ELANE", "PRTN3", "AZU1", "S100A8", "S100A9", "LYZ", "LTF"),
  adipose = c("FABP4", "PLIN1", "PLIN4", "ADIPOQ", "ADIRF"),
  myofibre = c(
    "MB", "CKM", "ACTA1", "MYH1", "MYH2", "MYH7", "ALDOA", "CASQ1", "PYGM",
    "ATP2A1", "DES"
  )
)
c(contaminant_list = nrow(contaminants))
```

The 95 blood proteins come from `blood_contaminants.csv`, chosen by identity: immunoglobulins,
complement, secreted plasma proteins, mature red-cell and leukocyte proteins. They replaced a
cut at `blood_cor > 0.45`, whose null did not reproduce (last section).

## Read and deduplicate

```{r read}
raw <- read_excel(here("00_Input", "HRvLR_raw.xlsx"))
metadata <- as.data.frame(read_csv(here("00_Input", "HRvLR_meta.csv"), show_col_types = FALSE))
rownames(metadata) <- metadata$Col_ID

annotation_cols <- c("uniprot_id", "protein", "gene", "description", "n_seq")
# Every run in the report must be on the sample sheet and every sheet row in the report, or a
# sample would silently drop out or come back all-NA.
runs <- setdiff(names(raw), annotation_cols)
stopifnot("sample sheet and report disagree" = setequal(runs, metadata$Col_ID))
annotation <- raw[, annotation_cols]
intensity <- data.matrix(raw[, metadata$Col_ID])
n_raw <- nrow(annotation)

# One row per accession: the highest-mean row wins.
keep_row <- tibble(
  i = seq_len(n_raw), id = annotation$uniprot_id, m = rowMeans(intensity, na.rm = TRUE)
) |>
  slice_max(m, n = 1, by = id, with_ties = FALSE) |>
  pull(i)
annotation <- annotation[keep_row, ]
intensity <- intensity[keep_row, ]
n_dedup <- length(keep_row)
c(proteins = nrow(intensity), samples = ncol(intensity))
```

## Blood index and contamination

The blood index is each sample's mean log2 haemoglobin intensity, computed before removal while
the anchor proteins are still present. Each protein's Spearman correlation with it, `blood_cor`,
is reported beside every call and gates nothing.

```{r blood-index}
log_int <- log2(intensity)
log_int[!is.finite(log_int)] <- NA
blood_index <- colMeans(log_int[annotation$gene %in% cfg$blood_anchor, , drop = FALSE],
  na.rm = TRUE
)
testable <- rowSums(!is.na(log_int)) >= cfg$cor_min_obs
blood_cor <- suppressWarnings(as.vector(cor(
  t(log_int), blood_index,
  method = "spearman", use = "pairwise.complete.obs"
)))
blood_cor[!testable] <- NA

qc_index <- imap(qc_panels, \(genes, panel) {
  tibble(
    Col_ID = colnames(intensity), panel = panel,
    pct_signal = 100 * colSums(intensity[annotation$gene %in% genes, , drop = FALSE],
      na.rm = TRUE
    ) / colSums(intensity, na.rm = TRUE)
  )
}) |>
  list_rbind() |>
  left_join(select(metadata, Col_ID, Subject_ID, Group, Timepoint), by = "Col_ID")

qc_index |>
  summarise(median = round(median(pct_signal), 3), max = round(max(pct_signal), 3), .by = panel)
```

## One call per protein

A protein is removed when it is on the curated list, or when HPA marks it as a plasma,
immunoglobulin or erythrocyte protein and the muscle rescue does not reach it. Absence from HPA
is unknown, never a reason to remove. The red-cell reference (`RBC_proteome_reference.tsv`)
is reported as `in_rbc` and removes nothing: two thirds of any muscle proteome appears in it.

```{r calls}
hpa <- read_tsv(here("00_Input", "HPA_annotations_full.tsv"), show_col_types = FALSE) |>
  transmute(
    acc = Uniprot, protein_class = `Protein class`, secretome = `Secretome location`,
    blood_conc = suppressWarnings(as.numeric(`Blood concentration - Conc. blood MS [pg/L]`)),
    ery = suppressWarnings(as.numeric(`Single Cell Type RNA - Erythrocytes [nCPM]`)),
    myo = suppressWarnings(as.numeric(`Single Cell Type RNA - Myonuclei [nCPM]`))
  ) |>
  filter(!is.na(acc), acc != "") |>
  separate_longer_delim(acc, delim = ", ") |>
  mutate(acc = strip_iso(acc)) |>
  distinct(acc, .keep_all = TRUE)
rbc <- read_tsv(here("00_Input", "RBC_proteome_reference.tsv"), show_col_types = FALSE)

protein_calls <- annotation |>
  mutate(acc = strip_iso(uniprot_id), blood_cor = blood_cor) |>
  left_join(hpa, by = "acc") |>
  left_join(
    select(contaminants, acc = uniprot_id, contam_class = class, contam_reason = reason),
    by = "acc"
  ) |>
  mutate(
    is_curated = !is.na(contam_class),
    is_ery = !is.na(ery) & ery >= cfg$ery_cut,
    is_ig = str_detect(coalesce(protein_class, ""), "Immunoglobulin genes"),
    is_plasma = !is.na(secretome) & secretome == "Secreted to blood",
    in_rbc = acc %in% na.omit(rbc$acc) | gene %in% na.omit(rbc$gene),
    rescued = !is.na(myo) & myo >= cfg$myo_cut &
      (is.na(blood_conc) | blood_conc < cfg$blood_max),
    is_blood = is_ery | is_ig | is_plasma,
    contaminant = is_curated | (is_blood & !rescued),
    verdict = case_when(
      is_curated ~ paste0("remove: ", contam_class),
      is_blood & rescued ~ "keep: rescued (muscle-expressed)",
      is_plasma ~ "remove: plasma",
      is_ig ~ "remove: immunoglobulin",
      is_ery ~ "remove: erythrocyte",
      TRUE ~ "keep"
    ),
    reason = coalesce(contam_reason, verdict)
  ) |>
  select(
    uniprot_id, gene, description, verdict, reason, contaminant, in_rbc,
    secretome, blood_conc, ery, myo, blood_cor
  )
# Every verdict branch has to agree with the contaminant flag.
stopifnot(protein_calls$contaminant == str_starts(protein_calls$verdict, "remove"))
count(protein_calls, verdict, sort = TRUE)
```

Removal is a plain subset. `filter_proteins_by_annotation()` would read a logical column
directly, but its guard calls `if()` on a length-2 class vector and errors on every real DAList.

```{r dalist}
keep <- !protein_calls$contaminant
int_df <- as.data.frame(intensity[keep, ])
annot_df <- as.data.frame(annotation[keep, ])
rownames(int_df) <- rownames(annot_df) <- annot_df$uniprot_id
dal <- zero_to_missing(DAList(data = int_df, annotation = annot_df, metadata = metadata))
c(proteins = nrow(dal$data), samples = ncol(dal$data))
```

## Outlier samples

Four methods flag a sample: missingness (Tukey fence on the sample's share and on its spread
within subject), PCA Mahalanobis distance, median intensity (Hampel) and median inter-sample
correlation. A sample goes when three of the four agree.

```{r outliers}
lg <- log2(dal$data + 1)
complete <- dal$data[rowSums(is.na(dal$data)) == 0, ]
pcs <- prcomp(t(log2(complete + 1)), center = TRUE, scale. = TRUE)$x[, 1:3]
med_cor <- apply(cor(lg, use = "pairwise.complete.obs"), 2, \(x) median(x[x < 1], na.rm = TRUE))
hampel <- function(x) abs(x - median(x)) > cfg$mad_k * mad(x)
tukey <- function(x) x > quantile(x, 0.75) + 1.5 * IQR(x)

outlier_diag <- dal$metadata |>
  select(Col_ID, Subject_ID, Group, Timepoint) |>
  mutate(
    pct_missing = colMeans(is.na(dal$data))[Col_ID] * 100,
    delta_missing = ave(pct_missing, Subject_ID, FUN = \(x) pmax(x - min(x), max(x) - x)),
    miss_flag = tukey(pct_missing) | tukey(delta_missing),
    pca_flag = mahalanobis(pcs, colMeans(pcs), cov(pcs)) > qchisq(1 - cfg$mahal_p, df = 3),
    mad_flag = hampel(apply(lg, 2, median, na.rm = TRUE)[Col_ID]),
    cor_flag = med_cor[Col_ID] < median(med_cor) - cfg$mad_k * mad(med_cor),
    n_flags = miss_flag + pca_flag + mad_flag + cor_flag,
    consensus_outlier = n_flags >= cfg$outlier_k
  )
stopifnot(identical(rownames(pcs), outlier_diag$Col_ID))
outlier_ids <- outlier_diag$Col_ID[outlier_diag$consensus_outlier]
dal <- filter_samples(dal, !(Col_ID %in% outlier_ids))
filter(outlier_diag, n_flags > 0) |> select(Col_ID, Group, Timepoint, n_flags, consensus_outlier)
```

## Detection

Run last, on the samples that survive, so no sample about to be discarded counts toward a
protein's detections.

```{r detection}
n_before <- nrow(dal$data)
dal <- filter_proteins_by_group(dal,
  min_reps = cfg$miss_min_reps, min_groups = cfg$miss_min_groups,
  grouping_column = "Group_Time"
)
removed <- count(filter(protein_calls, contaminant), verdict, name = "n_removed")
filter_log <- tibble(
  step = c("raw input", "duplicate accession", removed$verdict, "outlier samples", "detection"),
  n_removed = c(NA, n_raw - n_dedup, removed$n_removed, 0L, n_before - nrow(dal$data))
) |>
  mutate(
    n_after = n_raw - cumsum(coalesce(n_removed, 0L)),
    pct_of_raw = round(100 * n_after / n_raw, 1),
    n_samples = c(rep(nrow(metadata), nrow(removed) + 2), rep(ncol(dal$data), 2))
  )
filter_log
```

## Figures

```{r fig-cascade}
#| fig-width: 9
#| fig-height: 10
set.seed(42) # the jittered points only
waterfall <- filter_log |>
  filter(!is.na(n_removed)) |>
  mutate(step = fct_inorder(str_remove(step, "^remove: ")), ymax = n_after + n_removed)
cascade <- ggplot(waterfall, aes(step)) +
  geom_rect(aes(
    xmin = as.numeric(step) - 0.35, xmax = as.numeric(step) + 0.35,
    ymin = n_after, ymax = ymax
  ), fill = "#D6604D") +
  geom_step(aes(y = n_after, group = 1), direction = "mid", colour = "grey35") +
  geom_text(aes(y = ymax, label = if_else(n_removed > 0, paste0("-", n_removed), "")),
    vjust = -0.6, size = 3, colour = "#B2182B"
  ) +
  labs(
    title = "Filtering cascade",
    subtitle = sprintf("%d proteins measured, %d analysed", n_raw, nrow(dal$data)),
    x = NULL, y = "proteins retained"
  ) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))

contamination <- qc_index |>
  filter(panel != "myofibre") |>
  ggplot(aes(Timepoint, pct_signal, fill = Timepoint)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.75) +
  geom_jitter(aes(shape = Group), width = 0.15, size = 1.4) +
  facet_wrap(~panel, scales = "free_y", nrow = 1) +
  scale_fill_brewer(palette = "Blues", guide = "none") +
  scale_shape_manual(values = c(HR = 16, LR = 1), name = NULL) +
  labs(
    title = "Contamination per sample, before removal",
    subtitle = "share of summed intensity per panel", x = NULL, y = "% of sample signal"
  )

flags <- outlier_diag |>
  filter(n_flags > 0) |>
  pivot_longer(ends_with("_flag"), names_to = "method", values_to = "flagged") |>
  mutate(Col_ID = fct_reorder(Col_ID, n_flags))
consensus <- ggplot(flags, aes(method, Col_ID, fill = flagged)) +
  geom_tile(colour = "white") +
  scale_fill_manual(values = c(`TRUE` = "#D6604D", `FALSE` = "grey92"), name = NULL) +
  labs(
    title = "Outlier flags, flagged samples only",
    subtitle = sprintf(
      "%d of %d samples flagged; %d met the %d-of-4 consensus",
      n_distinct(flags$Col_ID), nrow(metadata), length(outlier_ids), cfg$outlier_k
    ),
    x = NULL, y = NULL
  )

(cascade / contamination / consensus) +
  plot_layout(heights = c(1, 1.2, 0.9)) &
  theme_minimal(base_size = 10)
```

## Blood at T3 by arm

Tests whether blood content rises more at T3 in one arm, with a mixed model on the analysed
samples (arm by timepoint, subject intercept) on two scales: the log2 haemoglobin index and the
blood panel's share of signal. The permutation p shuffles the HR/LR label across subjects 999 times.

```{r blood-by-arm}
blood_data <- dal$metadata |>
  select(Col_ID, Subject_ID, Group, Timepoint) |>
  mutate(
    Group = factor(Group, c("HR", "LR")),
    log2_index = unname(blood_index[Col_ID]),
  ) |>
  left_join(
    filter(qc_index, panel == "blood") |> select(Col_ID, pct_signal),
    by = "Col_ID"
  )
arm_by_t3 <- function(data, response) {
  fit <- suppressMessages(lme4::lmer(
    reformulate(c("Group * Timepoint", "(1 | Subject_ID)"), response),
    data = data
  ))
  coef(summary(fit))["GroupLR:TimepointT3", c("Estimate", "t value")]
}
set.seed(42)
arms <- distinct(blood_data, Subject_ID, Group)
blood_by_arm <- map(set_names(c("log2_index", "pct_signal")), \(response) {
  observed <- arm_by_t3(blood_data, response)
  null_t <- replicate(999, {
    shuffled <- set_names(sample(arms$Group), arms$Subject_ID)
    arm_by_t3(mutate(blood_data, Group = shuffled[Subject_ID]), response)[["t value"]]
  })
  tibble(
    scale = response, lr_minus_hr_at_t3 = round(observed[["Estimate"]], 3),
    t = round(observed[["t value"]], 2),
    permutation_p = (1 + sum(abs(null_t) >= abs(observed[["t value"]]))) / 1000
  )
}) |>
  list_rbind()
blood_by_arm
blood_data |>
  summarise(mean_log2_index = round(mean(log2_index), 2), .by = c(Group, Timepoint)) |>
  arrange(Group, Timepoint)
```

The estimate is LR's T3 rise over HR's, relative to T1. A confound on this scale does not cancel
in the interaction contrasts.

## Why blood_cor removes nothing

The old cut was 0.45, justified as rounding above a 300-permutation null recorded at 0.43.
Recomputed here, and split by how many samples a protein was seen in:

```{r blood-null}
set.seed(42)
tested <- log_int[testable, ]
null_rho <- replicate(300, abs(as.vector(suppressWarnings(cor(
  t(tested), sample(blood_index),
  method = "spearman", use = "pairwise.complete.obs"
)))))
n_obs <- cut(rowSums(!is.na(tested)), c(0, 25, 40, Inf), c("15-25", "26-40", "41-48"))
null_999 <- \(rows) round(quantile(null_rho[rows, ], 0.999, na.rm = TRUE, names = FALSE), 3)
bind_rows(
  map(levels(n_obs), \(bin) {
    tibble(observations = bin, proteins = sum(n_obs == bin), null_999 = null_999(n_obs == bin))
  }),
  tibble(observations = "all", proteins = length(n_obs), null_999 = null_999(TRUE))
)
```

A protein seen in 15 samples and one seen in 48 do not draw |rho| from the same distribution, so
no single cut serves both. Removal moved to the curated list.

## Outputs

`DAList_filtered.rds` is read by `02_Normalization`. The workbook records every call.

```{r write}
saveRDS(dal, file.path(out, "DAList_filtered.rds"))
write_xlsx(
  list(
    filter_log = filter_log,
    contaminants_removed = arrange(filter(protein_calls, contaminant), verdict, gene),
    rescued = arrange(filter(protein_calls, str_detect(verdict, "rescued")), desc(myo)),
    protein_calls = protein_calls,
    contamination_index = qc_index,
    blood_index = tibble(Col_ID = names(blood_index), blood_index = unname(blood_index)),
    outlier_diagnostics = outlier_diag,
    blood_by_arm = blood_by_arm
  ),
  file.path(out, "01_filter.xlsx")
)
```

```{r session}
sessionInfo()
```
