---
title: "HRvLR F04 — Co-expression Modules"
subtitle: "Coherent biology, no responder signal — and why the earlier positive was an artifact"
author: "A_HRvLR_2026"
date: today
bibliography: ../../shared/references.bib
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
    embed-resources: true
    df-print: paged
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

```{r}
#| label: load
#| echo: false
pacman::p_load(here, openxlsx, dplyr, knitr)
wb <- here("04_Figures", "F04_modules", "c_data", "F04_modules_source_data.xlsx")
atlas <- read.xlsx(wb, "module_atlas")
fry <- read.xlsx(wb, "module_fry")
pred <- read.xlsx(wb, "module_prediction")
lmm <- read.xlsx(wb, "module_phenotype_lmm")
ora <- read.xlsx(wb, "module_ora")

responder <- c(
  "Baseline_HRvLR", "Trained_HRvLR", "Acute_HRvLR",
  "Training_Interaction", "Acute_Interaction"
)
fry_r <- fry[fry$contrast %in% responder, ]
n_grey <- 1920 - sum(atlas$n_proteins)
```

## What F04 asks

Proteins do not act alone. If high and low responders differ, the difference might not sit
in any single protein but in a **co-expressed programme** — a set of proteins that move
together. F04 builds those programmes without looking at responder status, then asks two
questions of them:

1. Does any module **move** with responder status?
2. Does any module's **baseline** state predict how much a subject adapts?

The answer to both is no. This page explains how the modules were built, why an earlier
version of this analysis appeared to say yes, and why that answer was wrong.

## The trap, and how F04 avoids it

WGCNA correlates proteins **across samples, treating them as independent**. Our 45 samples
are 16 subjects × 3 timepoints. They are not independent, and subject identity carries a
large share of the variance — the WGCNA FAQ names exactly this as the *"strong driver that
makes a subset of the samples globally different from the rest"* that invalidates the
scale-free assumption and inflates protein–protein correlation.

Build the network on raw abundance and the modules substantially encode **which subject a
sample came from**, not co-regulation. That is not hypothetical:

| perturbation | adjusted Rand index vs the raw-built network |
|---|---|
| change `pamRespectsDendro` | 0.876 |
| swap the entire WGCNA engine | 0.876 |
| drop imputation (complete cases only) | 0.506 |
| **remove subject identity** (centre within subject) | **0.288** |

**Every WGCNA parameter is noise. The design is what moves the modules.** Per-module subject
ICC reached 0.51 in the raw-built network.

**That artifact is the whole explanation for the earlier result.** A previous version of this
analysis reported a baseline-prediction q² of **0.713**. If a module is largely a subject
fingerprint, predicting across subjects in-sample looks excellent and leave-one-subject-out
destroys it — because LOSO removes precisely the thing the module was tracking. It did: the
same cell fell to **−0.165** under an honest refit.

**F04's fix: define the modules on within-subject-centred abundance, score them on raw
abundance.** Centring removes subject identity as a driver, so the modules capture how
proteins move *together within a subject* — the training and acute response, which is the
question. Eigengenes are then computed on **raw** abundance, so the between-subject
HR-versus-LR contrast is still there to be tested. It simply no longer *defines* the modules.

It works. In the centred network the subject ICC has a median of
`r sprintf("%.3f", median(atlas$icc))` and a maximum of `r sprintf("%.3f", max(atlas$icc))`,
down from 0.51.

## What the modules are

```{r}
#| label: atlas
#| echo: false
atlas |>
  arrange(desc(n_proteins)) |>
  transmute(Module = module, Proteins = n_proteins, `Subject ICC` = round(icc, 3)) |>
  kable()
```

`r nrow(atlas)` modules; `r n_grey` proteins left unassigned (grey).
`r n_distinct(ora$module[!is.na(ora$term)])` of the `r nrow(atlas)` carry a significant
GO:BP term. **The modules are real, coherent biology.** That is the one positive finding on
this page, and it is descriptive.

## Do the modules move? (`limma::fry`)

```{r}
#| label: fry-table
#| echo: false
fry |>
  group_by(contrast) |>
  summarise(
    `Min FDR` = round(min(fdr), 3),
    `Significant modules` = sum(fdr < 0.05), .groups = "drop"
  ) |>
  arrange(`Min FDR`) |>
  kable()
```

Across `r nrow(fry)` module × contrast tests, **`r sum(fry$fdr < 0.05)` is significant**: the
purple module in `Acute_LR` (FDR = `r sprintf("%.4f", min(fry$fdr))`). That is a
**within-group** response — the acute bout moves the proteome inside the LR arm. It is the
positive control, and it agrees with F03.

**Across the five contrasts that actually test responder status, nothing.
`r sum(fry_r$fdr < 0.05)` of `r nrow(fry_r)` cells are significant, and the minimum FDR is
`r sprintf("%.2f", min(fry_r$fdr))`.**

### Why `fry` and not `fgsea`

The panel this replaces ran **fgsea with the WGCNA modules as the gene sets**, ranked by
moderated *t* **from the same matrix that defined those modules**. The sets were built to be
co-expressed, so of course their members rank together: the test was circular. It returned
**padj = 7 × 10⁻³³** in a study whose minimum *q* anywhere else is 0.36. That number was the
tell.

`limma::fry` [@wu2010] is the correct instrument. It is a **rotation-based, self-contained**
set test: it asks whether the proteins in a predefined set move together **in a specified
contrast**, against the same linear model that produced the DEP results, carrying the subject
block and the `duplicateCorrelation` through. It was already used correctly elsewhere in this
repo.

## Does a baseline module predict adaptation?

```{r}
#| label: pred
#| echo: false
pred |>
  arrange(desc(q2)) |>
  head(5) |>
  transmute(
    Module = sub("^ME", "", module), Trait = trait,
    r = round(r, 2), `q2` = round(q2, 3), `BH q` = round(fdr, 2)
  ) |>
  kable(caption = "The five best cells of 72, ranked by q².")
```

No. The best cell in the entire `r nrow(pred)`-cell grid has
**q² = `r sprintf("%.3f", max(pred$q2, na.rm = TRUE))`** — and that is the **argmax of
`r nrow(pred)` cells**, so it is nearly free. Only `r sum(pred$q2 > 0, na.rm = TRUE)` of
`r nrow(pred)` cells have a positive q² at all, and nothing survives FDR (minimum BH *q* =
`r sprintf("%.2f", min(pred$fdr, na.rm = TRUE))`).

Read the two columns together. **`r sum(pred$p < 0.05, na.rm = TRUE)` cells have a nominally
significant in-sample correlation**, and almost none of them predicts a held-out subject.
That gap — in-sample correlation without out-of-sample prediction — is what overfitting looks
like, and it is why q² is on the face of the panel rather than *r*.

The mixed-model supplement keeps the repeated-measures structure this panel collapses to T1,
asking whether a module's whole time-course bends with the size of the response
(`ME ~ trait * timepoint + (1 | subject)`). Minimum BH *q* =
`r sprintf("%.2f", min(lmm$fdr_trait_time, na.rm = TRUE))`. Nothing survives there either.

## Methods note

Every DOI was resolved against Crossref.

| Panel | Claim | Statistic | Fit to this design |
|---|---|---|---|
| **A** atlas | The modules exist, and subject identity no longer drives them | Module sizes; ICC of each eigengene by subject | The ICC column is the diagnostic that exposed the earlier artifact, so it sits on the face of the figure, not in a supplement. Raw R² would mislead here: 15 subject df on 45 samples gives an expected R² of 0.34 under pure noise, so the **adjusted** R² and the ICC are the honest numbers. |
| **B** pathways | The modules are coherent biology | GO:BP over-representation, universe = **the measured proteome** | The universe matters. A protein that was never quantified could not have been assigned to a module, so scoring against the whole genome would inflate every term. |
| **C** fry | No module moves with responder status | `limma::fry` [@wu2010], modules as the index, on the 03_DEP design with the subject block and `duplicateCorrelation` | Correct, and not circular: the contrast comes from the experimental design, not from the matrix that built the sets. A **self-contained** rotation test, not a competitive one. |
| **D** phenotype | No baseline module predicts adaptation | In-sample *r* and **leave-one-subject-out q²**, BH across the grid | LOSO is the right resampling unit: the dependency is *within subject*, so holding out a sample rather than a subject would leak. q² ≤ 0 means the model predicts a held-out subject worse than the trait's own mean does. |
| **D** supp | Nor does the module's trajectory bend with it | `ME ~ trait * timepoint + (1 \| subject)`, BH within trait | Keeps the repeated-measures structure. **`r sum(lmm$singular, na.rm = TRUE)` of `r nrow(lmm)` fits are singular** and are flagged, not dropped: a time-invariant trait can drive the random-intercept variance to zero, inflating the trait main effect — so the interaction is the term to trust. |

### The WGCNA parameters, and their sources

Signed network; `bicor` with `maxPOutliers = 0.05`; `deepSplit = 2`; `minModuleSize = 30`;
`mergeCutHeight = 0.15`; `pamStage = TRUE`; `pamRespectsDendro = TRUE`. WGCNA:
Langfelder & Horvath [@langfelder2008].

- **Signed and `bicor`** are the *only two* deviations from default the WGCNA FAQ explicitly
  recommends. `maxPOutliers = 0.05` is its "strongly recommend" — the package default of 1
  disables the outlier guard entirely, which is a silent and common error.
- **The soft power is what `pickSoftThreshold` itself returns** at its documented default
  (`RsquaredCut = 0.85`). The FAQ's sample-size table (signed, n > 40 → 12) is a **fallback
  for when the scale-free fit never clears the threshold.** Ours clears it, so the table does
  not apply. The supplement plots the sweep and reports the achieved R² and mean connectivity
  — neither of which the proteomics literature routinely reports.
- **`pamRespectsDendro = TRUE` is the package default.** No published source recommends
  setting it `FALSE`; that convention traces to an unexplained line in the official tutorial
  *script*. Every WGCNA-on-proteomics paper surveyed uses `TRUE`.
- `minModuleSize = 30` is the official tutorial's value. Module count is insensitive to
  `mergeCutHeight` across 0.07–0.25 on this data, so that parameter is not load-bearing.

### Where this sits outside the validated regime

Stated plainly, because it is the real limitation and nothing else covers it:

- **There is no published guidance for WGCNA on repeated-measures designs.** None. The
  within-subject centring here is a defensible response to a documented failure mode, not a
  validated method.
- **1,920 proteins is far below the ~20,000 features WGCNA was built around.** Langfelder
  concedes small feature counts are unlikely to be scale-free, so a mediocre fit is expected
  rather than a bug.
- **12.2 % of the matrix was imputed** by missForest, which predicts proteins *from other
  proteins* — manufacturing co-expression for a method that is entirely about correlation.
  Dropping imputation changes the module assignment substantially (ARI 0.506). Centring does
  not fix this. It is a separate exposure, and it is why the modules are presented as
  descriptive rather than as a discovery.

## Recap

- The modules are real and biologically coherent. That is the positive finding, and it is
  descriptive.
- **No module moves with responder status**: 0 of `r nrow(fry_r)` cells, minimum FDR
  `r sprintf("%.2f", min(fry_r$fdr))`.
- **No baseline module predicts adaptation**: best q² =
  `r sprintf("%.3f", max(pred$q2, na.rm = TRUE))`, the argmax of `r nrow(pred)` cells.
- The earlier q² of 0.713 was an artifact of modules that encoded subject identity. Define
  them properly and the signal was never there.
- The one significant module × contrast cell is a **within-group acute response** — the
  positive control, not a responder difference.

## References

::: {#refs}
:::

```{r sessioninfo}
sessionInfo()
```
