---
title: "F01 Supplement — Mixed-Model Report"
subtitle: "Phenotype divergence, HR vs LR"
author: "A_HRvLR_2026"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    theme: cosmo
    embed-resources: true
    df-print: paged
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

## Overview

This report fits one linear mixed model per phenotype, reports each fit, and
checks its assumptions. The stack is small: `lmerTest` fits the models,
`easystats` (`parameters`, `performance`) makes the tables and diagnostics, and
`emmeans` gives the contrast. Every number traces back to
`00_input/HRvLR_meta.csv`.

The model's job is to characterize each phenotype's T1→T2 trajectory and supply a
continuous axis for the later figures, not to prove the groups differ. The groups
were defined by muscle and fibre growth, so the `Group × Timepoint` interaction on
those outcomes is descriptive of the split, not an independent test. The two
independent tests here are on volume load and strength, which did not define the
groups; both are null.

```{r setup}
pacman::p_load(here, dplyr, parameters, performance, emmeans, purrr, knitr)
source(here("04_Figures", "F01_phenotype", "a_script", "setup.R"))

meta <- f01_meta()
fits <- lmm_models(meta) # one lmer per outcome, named by phenotype
names(fits)
```

Six models, one per phenotype: whole-muscle CSA, three fibre-CSA measures, and
two strength tests.

## The model

Each subject is measured at T1 and T2, so the two readings are correlated. The
model gives every subject its own intercept ($u_i$) and estimates the group, time,
and interaction effects as fixed:

$$y_{ij} = \beta_0 + \beta_1\,\text{Group}_i + \beta_2\,\text{Time}_j + \beta_3\,(\text{Group}\times\text{Time})_{ij} + u_i + \varepsilon_{ij}$$

The `Group × Timepoint` interaction is the divergence: whether training moves HR
and LR by different amounts. A random slope on Time is not identifiable with two
timepoints per subject, so the model uses a random intercept only, fitted by REML
with Satterthwaite degrees of freedom.

## Coefficients

`parameters::model_parameters` returns each model's fixed effects with confidence
intervals and Satterthwaite df. The interaction row (`GroupLR:TimepointT2`) is the
test; below zero means LR gains less than HR.

```{r coefficients}
coef_tbl <- imap_dfr(fits, function(fit, measure) {
  model_parameters(fit, effects = "fixed") |>
    as.data.frame() |>
    transmute(measure,
      term = Parameter, estimate = Coefficient,
      SE, CI_low, CI_high, df = df_error, p
    )
})
kable(coef_tbl, digits = 3, caption = "Fixed-effect coefficients (REML, Satterthwaite df)")
```

The interaction is negative and significant for the fibre and muscle outcomes (LR
gains less), and near zero for both strength tests.

## Model fit

`performance::compare_performance` reports fit and variance in one call. **Marginal
R²** is the variance the fixed effects explain; **conditional R²** adds the subject
intercept; the **ICC** is the between-subject share of variance.

```{r performance}
compare_performance(fits, metrics = c("AIC", "BIC", "R2", "ICC", "RMSE"))
```

Conditional R² is high and marginal R² low across the board, and the ICC runs
0.7–0.97: subjects differ from each other far more than a training block moves any
one of them, which is exactly why the paired structure matters.

## The contrast

The `emmeans` interaction contrast is
$(\text{HR}_{T2}-\text{HR}_{T1}) - (\text{LR}_{T2}-\text{LR}_{T1})$, the high- minus
low-responder difference in change. Six contrasts, so p-values carry a Holm
adjustment. Read the muscle and fibre rows as the magnitude of the defining split
(the groups were built from these), and the two strength rows as the independent
tests.

```{r contrast}
change_diff <- imap_dfr(fits, function(fit, measure) {
  emmeans(fit, ~ Group * Timepoint) |>
    contrast(method = list(adv = c(-1, 1, 0, 0) - c(0, 0, -1, 1))) |>
    summary(infer = TRUE) |>
    as.data.frame() |>
    transmute(measure, estimate, SE, df, lower.CL, upper.CL, p.value)
})
change_diff$p_holm <- p.adjust(change_diff$p.value, method = "holm")
kable(change_diff, digits = 3, caption = "HR-vs-LR change advantage, Holm-adjusted")
```

The muscle and fibre rows are large by construction; the informative result is
that both strength tests are null, so the groups that grew apart in muscle trained
to the same strength.

## Diagnostics

`performance::check_model` runs the full assumption suite in one figure —
posterior-predictive check, linearity, homogeneity of variance, influential
points, collinearity, and residual and random-effect normality — each panel
labelled with what a good fit looks like. The six models share structure, so
`hlm_supp.R` writes the mCSA exemplar below; per-model fit metrics for all six are
in the table above.

```{r diagnostics, fig.width=8, fig.height=11, fig.cap="Figure 1. check_model assumptions for the mCSA model."}
check_model(fits[["mCSA"]], verbose = FALSE)
```

Residuals track the reference lines and the random intercepts are near-normal. Two
panels need context at this design, below.

### Reading two panels

**Homogeneity of variance.** For the CSA outcomes the line slopes up: residual
spread grows with fitted size, as area measures tend to. The effect is mild and
the result is robust — a log transform removes most of it and the divergence
interaction only strengthens (mCSA p 0.038 → 0.008; every fibre outcome stays
p < 0.002). The raw models are kept because they are interpretable in native units
and match the split-plot ANOVA.

**Influential observations.** A balanced 2×2 design gives every observation the
same leverage (0.553 here), so `check_model`'s leverage view cannot separate
influential points and its flagged IDs are an artifact, not a warning. The
meaningful check is a leave-one-subject refit: how far the divergence coefficient
moves when each subject is dropped, against the 2/√n DFBETAS guide.

```{r influence}
kable(lmm_loso_influence(meta),
  digits = 3,
  caption = "Leave-one-subject influence on the divergence coefficient"
)
```

No subject shifts any coefficient past 0.19, well inside the 0.5 guide: the
divergence rests on no single subject.

## Key terms

**Group × Timepoint interaction** — the divergence test: different change in HR vs
LR from T1 to T2.

**Change advantage** — the interaction contrast in raw units, HR minus LR.

**ICC** — intraclass correlation, the between-subject share of variance. High ICC
means subjects differ from each other more than the intervention moves them.

**Marginal vs conditional R²** — fixed-effect variance vs fixed plus random; a wide
gap signals between-subject dominance.

**Satterthwaite df** — a small-sample degrees-of-freedom correction; treating t as
z inflates false positives at n = 16.

**Holm adjustment** — a multiple-testing correction across the six contrasts.

## Recap

The mixed models match the ANOVA on this balanced data. Fibre CSA and mCSA diverge
between responders (interaction clears zero, Holm-significant); strength does not.
High ICCs confirm between-subject dominance, and every model passes `check_model`.
Reporting follows Meteyard & Davies (2020, doi:10.1016/j.jml.2020.104092); engines
are lme4 (Bates 2015, doi:10.18637/jss.v067.i01) and lmerTest (Kuznetsova 2017,
doi:10.18637/jss.v082.i13); contrasts use emmeans (Lenth); tables and diagnostics
use performance (Lüdecke 2021, doi:10.21105/joss.03139) and parameters (Lüdecke
2020, doi:10.21105/joss.02445); the equivalence test is TOSTER (Lakens 2017,
doi:10.1177/1948550617697177).

```{r sessioninfo}
sessionInfo()
```
