---
title: "Stage 1: Normalization"
subtitle: "HPA Filter → Missingness Filter → Outlier Detection → Cycloess"
author: "DTL"
date: today
bibliography: references.bib
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    theme:
      light: flatly
      dark: darkly
    fig-width: 10
    fig-height: 7
    df-print: paged
    smooth-scroll: true
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
#| eval: true

knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)
```

## Overview

This document traces the complete normalization pipeline for the HRvLR DIA-MS skeletal muscle proteomics dataset. The study follows a 2 × 3 factorial design: Responder (HR / LR) × Timepoint (T1 / T2 / T3), with n = 8 subjects per group.

Stage 1 transforms raw DIA-NN intensities into a quality-controlled, normalized protein matrix suitable for differential abundance analysis. The pipeline is split across two scripts:

| Script | Role |
|--------|------|
| `01_run_normalization.R` | Core computation: filtering, outlier detection, cycloess normalization, PRONE benchmark, export |
| `02_norm_reports.R` | Visualization: reads `00_report_intermediates.rds` and produces the 4-page diagnostic PDF + supplementary workbook |

This two-script design separates compute-heavy operations from reporting. The intermediates `.rds` file serializes all plot-ready objects so that `02_norm_reports.R` can produce publication-quality diagnostics without re-running the pipeline. Every filtering decision, sample quality judgment, and normalization choice is recorded here alongside the code and quantitative evidence that supports it.

## Script 1: `01_run_normalization.R`

### Data loading and validation

Raw protein intensities and annotation are stored together in a single Excel workbook (`00_input/HRvLR_raw.xlsx`). The first step separates the annotation columns — UniProt ID, protein name, gene symbol, description, and number of peptide sequences — from the sample intensity columns, which are named by `Col_ID` to match the sample metadata CSV.

```{r}
#| label: norm-load
#| eval: false

raw        <- read_excel(cfg$raw_file)
annot_cols <- c("uniprot_id", "protein", "gene", "description", "n_seq")
annotation <- raw[, annot_cols]
intensity  <- raw[, setdiff(names(raw), annot_cols)]

metadata <- as.data.frame(read_csv(cfg$meta_file, show_col_types = FALSE))
rownames(metadata) <- metadata$Col_ID
stopifnot("Sample mismatch" = setequal(colnames(intensity), metadata$Col_ID))
intensity <- intensity[, metadata$Col_ID]  # enforce consistent column order
```

The `stopifnot()` call is a hard contract check. If any sample name in the intensity matrix does not exactly match a `Col_ID` in the metadata — whether due to a typo, a missing sample, or a file version mismatch — the pipeline stops immediately. R's silent column recycling and partial matching mean a sample order mismatch would not throw an error on its own, yet it would corrupt every downstream association between a sample's intensities and its group label.

### HPA tissue filter

DIA-NN's library-free search mode matches peptide spectra against a complete human proteome reference, which inevitably yields non-specific identifications from blood (hemoglobin chains, serum albumin), skin (keratins), and connective tissue contaminants. A blanket abundance filter would remove low-abundance muscle proteins before removing high-abundance contaminants; the tissue-relevance filter is the biologically correct first step.

The Human Protein Atlas (HPA) skeletal muscle annotation file provides a gene-symbol reference set. Only proteins whose gene symbol is present in this set are retained.

```{r}
#| label: hpa-filter
#| eval: false

hpa <- read_tsv(cfg$hpa_file, show_col_types = FALSE) |>
  select(Gene, Ensembl, Evidence,
         Protein_class    = `Protein class`,
         Subcellular_main = `Subcellular main location`,
         Interactions) |>
  distinct(Gene, .keep_all = TRUE)

n_before   <- nrow(annotation)
keep_hpa   <- annotation$gene %in% hpa$Gene
intensity  <- intensity[keep_hpa, ]
annotation <- annotation[keep_hpa, ] |> left_join(hpa, by = c("gene" = "Gene"))
```

The `left_join()` appends HPA annotation fields — subcellular location, protein class, interaction partners — directly to the annotation data frame, making these attributes available for downstream pathway and subcellular analyses without a separate merge step.

### Deduplication

DIA-NN can report multiple precursor groups for the same UniProt accession when a protein has several proteoform entries in the reference library. Retaining multiple rows per UniProt ID would artificially inflate protein counts and introduce redundant observations. The row with the highest mean intensity across all samples is kept, as it represents the most reliably detected peptide evidence.

```{r}
#| label: deduplication
#| eval: false

if (any(duplicated(annotation$uniprot_id))) {
  annotation$row_mean <- rowMeans(data.matrix(intensity), na.rm = TRUE)
  keep_idx <- annotation |>
    mutate(row_idx = row_number()) |>
    group_by(uniprot_id) |>
    slice_max(row_mean, n = 1, with_ties = FALSE) |>
    pull(row_idx)
  annotation <- annotation[keep_idx, ]
  intensity  <- intensity[keep_idx, ]
  annotation$row_mean <- NULL
}
```

### DAList assembly and missingness filter

The `proteoDA` package [@thurman_proteoDA_2023] organizes data, annotation, and metadata into a `DAList` object — a named list whose three slots are kept structurally synchronized through all filtering, normalization, and modeling operations. This eliminates the most common source of silent errors in proteomics pipelines: row- or column-order divergence between the intensity matrix and the annotation or metadata tables.

```{r}
#| label: assemble-dalist
#| eval: false

int_mat <- as.data.frame(data.matrix(intensity))
rownames(int_mat) <- annotation$uniprot_id
annot_df <- as.data.frame(annotation); rownames(annot_df) <- annotation$uniprot_id
meta_df  <- as.data.frame(metadata);   rownames(meta_df)  <- metadata$Col_ID

dal <- DAList(data = int_mat, annotation = annot_df, metadata = meta_df)
dal <- zero_to_missing(dal)
```

DIA-NN encodes non-detections as numeric zeros rather than `NA`. The `zero_to_missing()` function converts all zeros to `NA`, after which limma fits each protein using only its observed samples — the statistically correct treatment for instrument-censored data.

After zero conversion, proteins are filtered by group-wise detection: a protein must have at least 5 non-missing values (of 8) in at least 1 of the 6 Group_Time cells. The group-wise logic preserves condition-specific proteins that would be removed by a naive global filter.

```{r}
#| label: missingness-filter
#| eval: false

n_before <- nrow(dal$data)
dal <- filter_proteins_by_group(dal, min_reps = cfg$miss_min_reps,
                                min_groups = cfg$miss_min_groups,
                                grouping_column = "Group_Time")
```

```{r}
#| label: show-pipeline-summary
#| eval: true

suppressPackageStartupMessages({ library(dplyr); library(readr); library(readxl) })

if (file.exists("01_normalization/c_data/05_normalization_supp.xlsx")) {
  readxl::read_excel("01_normalization/c_data/05_normalization_supp.xlsx",
                     sheet = "Pipeline_Summary")
}
```

### Outlier detection (3-method consensus)

Sample outlier detection balances false positives — removing a biologically extreme but technically sound sample — against false negatives — retaining a failed sample that will inflate within-group variance. A multi-method consensus approach [@huang2024] addresses this by requiring independent corroboration: a sample is flagged as an outlier only if at least `k = 2` of 3 detection methods agree.

#### Method 1: Sample missingness

Per-sample missingness percentages are computed from the filtered data matrix. An absolute threshold is set using the Tukey fence (Q3 + 1.5 × IQR) of the distribution of per-sample missingness, and a within-subject delta threshold is computed from the distribution of pairwise missingness changes. A sample is flagged if it exceeds either threshold.

```{r}
#| label: outlier-missingness
#| eval: false

pct_missing <- colMeans(is.na(dal$data)) * 100
miss_thresh <- quantile(pct_missing, 0.75) + 1.5 * IQR(pct_missing)

miss_info <- dal$metadata |>
  select(Col_ID, Subject_ID, Group, Timepoint) |>
  mutate(pct_missing = pct_missing[Col_ID],
         prefix = str_remove(Col_ID, "_T[123]$")) |>
  group_by(prefix) |>
  mutate(delta_missing = sapply(pct_missing, \(x) max(abs(x - pct_missing)))) |>
  ungroup()

delta_thresh <- quantile(miss_info$delta_missing, 0.75) +
  1.5 * IQR(miss_info$delta_missing)
miss_info$miss_flag <- miss_info$pct_missing > miss_thresh |
  miss_info$delta_missing > delta_thresh
```

#### Method 2: PCA Mahalanobis distance

PCA is run on complete-case proteins after log2 transformation. Each sample's position in PC1–PC3 space is summarized by its Mahalanobis distance from the centroid, and samples exceeding the chi-squared critical value at 99% with 3 degrees of freedom are flagged.

```{r}
#| label: outlier-pca
#| eval: false

complete_mat <- dal$data[rowSums(is.na(dal$data)) == 0, ]
pca_pre  <- run_pca(complete_mat, dal$metadata, log_transform = TRUE)
pc3      <- pca_pre$pca$x[, 1:3]
mahal    <- mahalanobis(pc3, colMeans(pc3), cov(pc3))
pca_flags <- tibble(Col_ID = colnames(dal$data), mahal_dist = mahal,
                    pca_flag = mahal > qchisq(1 - cfg$mahal_p, df = 3))
```

#### Method 3: MAD on sample-median intensity

The median log2 intensity is computed per sample, and samples deviating by more than 3 MADs from the cohort median are flagged. MAD is preferred over standard deviation because it is not inflated by the very outliers being detected.

```{r}
#| label: outlier-mad
#| eval: false

samp_med   <- apply(log2(dal$data + 1), 2, median, na.rm = TRUE)
global_med <- median(samp_med)
mad_val    <- mad(samp_med)
mad_flags  <- tibble(Col_ID = names(samp_med), sample_median = samp_med,
                     mad_flag = abs(samp_med - global_med) > cfg$mad_k * mad_val)
```

#### Consensus and removal

The three flag vectors are combined per sample. A consensus outlier is any sample where at least `k = 2` flags are `TRUE`. Only the flagged sample is removed — not the subject's other timepoints — since `duplicateCorrelation` in limma explicitly supports incomplete block designs.

```{r}
#| label: outlier-consensus
#| eval: false

outlier_diag <- miss_info |>
  left_join(pca_flags, by = "Col_ID") |>
  left_join(mad_flags, by = "Col_ID") |>
  mutate(n_flags = miss_flag + pca_flag + mad_flag,
         consensus_outlier = n_flags >= cfg$outlier_k)

outlier_ids <- outlier_diag |> filter(consensus_outlier) |> pull(Col_ID)
if (length(outlier_ids) > 0) {
  dal <- filter_samples(dal, !(Col_ID %in% outlier_ids))
}
```

```{r}
#| label: show-outlier-diagnostics
#| eval: true

if (file.exists("01_normalization/c_data/05_normalization_supp.xlsx")) {
  readxl::read_excel("01_normalization/c_data/05_normalization_supp.xlsx",
                     sheet = "Outlier_Diagnostics") |>
    dplyr::select(any_of(c("Col_ID", "pct_missing", "delta_missing", "mahal_dist",
                            "miss_flag", "pca_flag", "mad_flag", "n_flags",
                            "consensus_outlier"))) |>
    dplyr::arrange(dplyr::desc(n_flags)) |>
    head(10)
}
```

### Normalization (cycloess)

Cyclic loess normalization [@bolstad_cyclic_loess_2003] corrects for non-linear, intensity-dependent systematic biases between samples. For each sample pair, an MA-plot loess smoother is fit and subtracted, iterating over all pairs until corrections stabilize. This approach is preferable to quantile normalization for DIA-MS data where genuine biological differences can manifest as global intensity shifts — cycloess corrects non-linear technical biases while preserving true biological differences in overall abundance level.

```{r}
#| label: cycloess-normalization
#| eval: false

write_norm_report(dal, grouping_column = "Group_Time",
                  output_dir = cfg$report_dir,
                  filename = "01_norm_comparison.pdf", overwrite = TRUE)

write_qc_report(dal, color_column = "Group_Time",
                output_dir = cfg$report_dir,
                filename = "02_qc_pre.pdf", overwrite = TRUE)

dal_pre <- dal
saveRDS(dal_pre, file.path(cfg$data_dir, "01_DAList_prenorm.rds"))

dal <- normalize_data(dal, norm_method = cfg$norm_method)

write_qc_report(dal, color_column = "Group_Time",
                output_dir = cfg$report_dir,
                filename = "03_qc_post.pdf", overwrite = TRUE)
```

```{r}
#| label: normalized-matrix-preview
#| eval: true

if (file.exists("01_normalization/c_data/02_normalized.csv")) {
  df <- readr::read_csv("01_normalization/c_data/02_normalized.csv",
                         show_col_types = FALSE)
  intensity_block <- df[, -(1:4)]
  cat(sprintf("Normalized matrix: %d proteins x %d sample columns\n",
              nrow(df), ncol(intensity_block)))
  cat(sprintf("Missing values: %d (%.1f%%)\n",
              sum(is.na(intensity_block)),
              100 * sum(is.na(intensity_block)) / (nrow(df) * ncol(intensity_block))))
  df |> dplyr::select(1:8) |> head(4)
}
```

### Normalization benchmark (PRONE framework)

To validate the cycloess choice empirically, eight normalization methods are benchmarked against three quantitative metrics from the PRONE framework [@arend_PRONE_2025]. PCV (pooled within-group coefficient of variation, computed on linear-scale intensities per @brenes2024) and PMAD (pooled within-group median absolute deviation) measure technical noise reduction, while COR (mean pairwise within-group Pearson correlation) measures biological coherence. A composite rank — the mean of the three per-metric ranks — identifies the best performer.

```{r}
#| label: normalization-benchmark
#| eval: false

methods <- c("log2", "median", "mean", "vsn", "quantile", "cycloess", "rlr", "gi")

norm_scores <- lapply(methods, function(m) {
  dal_n <- tryCatch(normalize_data(dal_pre, norm_method = m), error = function(e) NULL)
  if (is.null(dal_n)) return(NULL)
  mat <- as.matrix(dal_n$data); grps <- dal_n$metadata$group
  tibble(method = m,
         PCV  = round(norm_metric(mat, grps, "cv"),  4),
         PMAD = round(norm_metric(mat, grps, "mad"), 4),
         COR  = round(norm_metric(mat, grps, "cor"), 4))
}) |> bind_rows() |>
  mutate(PCV_rank = rank(PCV), PMAD_rank = rank(PMAD), COR_rank = rank(-COR),
         composite = round((PCV_rank + PMAD_rank + COR_rank) / 3, 2)) |>
  arrange(composite)

write_csv(norm_scores, file.path(cfg$data_dir, "04_norm_quality_scores.csv"))
```

```{r}
#| label: show-norm-scores
#| eval: true

if (file.exists("01_normalization/c_data/04_norm_quality_scores.csv")) {
  readr::read_csv("01_normalization/c_data/04_norm_quality_scores.csv",
                   show_col_types = FALSE) |>
    dplyr::select(method, PCV, PMAD, COR, composite) |>
    dplyr::arrange(composite)
}
```

### Export and intermediates

The script exports the normalized matrix as CSV, the full DAList as RDS, and computes all plot-ready data objects (filter bar chart data, missingness bar chart, per-subject variability, eta-squared variance partition, pre/post PCA scores) which are serialized into `00_report_intermediates.rds` for consumption by `02_norm_reports.R`.

```{r}
#| label: export
#| eval: false

export_df <- bind_cols(
  as_tibble(dal$annotation) |> select(uniprot_id, protein, gene, description),
  as_tibble(dal$data))
write_csv(export_df, file.path(cfg$data_dir, "02_normalized.csv"))
saveRDS(dal, file.path(cfg$data_dir, "03_DAList_normalized.rds"))

# Compute plot data and save intermediates for 02_norm_reports.R
saveRDS(intermediates, file.path(cfg$data_dir, "00_report_intermediates.rds"))
```

## Script 2: `02_norm_reports.R`

This script reads `00_report_intermediates.rds` and produces two outputs without re-running any computation:

### 4-page diagnostic PDF (`04_diagnostics.pdf`)

| Page | Content |
|------|---------|
| 1 | Protein filtering stacked bar chart + per-sample detection bar chart (outliers faded) |
| 2 | Outlier diagnostics: 3-panel figure (missingness scatter, PCA Mahalanobis, MAD median intensity) with shared legend |
| 3 | Post-normalization PCA with 68% confidence ellipses per Group_Time |
| 4 | Per-subject IQR variability plot + eta-squared histogram for variance partition |

### Supplementary Excel workbook (`05_normalization_supp.xlsx`)

| Sheet | Content |
|-------|---------|
| Pipeline_Summary | Cumulative protein counts at each filter step |
| Norm_Ranking | 8 methods ranked by PCV, PMAD, COR, composite |
| Outlier_Diagnostics | Per-sample flags for all 3 methods + consensus |
| Filtered_Proteins | Proteins removed by HPA filter or missingness filter |

```{r}
#| label: reports-overview
#| eval: false

int <- readRDS(file.path(cfg$data_dir, "00_report_intermediates.rds"))

# All plot objects constructed from intermediates (no recomputation)
# -> 04_diagnostics.pdf (4 pages)
# -> 05_normalization_supp.xlsx (4 sheets)
```

## Stage outputs

| File | Description |
|------|-------------|
| `b_reports/01_norm_comparison.pdf` | `write_norm_report()` — 8 normalization methods side-by-side (density, boxplot) |
| `b_reports/02_qc_pre.pdf` | QC before normalization: boxplots, correlation heatmap, PCA |
| `b_reports/03_qc_post.pdf` | QC after cycloess: same layout — compare to pre for normalization effect |
| `b_reports/04_diagnostics.pdf` | 4 pages: protein filtering + detection; outlier diagnostics (3 methods); post-norm PCA; variability summary |
| `c_data/00_report_intermediates.rds` | Serialized plot-ready objects for `02_norm_reports.R` |
| `c_data/01_DAList_prenorm.rds` | DAList pre-normalization (reference for PRONE benchmark) |
| `c_data/02_normalized.csv` | **PRIMARY** — cycloess-normalized matrix (4 annotation cols + sample cols) |
| `c_data/03_DAList_normalized.rds` | **PRIMARY** — normalized DAList with full metadata |
| `c_data/04_norm_quality_scores.csv` | PRONE rankings for 8 normalization methods |
| `c_data/05_normalization_supp.xlsx` | 4-sheet workbook: Pipeline_Summary, Norm_Ranking, Outlier_Diagnostics, Filtered_Proteins |
