---
title: "F04 association — proteins, trajectory view"
subtitle: "Does a protein's whole time-course track the training response"
author: "A_HRvLR_2026"
format:
  html:
    toc: true
    embed-resources: true
    df-print: paged
execute:
  eval: true
  echo: false
  warning: false
  message: false
---

## In plain terms

Use all three biopsies per person and ask whether a protein's change over
time tracks how much they responded, letting each person have their own
starting point.

## Method

For each protein and each continuous training response, a mixed model is
fit across all three timepoints: `score ~ response * timepoint + (1 |
subject)`, with the response-by-timepoint interaction F-test as the
statistic of interest. The random intercept lets each subject start from
their own baseline; p-values are corrected with Benjamini-Hochberg across
proteins within each trait. Descriptive, exploratory, n = 16.

Two things about this model need to be said plainly.

**Overfitting at small n.** With 16 people and some missing biopsies, a
per-protein mixed model overfits for a meaningful fraction of proteins: the
residual variance collapses toward zero and manufactures an impossibly
large, meaningless F-statistic. The analysis guards against this — any
degenerate fit (a boundary random effect, or a residual variance below 5%
of the protein's own spread) is dropped to `NA` rather than reported. The
counts below show how many proteins were guarded out.

**Anti-conservative even among the kept fits.** The model uses a random
intercept only. The stricter choice — a per-person random slope on
timepoint — is unidentifiable at this sample size, so even the fits that
pass the guard run mildly optimistic: their p-values are somewhat smaller
than they should be.

## Results

```{r}
#| label: load-results
pacman::p_load(here, openxlsx, dplyr)

assoc <- read.xlsx(
  here(
    "04_Figures", "F04_association", "HLM", "proteins", "c_data",
    "proteins_source_data.xlsx"
  ),
  "protein_association_hlm"
)

n_features <- n_distinct(assoc$feature)
n_traits <- n_distinct(assoc$trait)
n_cells <- nrow(assoc)
n_guarded <- sum(is.na(assoc$p))
n_kept <- sum(!is.na(assoc$p))
n_hits <- sum(assoc$bh < 0.05, na.rm = TRUE)
min_bh <- min(assoc$bh, na.rm = TRUE)
```

`r n_features` proteins were tested against `r n_traits` training-response
traits, for `r n_cells` protein-by-trait cells in total. `r n_guarded` of
those cells were guarded out as degenerate fits (`NA`, not reported);
`r n_kept` were kept. Of the kept cells, `r n_hits` pass BH < 0.05, and the
smallest BH-adjusted q anywhere in the table is `r sprintf("%.3f", min_bh)`.

After guarding, nothing survives FDR. That is a clean corroboration, not a
failure: the trajectory view agrees with the snapshot (limma) view and with
the study-wide null — two different methods asking the same question and
returning the same honest answer.

```{r sessioninfo}
sessionInfo()
```
