---
title: "03 · Imputation"
subtitle: "missForest, for the methods that need a complete matrix"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    df-print: kable
    fig-width: 7
    fig-height: 5
    fig-dpi: 150
    fig-align: center
execute:
  message: false
---

The differential fit reads the unimputed matrix. Three later methods cannot: `fry` rotates
residuals and takes no missing value, singscore ranks every protein in a sample, and WGCNA
correlates complete columns. They read this file. Nothing else does.

```{r setup}
library(here)
library(missForest)
library(tibble)
library(writexl)

dal <- readRDS(here("01_Preprocess", "02_Normalization", "c_data", "DAList_normalized.rds"))
out <- here("01_Preprocess", "03_Imputation", "c_data")
dir.create(out, recursive = TRUE, showWarnings = FALSE)
mat <- as.matrix(dal$data)
c(proteins = nrow(mat), samples = ncol(mat), pct_missing = round(100 * mean(is.na(mat)), 1))
```

## Impute

Random-forest imputation (Stekhoven & Bühlmann 2012). Rows are sorted before the stochastic fit
so the result reproduces. The backend is named: missForest 1.6.1 made ranger the default, which
is not the randomForest engine the paper describes.

```{r impute}
set.seed(42)
ord <- order(rownames(mat))
mf <- missForest(t(mat[ord, ]), maxiter = 10, ntree = 100, verbose = FALSE, backend = "ranger")
imputed <- t(mf$ximp)[rownames(mat), ]
dimnames(imputed) <- dimnames(mat)
stopifnot(!anyNA(imputed), identical(dim(imputed), dim(mat)))
c(oob_nrmse = round(unname(mf$OOBerror[1]), 4))
```

## Outputs

```{r write}
dal$data <- imputed
dal$imputation <- list(
  method = "missForest", backend = "ranger", maxiter = 10, ntree = 100,
  oob_error = unname(mf$OOBerror[1])
)
saveRDS(dal, file.path(out, "DAList_imputed.rds"))
write_xlsx(
  list(imputation = tibble(
    method = "missForest", backend = "ranger", maxiter = 10, ntree = 100,
    pct_imputed = 100 * mean(is.na(mat)), oob_nrmse = unname(mf$OOBerror[1])
  )),
  file.path(out, "03_impute.xlsx")
)
```

```{r session}
sessionInfo()
```
