---
title: "03 · Imputation"
subtitle: "missForest, for the methods that need a complete matrix"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    df-print: kable
    fig-width: 7
    fig-height: 5
    fig-dpi: 150
    fig-align: center
execute:
  message: false
---

The differential fit reads the unimputed matrix. Three later methods cannot: `fry` rotates
residuals and takes no missing value, singscore ranks every protein in a sample, and WGCNA
correlates complete columns. `03_Pathway_Enrichment/05` reads the sample sheet from it.

```{r setup}
library(here)
library(missForest)
library(tibble)
library(writexl)

inputs <- c(normalized = "01_Preprocess/02_Normalization/c_data/DAList_normalized.rds")
paths <- vapply(inputs, here, character(1))
stopifnot(file.exists(paths))
dal <- readRDS(paths[["normalized"]])
out <- here("01_Preprocess", "03_Imputation", "c_data")
dir.create(out, recursive = TRUE, showWarnings = FALSE)
mat <- as.matrix(dal$data)
c(proteins = nrow(mat), samples = ncol(mat), pct_missing = round(100 * mean(is.na(mat)), 1))
```

## missForest on ranger, rows sorted, seed 42

Random-forest imputation (Stekhoven & Bühlmann 2012). Rows are sorted before the stochastic fit
so the result reproduces. The backend is ranger, named rather than left to the default, so a
change in missForest's default cannot swap engines between runs.

```{r impute}
params <- list(method = "missForest", backend = "ranger", maxiter = 10, ntree = 100)
set.seed(42)
ord <- order(rownames(mat))
mf <- missForest(
  t(mat[ord, ]),
  maxiter = params$maxiter, ntree = params$ntree, backend = params$backend, verbose = FALSE
)
imputed <- t(mf$ximp)[rownames(mat), ]
dimnames(imputed) <- dimnames(mat)
stopifnot(!anyNA(imputed), identical(dim(imputed), dim(mat)))
c(oob_nrmse = round(unname(mf$OOBerror[1]), 4))
```

## Writes DAList_imputed.rds and the workbook

```{r write}
dal$data <- imputed
dal$imputation <- c(params, oob_error = unname(mf$OOBerror[1]))
saveRDS(dal, file.path(out, "DAList_imputed.rds"), compress = "xz")
sheets <- list(
  imputation = as_tibble(c(
    params,
    pct_imputed = 100 * mean(is.na(mat)), oob_nrmse = unname(mf$OOBerror[1])
  )),
  input_manifest = tibble(
    input = names(inputs), path = unname(inputs), md5 = unname(tools::md5sum(paths))
  ),
  package_versions = sessioninfo::package_info("loaded", dependencies = FALSE) |>
    as_tibble() |>
    dplyr::select(package, version = loadedversion, source)
)
read_me <- tibble(sheet = names(sheets), holds = c(
  "missForest settings, share of values imputed, out-of-bag error.",
  "Files read, with md5.",
  "Packages loaded at run time."
))
write_xlsx(c(list(read_me = read_me), sheets), file.path(out, "03_impute.xlsx"))
```
