---
title: "t0 · Proteomic signatures of hypertrophy response and supplement group"
format:
  html:
    toc: true
    df-print: kable
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(targets)
library(dplyr)
library(ggplot2)
# Quarto renders with cwd = t0/, so anchor the store to an absolute path and keep the
# resolved config out of the repo.
Sys.setenv(TAR_CONFIG = tempfile(fileext = ".yaml"))
tar_config_set(store = here::here("t0", "_targets"))
tar_source(here::here("t0", "R"))
cfg <- tar_read(cfg)
```

## What this pipeline asks

Two prediction targets from the same muscle proteome: composite-hypertrophy **responder class**
and **supplement group**. Each is predicted from three feature levels — protein, pathway
(singscore), module (WGCNA eigengene) — under leave-one-subject-out cross-validation with the
pathway and module transforms learned **inside each fold**, so nothing a subject predicts is derived
from that subject. Every model is scored against a label-permutation null.

## Quality control

```{r qc-pairing}
tar_read(qc_pair)
```

```{r qc-outliers}
tar_read(qc_out) |>
  filter(flagged) |>
  select(sample, cor_z, mahal, flag_cor, flag_pca)
```

Outliers are flagged, never dropped; the modeling frames keep every subject with a complete
timepoint pair.

```{r qc-fig}
#| fig-width: 7
#| fig-height: 4
fig_missingness(tar_read(qc_miss))
```

Per-sample signal spread (median intensity, IQR, coefficient of variation):

```{r qc-dist}
tar_read(qc_dist) |>
  transmute(sample, group, timepoint,
    median = round(median, 2), iqr = round(iqr, 2), cv = round(cv, 3)
  )
```

## Supervised sweep

Leave-one-subject-out AUC per (feature level × engine × contrast × target), with the
permutation-null p-value and its BH-adjusted q.

```{r sweep-table}
tar_read(sweep) |>
  transmute(contrast, target, level, model,
    AUC = round(roc_auc, 3), bal_acc = round(bal_accuracy, 3),
    p_perm = round(p_perm, 3), q = round(q_perm, 3)
  ) |>
  arrange(target, contrast, desc(AUC))
```

```{r sweep-fig}
#| fig-width: 8
#| fig-height: 6
fig_cv_performance(tar_read(sweep))
```

## Unsupervised mapping

Do whole-cohort clusters of each feature representation recover either label? Adjusted Rand index
against a 2-cluster k-means (0 = chance, 1 = perfect).

```{r unsup}
tar_read(unsupervised) |>
  transmute(contrast, level,
    ARI_responder = round(ari_responder, 3),
    ARI_supplement = round(ari_supplement, 3)
  )
```

## Reproducibility

```{r session}
tar_read(cfg)$seed
sessioninfo::session_info()$platform
```
