---
title: "02 · Differential expression"
subtitle: "Nine contrasts on the protein matrix"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    df-print: kable
    fig-width: 7
    fig-height: 5
    fig-dpi: 150
    fig-align: center
execute:
  message: false
---

proteoDA's sequence is `fit_limma_model`, `extract_DA_results`, `write_limma_plots`. Under it
sit `duplicateCorrelation`, `lmFit` with the subject block, and `eBayes(robust = TRUE)`. The fit
reads the unimputed matrix, so each protein is fitted on the samples where it was seen.

```{r setup}
library(here)
library(proteoDA)
library(limma)
library(dplyr)
library(ggplot2)
library(purrr)
library(tibble)
library(tidyr)
library(writexl)

d <- readRDS(here("02_Differential_Expression", "01_Design", "c_data", "design.rds"))
dal <- d$dal
contrast_names <- colnames(dal$design$contrast_matrix)
out <- here("02_Differential_Expression", "02_Differential", "c_data")
figures <- here("02_Differential_Expression", "02_Differential", "b_reports")
bundle <- file.path(figures, "02_differential_figures.pdf")
# Clear last render's numbered figures and bundle; write_limma_plots() output is left alone.
unlink(c(list.files(figures, "^0[0-9]_.*[.](png|pdf)$", full.names = TRUE), bundle))

# Same samples in the same order. limma compares only the counts.
stopifnot(identical(rownames(dal$design$design_matrix), colnames(dal$data)))
```

## Fit

```{r fit}
fit <- fit_limma_model(dal)
res <- extract_DA_results(fit, pval_thresh = 0.05, lfc_thresh = 0, adj_method = "BH")
eb <- fit$eBayes_fit
# write_limma_plots() knits its own R Markdown, which cannot run inside this render, so it
# gets a fresh R session.
invisible(callr::r(\(res, figures) {
  proteoDA::write_limma_plots(res,
    grouping_column = "group", output_dir = figures,
    table_columns = c("uniprot_id", "gene", "protein"), title_column = "gene", overwrite = TRUE
  )
}, args = list(res = res, figures = figures)))
c(within_subject_correlation = round(eb$correlation, 3))
```

`write_limma_plots()` writes one interactive report per contrast into `b_reports/`, with its
static plots beside them. Neither is tracked.

Effect size the primary contrast could detect at 80% power, two-sided 0.05. An empty contrast
cannot exclude effects below this.

```{r detectable-effect}
se <- eb$stdev.unscaled[, "Training_Interaction"] * sqrt(eb$s2.post)
detectable <- se * (qt(0.975, eb$df.total) + qt(0.80, eb$df.total))
round(quantile(detectable, c(0.5, 0.9), na.rm = TRUE), 3)
```

## Results

BH within each contrast, never across the nine. A protein with an empty cell in a contrast comes
back NA there and is not tested, so each contrast has its own denominator.

```{r results}
annotation <- fit$annotation[, c("uniprot_id", "gene", "protein", "description")]
stopifnot(identical(rownames(res$results[[1]]), rownames(annotation)))
results <- res$results |>
  map(\(r) bind_cols(annotation, r)) |>
  list_rbind(names_to = "contrast") |>
  as_tibble() |>
  mutate(
    pi_score = P.Value^abs(logFC),
    sig_pi = case_when(
      pi_score < 0.05 & logFC > 0 ~ 1L,
      pi_score < 0.05 & logFC < 0 ~ -1L,
      TRUE ~ 0L
    )
  )
n_obs <- rowSums(!is.na(fit$data))

contrast_summary <- results |>
  summarise(
    n_tested = sum(!is.na(P.Value)),
    n_untested = sum(is.na(P.Value)),
    n_nominal = sum(P.Value < 0.05, na.rm = TRUE),
    n_pi = sum(sig_pi != 0L),
    n_fdr_05 = sum(adj.P.Val < 0.05, na.rm = TRUE),
    n_fdr_10 = sum(adj.P.Val < 0.10, na.rm = TRUE),
    min_adj_p = min(adj.P.Val, na.rm = TRUE),
    .by = contrast
  ) |>
  left_join(d$roles, by = "contrast")
contrast_summary |> mutate(min_adj_p = signif(min_adj_p, 3))
```

## Is a contrast empty, or is it being tested conservatively

`propTrueNull()` saturates at 1, so it cannot separate a flat null from a conservative one.
Median |t| is 0.674 under a calibrated null and `p_below_0.2` is 0.2. Read the between-arm
contrasts against `Baseline_HRvLR` rather than against theory.

```{r null-calibration}
null_calibration <- results |>
  filter(!is.na(P.Value)) |>
  summarise(
    prop_true_null = round(propTrueNull(P.Value), 3),
    median_abs_t = round(median(abs(t)), 3),
    p_below_0.2 = round(mean(P.Value < 0.2), 3), .by = contrast
  )
null_calibration
```

All three summarise the same p-value histogram. A spike at zero on a flat background is signal; a
distribution sloping up toward 1 is a conservative test.

```{r p-histogram}
#| fig-height: 6
p_histogram <- results |>
  filter(!is.na(P.Value)) |>
  mutate(contrast = factor(contrast, levels = contrast_names)) |>
  ggplot(aes(P.Value)) +
  geom_histogram(binwidth = 0.05, boundary = 0, fill = "grey35") +
  geom_hline(
    aes(yintercept = h),
    results |>
      filter(!is.na(P.Value)) |>
      summarise(h = n() / 20, .by = contrast) |>
      mutate(contrast = factor(contrast, levels = contrast_names)),
    linetype = 2, colour = "firebrick"
  ) +
  facet_wrap(~contrast) +
  labs(
    title = "P-value distribution per contrast",
    subtitle = "limma via proteoDA, 20 bins", x = "p", y = "proteins",
    caption = paste(
      "Bars: proteins per p-value bin. Dashed line: the height of a flat null (tested / 20).",
      "Table: c_data/02_differential.xlsx, DEP_matrix."
    )
  ) +
  theme_minimal(base_size = 10) +
  theme(plot.caption = element_text(size = 7, colour = "grey45", hjust = 0))
p_histogram
```

## Significant and large enough

`treat()` tests against a fold-change floor fixed in advance at 1.15, and controls an error rate,
unlike filtering on logFC afterwards. `robust = TRUE` matches the main fit.

```{r treat}
ft <- treat(eb, lfc = log2(1.15), robust = TRUE)
treat_counts <- tibble(
  contrast = contrast_names,
  n_q_05 = map_int(contrast_names, \(cn) {
    sum(topTreat(ft, coef = cn, number = Inf)$adj.P.Val < 0.05, na.rm = TRUE)
  })
)
treat_counts
```

## The pi score

`pi_score` is `P.Value^abs(logFC)`, the Xiao et al. (2014) score written inverted. It controls no
error rate: it ranks within a contrast and selects nothing. Large fold changes come from sparsely
observed proteins, so each row carries the number of samples the protein was seen in. Applied to
the four within-arm contrasts only.

```{r pi-ranking}
pi_ranking <- results |>
  filter(contrast %in% c("Training_HR", "Training_LR", "Acute_HR", "Acute_LR")) |>
  slice_min(pi_score, n = 10, by = contrast, with_ties = FALSE) |>
  mutate(n_obs = n_obs[uniprot_id]) |>
  select(contrast, uniprot_id, gene, logFC, adj.P.Val, pi_score, n_obs)
pi_ranking
```

## Protein classification

How well one protein separates two sets of samples. Eight tasks mirror the contrasts. The four
within-arm tasks pair each subject with itself; the four between-arm tasks compare subjects on a
level or on their own change. AUC comes from `pROC` with `direction = "<"` fixed, so above 0.5
means higher in the later timepoint or in HR. The p comes from the Wilcoxon test, signed-rank
when paired. A protein needs three observations per side.

```{r tasks}
meta <- fit$metadata
by_subject <- function(mat, timepoint, arm = NULL) {
  rows <- filter(meta, time == timepoint, is.null(arm) | responder %in% arm)
  set_names(as.data.frame(mat[, rows$sample_id, drop = FALSE]), rows$subject) |> as.matrix()
}
change <- function(mat, from, to) {
  a <- by_subject(mat, from)
  b <- by_subject(mat, to)
  both <- intersect(colnames(a), colnames(b))
  b[, both, drop = FALSE] - a[, both, drop = FALSE]
}
arm_of <- set_names(as.character(meta$responder), meta$subject)
within_arm <- function(mat, from, to, arm) {
  a <- by_subject(mat, from, arm)
  b <- by_subject(mat, to, arm)
  both <- intersect(colnames(a), colnames(b))
  list(positive = b[, both, drop = FALSE], negative = a[, both, drop = FALSE], paired = TRUE)
}
between_arm <- function(values) {
  list(
    positive = values[, arm_of[colnames(values)] == "HR", drop = FALSE],
    negative = values[, arm_of[colnames(values)] == "LR", drop = FALSE],
    paired = FALSE
  )
}
build_tasks <- function(mat) {
  list(
    Training_HR = within_arm(mat, "T1", "T2", "HR"),
    Training_LR = within_arm(mat, "T1", "T2", "LR"),
    Acute_HR = within_arm(mat, "T2", "T3", "HR"),
    Acute_LR = within_arm(mat, "T2", "T3", "LR"),
    Baseline_HRvLR = between_arm(by_subject(mat, "T1")),
    Trained_HRvLR = between_arm(by_subject(mat, "T2")),
    Training_change_HRvLR = between_arm(change(mat, "T1", "T2")),
    Acute_change_HRvLR = between_arm(change(mat, "T2", "T3"))
  )
}

classify_row <- function(pos, neg, paired) {
  if (paired) {
    ok <- !is.na(pos) & !is.na(neg)
    pos <- pos[ok]
    neg <- neg[ok]
  } else {
    pos <- pos[!is.na(pos)]
    neg <- neg[!is.na(neg)]
  }
  if (min(length(pos), length(neg)) < 3) {
    return(c(auc = NA, p = NA, n_positive = length(pos), n_negative = length(neg)))
  }
  roc <- pROC::roc(controls = neg, cases = pos, direction = "<", quiet = TRUE)
  # Ties make wilcox.test fall back to its normal approximation and warn on every call.
  p <- suppressWarnings(wilcox.test(pos, neg, paired = paired)$p.value)
  c(auc = as.numeric(pROC::auc(roc)), p = p, n_positive = length(pos), n_negative = length(neg))
}
classify <- function(spec) {
  fits <- vapply(seq_len(nrow(spec$positive)), \(i) {
    classify_row(spec$positive[i, ], spec$negative[i, ], spec$paired)
  }, numeric(4))
  tibble(uniprot_id = rownames(spec$positive), paired = spec$paired, !!!as_tibble(t(fits)))
}

# The labels are shared with the set and module levels, so the three chance tables line up.
task_labels <- c(
  Training_HR = "Training, HR (T1 to T2)", Training_LR = "Training, LR (T1 to T2)",
  Acute_HR = "Acute bout, HR (T2 to T3)", Acute_LR = "Acute bout, LR (T2 to T3)",
  Baseline_HRvLR = "HR vs LR at T1 (floor)", Trained_HRvLR = "HR vs LR at T2",
  Training_change_HRvLR = "HR vs LR, training change",
  Acute_change_HRvLR = "HR vs LR, acute change"
)
tasks <- build_tasks(fit$data)
stopifnot(identical(names(tasks), names(task_labels)))
protein_auc <- imap(tasks, \(spec, name) {
  mutate(classify(spec), task = name, task_label = task_labels[[name]], .before = 1)
}) |>
  list_rbind() |>
  mutate(fdr = p.adjust(p, "BH"), .by = task) |>
  left_join(select(annotation, uniprot_id, gene), by = "uniprot_id") |>
  relocate(gene, .after = uniprot_id)

chance_expectation <- protein_auc |>
  filter(!is.na(p)) |>
  summarise(tested = n(), nominal = sum(p < 0.05), fdr_sig = sum(fdr < 0.05), .by = task_label) |>
  transmute(
    analysis = "classification", comparison = task_label, tested, nominal,
    expected = 0.05 * tested, ratio = round(nominal / expected, 2), fdr_sig
  )
chance_expectation
```

## Hit matrices

Every protein nominal in at least one column, as a dot matrix: fill is the effect, size is
−log10 p, a black ring marks BH < 0.05. 75 rows to a page.

```{r hit-matrices}
labels <- annotation |>
  mutate(label = if_else(
    is.na(gene) | duplicated(gene) | duplicated(gene, fromLast = TRUE),
    paste0(coalesce(gene, "?"), " (", uniprot_id, ")"), gene
  ))
label_of <- set_names(labels$label, labels$uniprot_id)

save_figure <- function(figure, name, width = 11, height = 8.5) {
  walk(c("png", "pdf"), \(extension) {
    ggsave(file.path(figures, paste0(name, ".", extension)), figure,
      width = width, height = height, dpi = 200, bg = "white"
    )
  })
}
# A paged figure is one PDF; a single-page one also gets a PNG.
save_pages <- function(pages, name, width = 11, height = 8.5) {
  if (length(pages) == 1) save_figure(pages[[1]], name, width, height)
  if (length(pages) > 1) {
    pdf(file.path(figures, paste0(name, ".pdf")), width = width, height = height, bg = "white")
    walk(pages, print)
    invisible(dev.off())
  }
}
hit_pages <- function(data, columns, title, subtitle, fill_label, table, midpoint = 0,
                      per_page = 75) {
  ranked <- data |>
    filter(!is.na(p)) |>
    summarise(n_nominal = sum(p < 0.05), best = min(p), .by = label) |>
    filter(n_nominal > 0) |>
    arrange(desc(n_nominal), best)
  page_of <- ceiling(seq_len(nrow(ranked)) / per_page)
  map(seq_len(max(page_of, 0)), \(page) {
    rows <- ranked$label[page_of == page]
    shown <- data |>
      filter(label %in% rows) |>
      mutate(
        label = factor(label, levels = rev(rows)),
        column = factor(column, levels = columns),
        nominal = !is.na(p) & p < 0.05
      )
    ggplot(shown, aes(column, label)) +
      geom_point(data = filter(shown, !nominal), colour = "grey85", size = 0.6) +
      geom_point(
        data = filter(shown, nominal), aes(size = -log10(p), fill = effect),
        shape = 21, colour = "grey30", stroke = 0.2
      ) +
      geom_point(
        data = filter(shown, !is.na(fdr), fdr < 0.05), aes(size = -log10(p)),
        shape = 21, colour = "black", stroke = 0.9, fill = NA
      ) +
      scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B", midpoint = midpoint) +
      scale_size_continuous(range = c(1, 4.5), limits = c(-log10(0.05), NA)) +
      scale_x_discrete(drop = FALSE) +
      labs(
        title = sprintf("%s (%d/%d)", title, page, max(page_of)),
        subtitle = sprintf(
          "%s; %d proteins nominal in at least one column; rows %d-%d",
          subtitle, nrow(ranked), min(which(page_of == page)), max(which(page_of == page))
        ),
        x = NULL, y = NULL, fill = fill_label, size = "-log10 p",
        caption = paste0(
          "Rows: proteins at nominal p < 0.05 in at least one column, most columns first, then ",
          "by best p. Filled dot: nominal, fill = ", fill_label, ", size = -log10 p. Black ring: ",
          "BH < 0.05. Grey speck: tested, not nominal. Table: ", table, "."
        )
      ) +
      theme_minimal(base_size = 9) +
      theme(
        axis.text.x = element_text(angle = 35, hjust = 1),
        axis.text.y = element_text(size = if (length(rows) > 30) 5.5 else 8),
        plot.caption = element_text(size = 7, colour = "grey45", hjust = 0)
      )
  })
}

contrast_hits <- hit_pages(
  transmute(results,
    label = label_of[uniprot_id], column = contrast,
    effect = logFC, p = P.Value, fdr = adj.P.Val
  ),
  contrast_names,
  title = "Protein contrast hits", subtitle = "limma via proteoDA, nominal p per contrast",
  fill_label = "log2 FC", table = "c_data/02_differential.xlsx, DEP_matrix"
)
task_hits <- hit_pages(
  transmute(protein_auc,
    label = label_of[uniprot_id], column = task,
    effect = auc, p = p, fdr = fdr
  ),
  names(tasks),
  title = "Protein classification hits", subtitle = "pROC AUC, Wilcoxon p per task",
  fill_label = "AUC", midpoint = 0.5, table = "c_data/02_differential.xlsx, protein_auc"
)
c(contrast_pages = length(contrast_hits), task_pages = length(task_hits))
```

## Outputs

The fit and the figures are written before the floor is checked, so a failure leaves them on disk.
The workbook follows the two appendices, whose tables it carries.

```{r write}
dep_matrix <- results |>
  mutate(n_obs = n_obs[uniprot_id]) |>
  select(uniprot_id, gene, n_obs, contrast, logFC, P.Value, adj.P.Val) |>
  pivot_wider(
    names_from = contrast, values_from = c(logFC, P.Value, adj.P.Val),
    names_glue = "{contrast}_{.value}"
  )

saveRDS(res, file.path(out, "fit.rds"))
save_figure(p_histogram, "01_p_histogram", height = 7)
save_pages(contrast_hits, "02_contrast_hits")
save_pages(task_hits, "03_classification_hits")
```

## Floor

`Baseline_HRvLR` compares the arms before training. The count is printed, not asserted.

```{r floor}
floor_rows <- filter(results, contrast == d$floor)
# sum(logical(0)) is zero, so a contrast name matching nothing would look like a clean floor.
stopifnot(nrow(floor_rows) == nrow(fit$data))
sum(floor_rows$adj.P.Val < 0.05, na.rm = TRUE)
slice_min(floor_rows, P.Value, n = 5) |> select(uniprot_id, gene, logFC, P.Value, adj.P.Val)
```

## Appendix: V1 equivalence

The fit reproduces V1's committed numbers. Skipped when the sibling V1 tree is absent.

```{r v1}
v1_book <- file.path(
  dirname(here()), "HRvLR_V1", "03_Features", "01_Proteins", "c_data",
  "05_results.xlsx"
)
v1_equivalence <- if (file.exists(v1_book)) {
  map(contrast_names, \(cn) {
    reference <- readxl::read_excel(v1_book, sheet = cn)
    joined <- inner_join(
      filter(results, contrast == cn),
      transmute(reference, uniprot_id, ref_lfc = logFC, ref_p = P.Value, ref_bh = adj.P.Val),
      by = "uniprot_id"
    )
    tibble(
      contrast = cn, matched = nrow(joined),
      max_abs_diff = max(abs(c(
        joined$logFC - joined$ref_lfc, joined$P.Value - joined$ref_p,
        joined$adj.P.Val - joined$ref_bh
      )), na.rm = TRUE)
    )
  }) |>
    list_rbind()
} else {
  tibble(contrast = character(), matched = integer(), max_abs_diff = numeric())
}
v1_equivalence
```

## Appendix: does one arm's training signature move in the other

The proteins that responded in one arm, tested as a set in the other arm's contrast with `fry`.
Only the mirrored within-arm pairs share no cell mean, so only they are tested. `fry` takes no
missing value, so it runs on the imputed matrix with the same design, block and a correlation
estimated on that matrix.

```{r fry-concordance}
imputed <- readRDS(here("01_Preprocess", "03_Imputation", "c_data", "DAList_imputed.rds"))
abundance <- as.matrix(imputed$data)
design <- fit$design$design_matrix
stopifnot(
  identical(colnames(abundance), rownames(design)), identical(meta$sample_id, rownames(design)),
  !anyNA(abundance)
)
fry_correlation <- duplicateCorrelation(abundance, design, block = meta$subject)
fry_correlation <- fry_correlation$consensus.correlation

member_rows <- function(cn, criterion, direction) {
  r <- filter(results, contrast == cn)
  sign_ok <- if (direction == "up") r$logFC > 0 else r$logFC < 0
  hit <- switch(criterion,
    pi = r$sig_pi == if (direction == "up") 1L else -1L,
    bh = !is.na(r$adj.P.Val) & r$adj.P.Val < 0.05 & sign_ok,
    nominal = !is.na(r$P.Value) & r$P.Value < 0.05 & sign_ok
  )
  which(rownames(abundance) %in% r$uniprot_id[coalesce(hit, FALSE)])
}
pairings <- tribble(
  ~set_from, ~ranked_on,
  "Training_HR", "Training_LR",
  "Training_LR", "Training_HR",
  "Acute_HR", "Acute_LR",
  "Acute_LR", "Acute_HR"
)
selections <- expand_grid(criterion = c("pi", "bh", "nominal"), direction = c("up", "down"))
fry_concordance <- pmap(pairings, \(set_from, ranked_on) {
  sets <- pmap(selections, \(criterion, direction) member_rows(set_from, criterion, direction)) |>
    set_names(paste(selections$criterion, selections$direction, sep = "_"))
  sets <- keep(sets, \(rows) length(rows) >= 5)
  # fry() drops its FDR column when handed a single set, so one set is not tested.
  if (length(sets) < 2) {
    return(NULL)
  }
  fry(abundance,
    index = sets, design = design, contrast = fit$design$contrast_matrix[, ranked_on],
    block = meta$subject, correlation = fry_correlation
  ) |>
    as_tibble(rownames = "set") |>
    transmute(set_from, ranked_on, set, n = NGenes, direction = Direction, p = PValue, fdr = FDR)
}) |>
  list_rbind()
arrange(fry_concordance, p)
```

```{r write-workbook}
write_xlsx(
  list(
    DEP_matrix = dep_matrix, contrast_summary = contrast_summary,
    null_calibration = null_calibration, treat = treat_counts, pi_ranking = pi_ranking,
    protein_auc = arrange(protein_auc, p), chance_expectation = chance_expectation,
    v1_equivalence = v1_equivalence, fry_concordance = fry_concordance
  ),
  file.path(out, "02_differential.xlsx")
)
saveRDS(
  list(
    contrast_summary = contrast_summary, null_calibration = null_calibration,
    treat = treat_counts, pi_ranking = pi_ranking, protein_auc = protein_auc,
    chance_expectation = chance_expectation, fry_concordance = fry_concordance,
    provenance = list(
      created_utc = format(Sys.time(), tz = "UTC", usetz = TRUE),
      inputs = tibble(
        input = "design", path = "02_Differential_Expression/01_Design/c_data/design.rds",
        md5 = unname(tools::md5sum(here(path)))
      )
    )
  ),
  file.path(out, "differential.rds"),
  compress = "xz"
)
pages <- sort(list.files(figures, "^0[0-9]_.*[.]pdf$", full.names = TRUE))
invisible(qpdf::pdf_combine(pages, bundle))
length(pages)
```

```{r session}
sessionInfo()
```
