---
title: "F05 prediction — WGCNA co-expression modules"
subtitle: "Responder class and training gains, out of sample: nothing, and even the upper bound is nothing"
author: "A_HRvLR_2026"
format:
  html:
    toc: true
    embed-resources: true
    df-print: paged
execute:
  eval: true
  echo: false
  warning: false
  message: false
---

## In plain terms

Can the proteome, summarised as WGCNA co-expression modules, predict a
held-out person? We train on 15 people and predict the 16th, rotating
through everyone, for two questions: which responder group they're in, and
how much they actually gained.

## Method

Each fold is a nested leave-one-subject-out: features are z-scored on the
training people only and that centre and scale are applied to the held-out
person, so nothing about the held-out subject touches the transform.
glmnet and sparse-PLS are tuned by an inner leave-one-out on the training
people alone, and the winning setting is refit on all of them before
scoring the held-out subject. A permutation null shuffles the outcome
across subjects and re-runs the entire nested procedure, giving each cell
its own p-value; Benjamini-Hochberg correction is then applied across the
whole metric grid. n = 16, exploratory.

**The module eigengenes here are derived from the whole cohort**, not
re-derived inside each training fold — the WGCNA network itself is fit
once, upstream, in F04_association. Only the downstream prediction step
(scaling, model tuning, scoring) is nested and leakage-free; the features
being predicted from already contain a trace of every subject, including
the held-out one. Any positive result in this family is therefore an
**optimistic upper bound**, not a clean out-of-sample estimate.

## Results

```{r}
#| label: load-results
pacman::p_load(here, openxlsx, dplyr)

class_metrics <- read.xlsx(
  here(
    "04_Figures", "F05_prediction", "WGCNA", "responder", "c_data",
    "class_source_data.xlsx"
  ),
  "metrics"
)
cont_metrics <- read.xlsx(
  here(
    "04_Figures", "F05_prediction", "WGCNA", "continuous", "c_data",
    "cont_source_data.xlsx"
  ),
  "metrics"
)

best_class <- class_metrics |> slice_max(estimate, n = 1, with_ties = FALSE)
best_cont <- cont_metrics |> slice_max(q2, n = 1, with_ties = FALSE)

n_class_cells <- nrow(class_metrics)
n_cont_cells <- nrow(cont_metrics)
n_class_hits <- sum(class_metrics$q_bh < 0.05, na.rm = TRUE)
n_cont_hits <- sum(cont_metrics$q_bh < 0.05, na.rm = TRUE)
```

Across the `r n_class_cells` phase-by-model cells that predict responder
class, the best is `r best_class$model` on the `r best_class$phase` phase:
AUC = `r sprintf("%.2f", best_class$estimate)` (permutation
p = `r sprintf("%.2f", best_class$perm_p)`, BH q = `r sprintf("%.2f", best_class$q_bh)`).
An AUC that low is not even a candidate hit, and its own null confirms it.

Across the `r n_cont_cells` outcome-by-phase-by-model cells that predict a
training gain, the best Q² anywhere is
`r sprintf("%.3f", best_cont$q2)` (`r best_cont$outcome`, `r best_cont$phase`
phase, `r best_cont$model`; permutation p = `r sprintf("%.2f", best_cont$perm_p_q2)`,
BH q = `r sprintf("%.2f", best_cont$q_bh)`). A negative Q² means the model
predicts a held-out subject worse than that subject's own group mean
would. `r n_class_hits` of `r n_class_cells` responder-class cells and
`r n_cont_hits` of `r n_cont_cells` continuous cells pass BH < 0.05.

**Nothing beats its permutation null, even with a whole-cohort-derived
feature space stacking the deck in its favour.** The proteome, summarised
as co-expression modules, does not predict responder class or training
gains out of sample. This agrees with the association null reported
elsewhere in F04 and with the study-wide null: the leaky raw-protein
feature space that an earlier version of this analysis used was dropped
rather than reported alongside a result it could not honestly support.

```{r sessioninfo}
sessionInfo()
```
