---
title: "HRvLR 00 — Inputs"
subtitle: "The raw DIA-MS matrix, sample sheet, phenotype table, and HPA annotations"
author: "A_HRvLR_2026"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
    embed-resources: true
    df-print: paged
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
pacman::p_load(here, readr, readxl, dplyr)
knitr::opts_knit$set(root.dir = here::here())
```

Stage 00 holds the four files the pipeline reads and never writes. Nothing is
computed here. This page records what each file is, why it is shaped the way it
is, and which downstream stage consumes it, so the contract is explicit before
any filtering or modelling begins.

The study is a two-group repeated-measures design: eight high responders (HR)
and eight low responders (LR), each sampled at three timepoints (T1 baseline,
T2 post-training, T3 post-acute-bout). Two subjects are partial, so 48
subject-timepoint slots reduce to 45 quantified samples downstream.

# The protein matrix: `HRvLR_raw.xlsx`

The DIA-MS quantification, one row per protein group and one column per sample,
plus leading annotation columns.

```{r}
#| label: raw
raw <- read_excel(here("00_input", "HRvLR_raw.xlsx"))
dim(raw)
names(raw)[1:8]
```

The first five columns identify the protein (`uniprot_id`, `protein`, `gene`,
`description`, `n_seq`); the remaining columns are intensities named
`Subject_Timepoint` (`S03_T1`, `S03_T2`, …). proteoDA does not read a wide sheet
directly — stage 01 splits it into the annotation table and the intensity matrix
and pairs the intensity columns to the sample sheet by name. Keeping the sample
columns named exactly as `Col_ID` in the metadata is what lets that pairing
succeed without a manual key.

Intensities arrive linear, with unquantified proteins recorded as zero. Stage 01
converts those zeros to `NA` (`zero_to_missing`) because the missing-value filter
and, later, limma treat absence as missing rather than as a measured zero.

# The sample sheet: `HRvLR_meta.csv`

One row per sample-timepoint. The left block defines the experimental design; the
right block carries the phenotype measurements used by the Figure 1 atlas.

```{r}
#| label: meta
meta <- read_csv(here("00_input", "HRvLR_meta.csv"), show_col_types = FALSE)
dim(meta)
names(meta)
```

- `Col_ID` matches the intensity column names in the protein matrix; it is the
  join key between data and metadata.
- `Subject_ID` groups the three timepoints of one person. It becomes the
  blocking factor in stage 03, where `duplicateCorrelation` uses it to account
  for repeated measures on the same subject.
- `Group` (HR/LR), `Timepoint` (T1/T2/T3), and their interaction `Group_Time`
  (six cells: `HR_T1` … `LR_T3`) define the cell-means design. `Group_Time` is
  also the grouping column for the missingness filter, so a protein is kept when
  it is quantified in enough samples of at least one timepoint cell.
- The phenotype columns (`ACCUM_VL`, `mCSA_Pre`, the `fCSA_*` and
  `MyoVision_*` fibre areas, `COMP.HYPERTROPHY`) are strength and
  cross-sectional-area readouts. Figure 1 reads them directly; the proteomics
  stages ignore them.

# The responder definition: `Phenotype.xlsx`

```{r}
#| label: pheno
excel_sheets(here("00_input", "Phenotype.xlsx"))
```

The `RES&NR data` sheet is the source for the HR/LR split. Group membership is
defined by the `COMP.HYPERTROPHY` composite, not assigned by proteome — Figure 1
states this openly so the composite is read as the grouping variable rather than
as an outcome the proteome predicts.

# The contaminant reference: `HPA_annotations.tsv`

Human Protein Atlas evidence, keyed by UniProt, that stage 01 uses to remove
blood-plasma leakage from a muscle-biopsy proteome.

```{r}
#| label: hpa
hpa <- read_tsv(here("00_input", "HPA_annotations.tsv"), show_col_types = FALSE)
names(hpa)
```

A muscle biopsy carries blood, so abundant plasma and erythrocyte proteins can
masquerade as muscle signal. The gate combines several HPA lines of evidence:
secretome/immunoglobulin status and erythrocyte marker RNA flag a protein as
blood-derived, while a high myonuclei RNA value (`Single Cell Type RNA -
Myonuclei`) with a low measured blood concentration rescues genuine muscle
proteins. A protein is dropped only when it looks blood-derived and does not
clear the muscle-rescue test.

`HPA_skeletal_muscle_annotations.tsv` sits alongside as a muscle-specific HPA
reference. The current filter does not read it; it is retained for provenance and
possible future use.

# What stage 01 reads

| File | Read by | Purpose |
|:--|:--|:--|
| `HRvLR_raw.xlsx` | `01_run_filtering.R` | protein intensities + annotation |
| `HRvLR_meta.csv` | `01_run_filtering.R` | sample sheet, design, phenotype |
| `HPA_annotations.tsv` | `01_run_filtering.R` | blood-contaminant evidence |
| `Phenotype.xlsx` | Figure 1 | responder grouping, phenotype source |

Stage 01 turns the matrix and sample sheet into a filtered `DAList`, the object
every later stage builds on.
