---
title: "HRvLR 00 — Inputs"
subtitle: "The raw DIA-MS matrix, sample sheet, phenotype table, and HPA annotations"
author: "A_HRvLR_2026"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
    embed-resources: true
    df-print: paged
bibliography: ../../04_Figures/shared/references.bib
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
pacman::p_load(here, readr, readxl, dplyr)
knitr::opts_knit$set(root.dir = here::here())
```

Stage 00 holds the four files the pipeline reads and never writes. Nothing is
computed here. This page records what each file is, why it is shaped the way it
is, and which downstream stage consumes it, so the contract is explicit before
any filtering or modelling begins.

The study is a two-group repeated-measures design: eight high responders (HR)
and eight low responders (LR), each sampled at three timepoints (T1 baseline,
T2 post-training, T3 post-acute-bout). Two subjects are partial, so 48
subject-timepoint slots reduce to 45 quantified samples downstream.

# What DIA-MS is, and what it means for the matrix

The subtitle says DIA, so it is worth saying what that is before reading a single
number. In data-independent acquisition the mass spectrometer marches through a
fixed ladder of m/z windows and fragments *everything* inside each window, every
cycle, in every run. The older alternative, data-dependent acquisition (DDA),
instead looks at each survey scan and picks the top-N most intense precursors to
fragment. That choice is made on the fly and depends on what else happened to be
eluting at that instant, so two runs of the same sample can fragment different
peptides.

The consequence for us is a matter of consistency. Because the DIA windows are
fixed in advance, the same peptide is targeted in every run, and quantification
across the 45 samples is far more consistent than DDA would give. What DIA does
not do is fill the matrix. There are still gaps, and the interpretation of a gap
is specific: it almost always means the protein fell below detection in that run,
not that the protein was absent from that person's muscle. That is the whole
reason stage 01 converts a reported zero to `NA`. A zero would tell the model "we
measured this and it was nothing," which is a claim the instrument never made.

One more piece of vocabulary. A row in the matrix is a **protein group**, not
necessarily a gene. Mass spectrometry sees peptides, and when several proteins
share the peptides that were actually observed, no amount of arithmetic can tell
them apart. Those indistinguishable proteins are reported together as one group,
carrying one intensity. So a row is a group of proteins that the evidence cannot
separate; most of the time it maps to one gene, but the identifier is a claim
about what was measurable, not about what was present.

# The protein matrix: `HRvLR_raw.xlsx`

The DIA-MS quantification, one row per protein group and one column per sample,
plus leading annotation columns.

```{r}
#| label: raw
raw <- read_excel(here("00_input", "HRvLR_raw.xlsx"))
dim(raw)
names(raw)[1:8]
```

The first five columns identify the protein (`uniprot_id`, `protein`, `gene`,
`description`, `n_seq`); the remaining columns are intensities named
`Subject_Timepoint` (`S03_T1`, `S03_T2`, …). proteoDA does not read a wide sheet
directly — stage 01 splits it into the annotation table and the intensity matrix
and pairs the intensity columns to the sample sheet by name. Keeping the sample
columns named exactly as `Col_ID` in the metadata is what lets that pairing
succeed without a manual key.

Intensities arrive linear, with unquantified proteins recorded as zero. Stage 01
converts those zeros to `NA` (`zero_to_missing`) because the missing-value filter
and, later, limma treat absence as missing rather than as a measured zero.

# The sample sheet: `HRvLR_meta.csv`

One row per sample-timepoint. The left block defines the experimental design; the
right block carries the phenotype measurements used by the Figure 1 atlas.

```{r}
#| label: meta
meta <- read_csv(here("00_input", "HRvLR_meta.csv"), show_col_types = FALSE)
dim(meta)
names(meta)
```

- `Col_ID` matches the intensity column names in the protein matrix; it is the
  join key between data and metadata.
- `Subject_ID` groups the three timepoints of one person. It becomes the
  blocking factor in stage 03, where `duplicateCorrelation` uses it to account
  for repeated measures on the same subject.
- `Group` (HR/LR), `Timepoint` (T1/T2/T3), and their interaction `Group_Time`
  (six cells: `HR_T1` … `LR_T3`) define the cell-means design. `Group_Time` is
  also the grouping column for the missingness filter, so a protein is kept when
  it is quantified in enough samples of at least one timepoint cell.
- The phenotype columns (`ACCUM_VL`, `mCSA_Pre`, the `fCSA_*` and
  `MyoVision_*` fibre areas, `COMP.HYPERTROPHY`) are strength and
  cross-sectional-area readouts. Figure 1 reads them directly; the proteomics
  stages ignore them.

# The responder definition: `Phenotype.xlsx`

```{r}
#| label: pheno
excel_sheets(here("00_input", "Phenotype.xlsx"))
```

The `RES&NR data` sheet is the source for the HR/LR split. Group membership is
defined by the `COMP.HYPERTROPHY` composite, not assigned by proteome — Figure 1
states this openly so the composite is read as the grouping variable rather than
as an outcome the proteome predicts.

# The contaminant reference: `HPA_annotations_full.tsv`

Human Protein Atlas evidence [@uhlen2015], keyed by UniProt, that stage 01 uses
to remove blood-plasma leakage from a muscle-biopsy proteome.

```{r}
#| label: hpa
hpa <- read_tsv(here("00_input", "HPA_annotations_full.tsv"), show_col_types = FALSE)
names(hpa)
```

A muscle biopsy carries blood, so abundant plasma and erythrocyte proteins can
masquerade as muscle signal. The gate combines several HPA lines of evidence:
secretome/immunoglobulin status and erythrocyte marker RNA flag a protein as
blood-derived, while a high myonuclei RNA value (`Single Cell Type RNA -
Myonuclei`) with a low measured blood concentration rescues genuine muscle
proteins. A protein is dropped only when it looks blood-derived and does not
clear the muscle-rescue test.

`HPA_skeletal_muscle_annotations.tsv` sits alongside as a muscle-specific HPA
reference. The current filter does not read it; it is retained for provenance and
possible future use.

# What stage 01 reads

| File | Read by | Purpose |
|:--|:--|:--|
| `HRvLR_raw.xlsx` | `01_run_filtering.R` | protein intensities + annotation |
| `HRvLR_meta.csv` | `01_run_filtering.R` | sample sheet, design, phenotype |
| `HPA_annotations_full.tsv` | `01_run_filtering.R` | blood-contaminant evidence |
| `Phenotype.xlsx` | Figure 1 | responder grouping, phenotype source |

Stage 01 turns the matrix and sample sheet into a filtered `DAList`, the object
every later stage builds on. A `DAList` is proteoDA's container [@thurman2023]:
three name-aligned tables held together, `data` (the intensity matrix, one row per
protein and one column per sample), `annotation` (one row per protein), and
`metadata` (one row per sample). The alignment is the point — the row names of
`data` match `annotation`, the column names of `data` match the row names of
`metadata`, and every proteoDA filter preserves that correspondence, so a protein
can never drift away from its own annotation. Stage 01 builds it and defines it in
full; it is named here because everything downstream of this page speaks in terms
of it.

## Key terms

**DIA (data-independent acquisition)** — the mass spectrometer fragments
everything inside a fixed ladder of m/z windows on every cycle, rather than
choosing the most intense precursors as it goes. Quantification is consistent
across runs, but the matrix still has gaps.

**DDA (data-dependent acquisition)** — the older alternative, which picks the
top-N most intense precursors in each survey scan. Which peptides get fragmented
depends on what happened to be eluting, so runs disagree with each other.

**Protein group** — one row of the matrix. A set of proteins that share the
observed peptides and therefore cannot be told apart. The row is the group, not
guaranteed to be a single gene.

**Missing, not zero** — an unquantified protein arrives as a zero and becomes
`NA` in stage 01. The gap usually means "below detection in this run", not
"absent from this muscle", and a zero would assert a measurement that was never
made.

**`Col_ID`** — the join key. It names an intensity column in the protein matrix
and a row in the sample sheet, which is what pairs data to design without a
manual key.

**`Group_Time`** — the six-cell interaction factor (`HR_T1` … `LR_T3`). It is the
cell-means design in stage 03 and the grouping column for the missingness filter
in stage 01.

**`DAList`** — proteoDA's three-table container (`data`, `annotation`,
`metadata`), name-aligned so filtering can never desynchronize a protein from its
annotation or a sample from its design.

## Recap

- Stage 00 is read-only. Four files enter the pipeline here and nothing is
  computed or written.
- The design is 8 HR and 8 LR across three timepoints; two partial subjects turn
  48 slots into 45 quantified samples.
- DIA gives consistent quantification but a gapped matrix, and the gaps mean
  "below detection", which is why zeros become `NA`.
- `Col_ID` pairs the matrix to the sample sheet, and `Group_Time` carries the six
  design cells used by both the missingness filter and the models.
- HPA evidence is the raw material for the blood-contaminant rule; `Phenotype.xlsx`
  defines the HR/LR split and is read only by Figure 1.
- Everything downstream is a `DAList`, which stage 01 builds.

```{r sessioninfo}
sessionInfo()
```

## References

::: {#refs}
:::
