---
title: "F02 — Proteome Overview (HR vs LR)"
subtitle: "Shared global response, protein-specific leads"
author: "A_HRvLR_2026"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
    embed-resources: true
    df-print: paged
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

## What F02 establishes

F02 asks whether high and low responders differ across the whole proteome, not
just protein by protein. The answer is no: every global test — the proteome
state, the size of the training and acute response, and its direction and spread —
comes back null. The only HR-vs-LR signal is protein-specific, and even that is a
list of exploratory leads that no protein clears at FDR. So the honest headline is
a **shared global response with protein-specific leads**, stated at the scale the
data actually support.

```{r setup}
source(here::here("04_Figures", "F02", "a_script", "HRvLR_F02_setup.R"))
panel <- function(name) {
  source(here::here("04_Figures", "F02", "a_script", "panels", paste0("panel_", name, ".R")))
}
```

## Panel A — global proteome state (PERMANOVA)

PERMANOVA asks whether two groups' clouds sit in different places in proteome
space, comparing the distance between group centroids to a null built by shuffling
the labels. Two cautions drive the design: it can be fooled by a difference in
*spread* rather than location (checked by PERMDISP in the supplement), and repeated
measures must be respected. So timepoint permutes *within* subject, and the group
test collapses each subject to their mean and treats subjects as the unit, so the
partial subjects cannot act as unequal blocks.

```{r panel-a}
panel("A")
pA
```

Groups overlap (p = 0.62); timepoint does separate (p = 0.02) — everyone moves
with training, the responder split just isn't a global axis.

## Panel B — response magnitude (RRPP)

RRPP (randomised residual permutation) is the valid permutation scheme when
proteins far outnumber subjects. Here it tests one number per subject: the length
of their before→after change vector, ‖Δ‖. If HR truly "respond more," their ‖Δ‖
should be larger.

```{r panel-b}
panel("B")
pB
```

It isn't (training p = 0.38, acute p = 0.92). HR and LR move about equally far.

## Panel C — protein-specific divergence leads

The correct test of "does this protein change differently in HR vs LR" is the
Group × Timepoint interaction, not the overlap of two separate hit lists (Gelman &
Stern 2006). The panel plots each protein's HR change against its LR change;
off-diagonal points are candidates flagged by the interaction pi-gate.

The pi-gate is `p^|log2FC|`, a fold-change-weighted p-value (Xiao 2014, Eq. 2). It
controls no error rate. Under BH it is unforgiving:

```{r pi-vs-fdr}
int_counts <- function(ph) {
  p <- dep_df[[paste0("P.Value_", ph, "_Interaction")]]
  q <- dep_df[[paste0("adj.P.Val_", ph, "_Interaction")]]
  pi <- dep_df[[paste0("sig_pi_", ph, "_Interaction")]]
  data.frame(
    phase = ph,
    raw_p_lt_05 = sum(p < 0.05, na.rm = TRUE),
    pi_gated = sum(pi != 0, na.rm = TRUE),
    bh_lt_05 = sum(q < 0.05, na.rm = TRUE),
    min_q = round(min(q, na.rm = TRUE), 3)
  )
}
knitr::kable(rbind(int_counts("Training"), int_counts("Acute")))
```

The pi-gate is no stricter than raw p (55 vs 51, 86 vs 114), and **zero proteins
survive FDR** (minimum q ≈ 0.76–0.89). So the divergent points are exploratory
leads, not discoveries; the panel prints both counts so the distinction is visible.
An error-rate-controlled alternative with an effect-size floor is `limma::treat()`.

```{r panel-c}
panel("C")
pC
```

## Panels D–F — the proteome overview

Descriptive scale for the figure. **D** counts DEPs per contrast as a percent of
the proteome against a chance line; **E** shows each contrast's log2FC density; **F**
runs preranked GO GSEA per contrast, collapsing redundant sets, and is the only
FDR-controlled panel. Read F's asymmetry (HR 85 vs LR 10 GO sets in the acute bout)
as a per-contrast count, not a between-group difference test.

```{r panels-def}
panel("D")
panel("E")
panel("F")
pE
pD
pF
```

## Supplements — the rest of the null argument

Two supplements corroborate "no global divergence" with subjects as the unit. The
**dispersion** supplement uses PERMDISP (distance to the group centroid per
timepoint) to check responders aren't simply more heterogeneous — they aren't
(p = 0.33 / 0.20 / 0.35). The **direction** supplement runs RRPP on the full
per-subject change matrix and the angle between the groups' mean change vectors —
no directional difference either (p = 0.87 / 0.18).

```{r supplements}
source(here::here("04_Figures", "F02", "a_script", "supp", "panel_C.R"))
source(here::here("04_Figures", "F02", "a_script", "supp", "panel_direction.R"))
```

The ordination and magnitude panels run on the missForest-imputed matrix so the
sample sets align; imputation borrows across proteins, which shrinks per-protein
variance and can compress ‖Δ‖. The DEP-derived panels (C–F) use the non-imputed
limma arm.

## Key terms

**PERMANOVA** — a permutation test for a difference in multivariate location.
Sensitive to dispersion, so it is read alongside PERMDISP.

**PERMDISP** — tests whether one group is more spread out than the other
(dispersion), the confound PERMANOVA cannot separate on its own.

**RRPP** — permutation of model residuals, the valid scheme when proteins far
outnumber subjects. Used here on one change vector per subject, so nothing is
pseudoreplicated.

**Interaction (Group × Timepoint)** — the correct test of "changes differently in
HR vs LR." Comparing two separate hit lists is not that test.

**pi-gate vs FDR** — the pi-value `p^|log2FC|` ranks by fold-change-weighted
significance but controls no error rate; BH-FDR is the honest bar, and here it
leaves nothing, so the divergent proteins are leads.

## Recap

- Shared global response: PCA/PERMANOVA, RRPP magnitude and direction, and PERMDISP
  all null with subjects as the unit.
- Protein-specific divergence is a pi-gated lead list; no protein survives FDR, so
  the panel reports both counts.
- The one FDR-clean positive is pathway-level (Panel F), and even that is a
  per-contrast count, not a difference test.

```{r sessioninfo}
sessionInfo()
```
