---
title: "Predicting responder class from the muscle proteome"
subtitle: "Nested leave-one-subject-out, honestly scored against a permutation null"
author: "HRvLR proteomics"
format:
  html:
    self-contained: true
    toc: true
---

# The question

Can the muscle proteome tell a high responder (HR) from a low responder
(LR) *out of sample*? We ask it three ways, one per phase:

- **baseline** — the T1 proteome, before any training.
- **training** — the T1 to T2 change, the adaptation to the programme.
- **acute** — the T2 to T3 change, the response to a single bout.

Within each phase we compare three feature spaces — singscore pathway
scores (primary), individual proteins, and the F06 module eigengenes —
and two models, an elastic net (`glmnet`) and sparse PLS-DA (`mixOmics`).

Sixteen subjects is a small cohort, so a single accuracy number is close
to meaningless. Everything below is cross-validated and every number
carries a permutation p. A null result here is a real result: it is the
honest answer to "is there anything an out-of-sample model can grab?"

# Why this is leakage-free

The danger with a cohort this size is that a preprocessing step quietly
peeks at the held-out subject and hands the model an unfair hint. Two
choices rule that out.

**singscore is single-sample.** Each sample's pathway score is built from
that sample's own within-sample gene ranks. It never looks at the rest of
the cohort, so scoring all 45 samples at once cannot leak the held-out
subject into the training features. That is the whole reason we score
with singscore rather than a cohort-relative method such as GSVA, whose
scores shift when you add or remove a sample.

**Everything else is fit train-only, inside the fold.** The outer loop
leaves one subject out. On the remaining subjects we z-score the features
(centre and scale learned on the training subjects, then applied to the
held-out subject), and we tune the model — `glmnet`'s alpha and lambda,
sPLS's keepX — with a second, inner leave-one-subject-out loop that never
sees the held-out subject. Only then is the held-out subject predicted.
The permutation null re-runs that entire nested procedure on shuffled
labels, 200 times, so the null carries the same optimism the real fit
does.

# The results

```{r}
#| message: false
pacman::p_load(here, dplyr, readr, purrr, knitr)

phases <- c("baseline", "training", "acute")
metrics <- map_dfr(phases, function(ph) {
  read_csv(
    here(
      "05_test", "prediction_responder_class", ph, "c_data",
      "class_metrics.csv"
    ),
    show_col_types = FALSE
  )
})

metrics |>
  transmute(phase, space, model, n,
    AUC = round(estimate, 2),
    CI = sprintf("[%.2f, %.2f]", ci_lo, ci_hi),
    perm_p = round(perm_p, 3),
    BH_q = round(q_bh, 3)
  ) |>
  arrange(perm_p) |>
  kable()
```

A few things to read off the table.

**The effective sample size is small and phase-dependent.** Baseline uses
15 subjects (S29 has no T1), training 14 (S28 and S29 each miss an end of
the T1-to-T2 window), acute 15 (S28 has no post samples). Leave-one-out on
14-15 subjects is about as thin as cross-validation gets.

**Point AUCs wander below 0.5, and that is expected.** With leave-one-out
on an unbalanced cohort, removing an HR subject shifts the training
majority toward LR, which nudges the held-out prediction the wrong way.
An intercept-only model can therefore land at an AUC well under chance.
This is exactly why the reference is the permutation null and not the
0.5 line: the same bias sits in every permutation, so the p-value stays
honest.

**Nothing survives multiple testing.** The strongest single cell is the
baseline singscore sPLS-DA classifier, and even it does not clear a
nominal permutation p, let alone BH correction across the six cells in
its leaf. Read together with the near-chance protein and eigengene
models, the message is consistent: the baseline proteome does not
separate future responders out of sample, and the training and acute
windows do not rescue it.

# Recap

- Prediction was scored with nested leave-one-subject-out and a
  200-permutation null, because point estimates on n = 16 cannot be
  trusted on their own.
- singscore was chosen precisely because it is single-sample and rank
  based, so full-cohort scoring introduces no leakage; every fold-specific
  transform stayed train-only.
- Across three phases, three feature spaces, and two models, no classifier
  separated HR from LR beyond its permutation null after BH correction.
  At this sample size that null is the finding, not a failure.

```{r}
sessionInfo()
```
