---
title: "Stage 3: Differential Expression"
subtitle: "limma + duplicateCorrelation -> 7 Contrasts -> Pi-Score -> Robustness"
author: "DTL"
date: today
bibliography: references.bib
format:
  html:
    toc: true
    toc-depth: 4
    toc-location: left
    toc_float: true
    number-sections: true
    code-fold: show
    code-summary: "Show code"
    code-tools: true
    theme:
      light: flatly
      dark: darkly
    fig-width: 10
    fig-height: 7
    df-print: paged
    smooth-scroll: true
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
---

## Setup

```{r}
#| label: setup
#| include: false
#| eval: true

knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)
```

```{r}
#| label: libraries
#| eval: true

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(readr)
  library(readxl)
  library(ggplot2)
})
```

## Introduction

Welcome to Stage 3 of the HRvLR proteomics analysis pipeline --- the **differential expression** stage. This is where we ask the central question of the study: **which proteins change in abundance between conditions, and do High Responders (HR) and Low Responders (LR) differ in their proteomic response to exercise?**

### What is differential expression analysis?

**Differential expression analysis** (or more precisely, **differential abundance analysis** for proteomics) is a statistical procedure that tests whether a protein's abundance differs significantly between two conditions. For each protein, we compute a **log-fold change (logFC)** --- how much the protein's abundance changes on the log2 scale --- and a **p-value** that quantifies the probability of observing that change by chance alone. A **log2-fold change** of 1 means the protein doubled in abundance; a logFC of -1 means it halved.

### Why limma?

We use **limma** (Linear Models for Microarray Data) [@ritchie2015], a statistical framework originally developed for microarray gene expression but widely adopted in proteomics. Limma's key innovation is **empirical Bayes moderation**: for each protein, the per-protein variance estimate is stabilized by borrowing information from the distribution of all protein variances in the experiment. This is especially important for our study, where we have only n = 8 subjects per group --- too few to get stable variance estimates for individual proteins.

Peng et al. (2024) demonstrated that limma applied directly to data with missing values --- where per-protein models are fitted using only the complete observations for that protein --- consistently outperforms alternatives that impute missing values before modeling [@peng2024]. Imputation introduces distributional assumptions that can distort variance estimates and inflate precision. Accordingly, the **non-imputed** matrix from Stage 1 is our primary input.

### Pipeline overview

The pipeline is implemented across three scripts:

| Script | Role |
|--------|------|
| `01_run_dep.R` | Core computation: model fitting, contrast testing, Pi-score, results export |
| `02_dep_reports.R` | Visualization: per-contrast volcano plots, overview bar charts, outlier sensitivity |
| `03_dep_robustness.R` | Robustness: response profiling, bootstrap CIs, power analysis, imputation sensitivity, supplementary workbook |

## Prerequisites

### Required R packages

- **proteoDA** --- wraps limma with a proteomics-friendly interface; handles `DAList` objects, `duplicateCorrelation`, and report generation [@thurman_proteoDA_2023]
- **limma** --- the statistical engine for linear modeling and empirical Bayes moderation [@ritchie2015]
- **boot** --- for bootstrap confidence intervals
- **pwr** --- for power analysis calculations
- **openxlsx** --- for Excel workbook assembly

### Input files

| File | Location | Description |
|------|----------|-------------|
| `02_normalized.csv` | `01_normalization/c_data/` | Cycloess-normalized, non-imputed protein matrix |
| `03_DAList_normalized.rds` | `01_normalization/c_data/` | Full DAList with canonical metadata |
| `01_imputed.csv` | `02_Imputation/c_data/` | Imputed matrix (for sensitivity analysis only) |

### Expected directory structure

```
A_HRvLR_2026/
  01_normalization/c_data/   <- Stage 1 outputs
  02_Imputation/c_data/      <- Stage 2 outputs (sensitivity analysis)
  03_DEP/
    a_script/                <- this file lives here
    b_reports/               <- PDF reports + proteoDA outputs
    c_data/                  <- results tables and workbooks
```

## Loading Data and Building Metadata

We read the normalized abundance matrix from Stage 1, then load the canonical metadata object --- the `DAList` produced by `proteoDA` during normalization --- rather than parsing sample names by regular expression. This ensures that **factor levels** for `Group`, `Timepoint`, and `Group_Time` are identical to those used in all other stages.

> **Note:** Deriving metadata from column names (e.g., parsing "S03_T1" to extract subject and timepoint) would introduce fragility: any renaming or reordering of samples would silently produce incorrect factor assignments. Loading the canonical metadata object removes that risk entirely.

```{r}
#| label: dep-load
#| eval: false

library(proteoDA)

df <- read_csv("01_normalization/c_data/02_normalized.csv", show_col_types = FALSE)

ann_cols   <- c("uniprot_id", "protein", "gene", "description")
ann        <- df[, ann_cols]
samp_names <- setdiff(names(df), ann_cols)
mat        <- as.matrix(df[, samp_names])
rownames(mat) <- ann$uniprot_id

cat(sprintf("Loaded: %d proteins x %d samples | missing: %d (%.1f%%)\n",
            nrow(mat), ncol(mat), sum(is.na(mat)),
            100 * sum(is.na(mat)) / length(mat)))

# Canonical metadata from the normalization DAList (not regex-derived)
dal_norm <- readRDS("01_normalization/c_data/03_DAList_normalized.rds")
dal_meta <- as.data.frame(dal_norm$metadata)
meta <- tibble(
  sample_id = dal_meta$Col_ID,
  responder = factor(dal_meta$Group,      levels = c("HR", "LR")),
  time      = factor(dal_meta$Timepoint,  levels = c("T1", "T2", "T3")),
  group     = factor(dal_meta$Group_Time,
                     levels = c("HR_T1", "HR_T2", "HR_T3",
                                "LR_T1", "LR_T2", "LR_T3")),
  subject   = sub("_T[123]$", "", dal_meta$Col_ID)
)

print(table(meta$responder, meta$time))
stopifnot(setequal(colnames(mat), meta$sample_id))
```

### Sample composition

```{r}
#| label: show-metadata
#| eval: true

dal_norm <- readRDS("01_normalization/c_data/03_DAList_normalized.rds")
table(dal_norm$metadata$Group, dal_norm$metadata$Timepoint)
cat(sprintf("Proteins: %d | Samples: %d\n",
            nrow(dal_norm$data), ncol(dal_norm$data)))
```

**Interpretation:** Each cell in the table should contain exactly n = 8 samples. Any cell with fewer than 8 indicates that an outlier was removed during Stage 1. The `stopifnot` guard ensures the sample order in the matrix matches the metadata exactly before any model is fitted.

With the metadata confirmed, we now specify the statistical design that governs how those samples are modeled.

## Statistical Design

### Cell-means parametrization

We parametrize the model using a **cell-means formulation** (`~ 0 + group`) rather than the default reference-cell parametrization (`~ group`). In a reference-cell model, the intercept represents one particular group mean, and all other coefficients are differences from that reference --- making contrasts harder to specify and audit. The cell-means parametrization avoids an intercept entirely: each coefficient is the estimated mean log-abundance for one of the six Group_Time combinations (HR_T1, HR_T2, HR_T3, LR_T1, LR_T2, LR_T3). Every planned contrast is then a simple arithmetic expression over those six means.

### Within-subject correlation (duplicateCorrelation)

Our study has a **repeated-measures structure**: each subject contributes three observations (T1, T2, T3), one per timepoint. Ignoring this structure would treat all observations as independent, overestimating the available degrees of freedom and producing **anti-conservative p-values** (too many false positives).

The **`duplicateCorrelation`** approach from @smyth2005 estimates a single average within-subject correlation, **rho**, by pooling information across all proteins. This pooled estimate is used as a fixed blocking correlation when fitting the per-protein linear models. Because rho is estimated globally rather than per-protein, its estimation is stable even with small n. The combination of a global blocking correlation with per-protein `eBayes` moderation is substantially more powerful than either mixed models or naive repeated-measures ANOVA in this setting.

```{r}
#| label: dep-design
#| eval: false

dal <- add_design(dal, "~ 0 + group + (1 | subject)")

# Strip the "group" prefix that model.matrix() prepends to factor-level column names
colnames(dal$design$design_matrix) <- gsub("^group", "",
                                            colnames(dal$design$design_matrix))
```

> **`add_design()`** is a `proteoDA` function that builds the design matrix and specifies the blocking factor in one call. The `(1 | subject)` term instructs `duplicateCorrelation` to treat subject as the blocking variable. The resulting design matrix has one row per sample and one column per Group_Time level.

### Estimated within-subject correlation

```{r}
#| label: show-within-cor
#| eval: true

dal_dep <- readRDS("03_DEP/c_data/01_limma_DAList.rds")
within_cor <- dal_dep$eBayes_fit$correlation %||%
  dal_dep$tags$duplicate_correlation %||% NA_real_
if (!is.na(within_cor)) {
  cat(sprintf("Within-subject correlation (rho): %.3f\n", within_cor))
}
```

**Interpretation:** A rho above 0.5 indicates that subjects are highly consistent in their proteomic profiles across timepoints --- repeated measures from the same individual are strongly correlated, and accounting for this structure provides substantial power gains. A rho below 0.3 would suggest high inter-individual variability relative to the within-subject signal. A rho near zero would imply the repeated-measures structure provides negligible additional information.

## Contrasts

### The seven planned comparisons

Seven contrasts were pre-specified before model fitting to address the principal biological questions of the study:

| Contrast | Formula | Biological question |
|:---------|:--------|:-------------------|
| **Training_HR** | HR_T2 - HR_T1 | Do HR subjects change after training? |
| **Training_LR** | LR_T2 - LR_T1 | Do LR subjects change after training? |
| **Acute_HR** | HR_T3 - HR_T2 | Do HR subjects respond to the acute bout? |
| **Acute_LR** | LR_T3 - LR_T2 | Do LR subjects respond to the acute bout? |
| **Baseline_HRvLR** | HR_T1 - LR_T1 | Do HR and LR differ at baseline? |
| **Training_Interaction** | (HR_T2 - HR_T1) - (LR_T2 - LR_T1) | Does the training response differ between HR and LR? |
| **Acute_Interaction** | (HR_T3 - HR_T2) - (LR_T3 - LR_T2) | Does the acute response differ between HR and LR? |

The two **interaction contrasts** are the analytically most important: they ask whether the proteomic response to exercise differs between responder groups, which is the central question of the study.

```{r}
#| label: dep-contrasts
#| eval: false

dal <- add_contrasts(dal, contrasts_vector = c(
  "Training_HR          = HR_T2 - HR_T1",
  "Training_LR          = LR_T2 - LR_T1",
  "Acute_HR             = HR_T3 - HR_T2",
  "Acute_LR             = LR_T3 - LR_T2",
  "Baseline_HRvLR       = HR_T1 - LR_T1",
  "Training_Interaction = (HR_T2 - HR_T1) - (LR_T2 - LR_T1)",
  "Acute_Interaction    = (HR_T3 - HR_T2) - (LR_T3 - LR_T2)"
))
```

### Contrast matrix

The resulting contrast matrix shows how each Group_Time cell mean contributes to each comparison. A value of +1 means that cell mean is added, -1 means subtracted, and 0 means it does not contribute:

| Group_Time | Training_HR | Training_LR | Acute_HR | Acute_LR | Baseline_HRvLR | Training_Int | Acute_Int |
|:-----------|:-----------:|:-----------:|:--------:|:--------:|:--------------:|:------------:|:---------:|
| HR_T1      |     -1      |      0      |    0     |    0     |       1        |      -1      |     0     |
| HR_T2      |      1      |      0      |   -1     |    0     |       0        |       1      |     0     |
| HR_T3      |      0      |      0      |    1     |    0     |       0        |       0      |     1     |
| LR_T1      |      0      |     -1      |    0     |    0     |      -1        |       1      |     0     |
| LR_T2      |      0      |      1      |    0     |   -1     |       0        |      -1      |     0     |
| LR_T3      |      0      |      0      |    0     |    1     |       0        |       0      |    -1     |

For the interaction contrasts, reading down the column reveals the **difference-of-differences** structure: the Training_Interaction column has -1 at HR_T1, +1 at HR_T2, +1 at LR_T1, and -1 at LR_T2, which implements `(HR_T2 - HR_T1) - (LR_T2 - LR_T1)` exactly.

## Fitting the Model and Extracting Results

### How limma works (step by step)

Model fitting proceeds in three internally chained steps within `fit_limma_model()`:

1. **`lmFit()`** --- fits a per-protein linear model using the design matrix and blocking correlation rho from `duplicateCorrelation`
2. **`contrasts.fit()`** --- reparametrizes from cell-means space into contrast space, producing contrast-specific coefficient estimates and standard errors
3. **`eBayes()`** --- applies **empirical Bayes moderation**: each protein's variance estimate is shrunk toward a prior variance derived from the global distribution of all protein variances [@ritchie2015]. This stabilizes proteins quantified from few observations.

### FDR threshold

The **False Discovery Rate (FDR)** threshold for the primary significance call is set at **0.10**. The **FDR** is the expected proportion of false positives among all proteins called significant [@benjamini1995]. An FDR of 0.10 means we accept that up to 10% of our "significant" proteins may be false discoveries.

This threshold is appropriate for discovery proteomics with n = 8 per group, where strict FDR control at 0.05 would discard genuinely altered proteins with modest statistical support. The **Pi-score** (see next section) provides a secondary filter that additionally requires a biologically meaningful fold change.

```{r}
#| label: dep-fit
#| eval: false

dal <- fit_limma_model(dal)
dal <- extract_DA_results(dal, pval_thresh = 0.10, lfc_thresh = 0, adj_method = "BH")

saveRDS(dal, file.path(DATA_DIR, "limma_DAList.rds"))
```

> **`extract_DA_results()`** applies the Benjamini-Hochberg procedure to adjust p-values for multiple testing. The `pval_thresh` argument sets the FDR threshold for calling proteins as differentially abundant. `lfc_thresh = 0` means we do not impose a minimum fold-change requirement at this stage (the Pi-score handles that).

### Significance summary

```{r}
#| label: show-da-summary
#| eval: true

da_sum <- read_csv("03_DEP/c_data/02_DA_summary.csv", show_col_types = FALSE)
da_sum |>
  filter(type != "nonsig") |>
  select(contrast, type, any_of(c("sig.Pi", "sig.FDR.05", "sig.FDR.10")))
```

**Interpretation:** Where FDR and Pi counts are zero, this does not necessarily indicate biological null results. In a small-n DIA-MS study, the analysis is powered to detect proteins with large fold changes. The interaction contrasts in particular test a subtle hypothesis (difference in response magnitude) that requires substantial biological effect to clear a stringent multiple-testing threshold. The bootstrap confidence intervals later provide an effect-size summary that is interpretable regardless of significance.

## Pi-Score

### What is the Pi-score?

The **Pi-score**, introduced by @xiao2014, integrates statistical significance with fold-change magnitude into a single scalar:

$$\text{Pi} = p^{|\log_2 \text{FC}|}$$

The logic: the exponent (absolute logFC) modulates how much a given p-value is penalized:

- **Small fold change** (|logFC| < 1): the exponent is less than 1, so Pi is *larger* than the raw p-value --- harder to reach significance
- **Large fold change** (|logFC| > 1): the exponent exceeds 1, so Pi *shrinks* below the raw p-value --- easier to reach significance

Neither criterion alone is sufficient. A protein must show **both** meaningful abundance change **and** robust statistical support to be classified as significant by Pi < 0.05.

### Concrete examples

| p-value | |logFC| | Pi = p^|logFC| | Significant? |
|---------|--------|----------------|-------------|
| 0.01 | 0.5 | 0.01^0.5 = 0.100 | No (Pi > 0.05) |
| 0.01 | 1.0 | 0.01^1.0 = 0.010 | Yes |
| 0.01 | 1.5 | 0.01^1.5 = 0.001 | Yes |
| 0.05 | 0.5 | 0.05^0.5 = 0.224 | No |
| 0.05 | 2.0 | 0.05^2.0 = 0.003 | Yes |

```{r}
#| label: pi-score-computation
#| eval: false

# Applied to each per-contrast CSV after writing by write_limma_tables()
res <- res |>
  mutate(
    pi_score = P.Value ^ abs(logFC),
    sig_pi = case_when(
      pi_score < 0.05 & logFC > 0 ~  1L,   # significant up-regulation
      pi_score < 0.05 & logFC < 0 ~ -1L,    # significant down-regulation
      TRUE ~ 0L                               # not significant
    )
  )
```

### Top proteins by Pi-score (Training_HR)

```{r}
#| label: show-per-contrast
#| eval: true

f <- "03_DEP/c_data/04_per_contrast_results/Training_HR.csv"
if (file.exists(f)) {
  res <- read_csv(f, show_col_types = FALSE)
  cat(sprintf("Training_HR: %d proteins\n", nrow(res)))
  res |>
    arrange(pi_score) |>
    select(gene, logFC, P.Value, adj.P.Val, pi_score, sig_pi) |>
    head(5)
}
```

**Interpretation:** A protein ranking in the top 5 by Pi but not reaching FDR < 0.10 has a large fold change that pushes its Pi below 0.05, but its nominal p-value is not small enough to survive Benjamini-Hochberg correction. Such a protein is biologically interesting but statistically marginal --- it warrants attention in pathway context and sensitivity analysis, but should not be reported as a definitive DEP without corroborating evidence.

## Combined Results Matrix

Results across all seven contrasts are assembled into a single wide-format matrix with one row per protein and contrast-suffixed columns (e.g., `logFC_Training_HR`, `adj.P.Val_Training_HR`, `pi_score_Training_HR`). This structure is convenient for downstream analysis: any per-protein, cross-contrast comparison can be performed with a single `select()` or `filter()` call.

```{r}
#| label: show-combined-results
#| eval: true

if (file.exists("03_DEP/c_data/03_combined_results.csv")) {
  comb <- read_csv("03_DEP/c_data/03_combined_results.csv", show_col_types = FALSE)
  cat(sprintf("Combined results: %d proteins x %d columns\n", nrow(comb), ncol(comb)))
  comb |>
    select(uniprot_id, gene,
           logFC_Training_HR, P.Value_Training_HR,
           adj.P.Val_Training_HR, pi_score_Training_HR, sig_pi_Training_HR) |>
    head(4)
}
```

With the core results assembled, the pipeline generates visualizations for each contrast and performs robustness analyses. We describe these next.

## Per-Contrast Visualization

For each of the seven contrasts, the pipeline produces a **triple volcano plot** --- three panels representing progressively more stringent significance filters:

1. **Left panel**: -log10(nominal P.Value) vs logFC, threshold at P < 0.01
2. **Middle panel**: -log10(adj.P.Val) vs logFC, threshold at FDR < 0.10
3. **Right panel**: -log10(Pi) vs logFC, threshold at Pi < 0.05

Reading from left to right moves from the broadest candidate set to the most stringent joint filter. Upregulated proteins are red, downregulated are blue, non-significant are grey.

```{r}
#| label: volcano-code
#| eval: false

# Volcano factory --- defined once, called three times per contrast
make_vol <- function(df, ycol, ylab, top_df, thresh) {
  ggplot(df, aes(logFC, .data[[ycol]], color = dir_pi)) +
    geom_point(alpha = 0.35, size = 1) +
    geom_hline(yintercept = thresh, linetype = "dashed",
               color = "grey40", linewidth = 0.4) +
    geom_text_repel(data = top_df, aes(label = gene), size = 2.5,
                    max.overlaps = 15, show.legend = FALSE, seed = 42) +
    scale_color_manual(values = c(Up = "#D6604D", Down = "#4393C3", NS = "grey70"),
                       drop = FALSE) +
    labs(x = expression(log[2]~FC), y = ylab) +
    theme_bw(base_size = 11) +
    theme(legend.position = "none")
}

# For each contrast (shown here for Training_HR):
p1 <- make_vol(res, "nlog10_pval", expression(-log[10](P.Value)),
               top_nom, -log10(0.01)) + ggtitle("Nominal P < 0.01")
p2 <- make_vol(res, "nlog10_adj",  expression(-log[10](adj.P.Val)),
               top_adj, -log10(0.10)) + ggtitle("FDR < 0.10")
p3 <- make_vol(res, "nlog10_pi",   expression(-log[10](Pi)),
               top_pi,  -log10(0.05)) + ggtitle("Pi < 0.05")
```

**How to read a volcano plot:** Each point is a protein. The x-axis is the logFC (positive = up in the numerator condition, negative = down). The y-axis is statistical significance on a -log10 scale (higher = more significant). Proteins in the upper corners are the strongest candidates --- they have both large fold changes and small p-values.

## Response Profiling

### Group-level proteomic plasticity

Response profiling asks a different question from per-protein DEP tests: not which *specific* proteins change, but whether the **overall magnitude** of proteomic response differs between HR and LR. A broader distribution of |logFC| values in HR relative to LR across all proteins --- including those that individually do not reach significance --- would suggest that HR subjects mount a globally more pronounced proteomic response.

Four complementary statistical tests are applied:

- **Kolmogorov-Smirnov test**: tests whether the shapes of the |logFC| distributions differ (sensitive to location, spread, and skew)
- **Fligner-Killeen test**: specifically tests whether the *variance* of |logFC| differs between groups
- **Wilcoxon signed-rank test**: treats per-protein |logFC| values as paired and tests for a systematic shift
- **Cliff's delta**: a non-parametric effect size ranging from -1 to +1; positive means HR has larger fold changes. Effect sizes: negligible (< 0.147), small (0.147--0.33), medium (0.33--0.474), large (> 0.474) [@romano2006]

```{r}
#| label: response-profiling-code
#| eval: false

# run_response_comparison() is defined in 03_dep_robustness.R
# Two calls --- one for Training, one for Acute:
rp_training <- run_response_comparison(results_list,
  "Training_HR", "Training_LR", "Training (HR vs LR)")
rp_acute <- run_response_comparison(results_list,
  "Acute_HR", "Acute_LR", "Acute (HR vs LR)")
```

### Response profiling results

```{r}
#| label: show-response-profiling
#| eval: true

if (file.exists("03_DEP/c_data/06_response_profiling_diagnostics.csv")) {
  rp <- read_csv("03_DEP/c_data/06_response_profiling_diagnostics.csv",
                 show_col_types = FALSE)
  rp |> select(comparison, test, statistic, p_value, interpretation)
}
```

**Interpretation:** A significant KS test (p < 0.05) means the shape of the |logFC| distribution differs between HR and LR. A positive Cliff's delta with medium or large magnitude provides the clearest evidence that HR subjects show globally larger proteomic responses. If all four tests are non-significant with negligible Cliff's delta, the data do not support a global difference in response magnitude --- any observed DEP differences should be interpreted as protein-specific rather than reflecting general proteomic plasticity.

## Bootstrap Confidence Intervals

### Effect-size summary per contrast

The per-protein significance tests report which proteins cross a threshold but give little intuition about the **overall scale** of proteomic change. The bootstrap analysis computes the **median absolute log-fold change** across all quantified proteins per contrast, together with a 95% **bias-corrected and accelerated (BCa)** confidence interval from 10,000 bootstrap resamples [@efron1993].

The median |logFC| answers: "how large are the protein abundance changes in this contrast, on average?" The BCa method corrects for both bias and skewness in the bootstrap distribution, producing more accurate interval coverage with heavy-tailed data.

```{r}
#| label: bootstrap-code
#| eval: false

library(boot)

# Applied per contrast (shown for a single contrast named cname):
vals <- abs(results_list[[cname]]$logFC)
vals <- vals[!is.na(vals)]
b    <- boot(vals, function(d, i) median(d[i], na.rm = TRUE), R = 10000)
ci   <- tryCatch(boot.ci(b, type = "bca"),
                 error = function(e) boot.ci(b, type = "perc"))
```

### Bootstrap results

```{r}
#| label: show-bootstrap
#| eval: true

if (file.exists("03_DEP/c_data/09_effect_size_bootstrap.csv")) {
  boot_df <- read_csv("03_DEP/c_data/09_effect_size_bootstrap.csv",
                      show_col_types = FALSE)
  boot_df |> arrange(desc(median_absLFC))
}
```

**Interpretation:** The ordering of contrasts by descending median |logFC| reveals which conditions drive the largest proteomic changes. Interaction contrasts typically have smaller median |logFC| than within-group contrasts (they measure differences in responses rather than responses themselves). Narrow confidence intervals reflect high precision; wide intervals arise from highly variable logFC distributions or many missing values.

## Power Analysis

### What the study can detect

The **power analysis** provides a lower bound on the **minimum detectable log-fold change** for each contrast at 80% power and alpha = 0.10 [@cohen1988]. This uses `pwr.t.test()`, which implements the standard t-test power formula. This is a **conservative** bound because limma's moderated t-statistic has greater power than the ordinary t-statistic --- empirical Bayes shrinkage effectively increases the degrees of freedom.

For **within-group contrasts** (Training_HR, Training_LR, Acute_HR, Acute_LR), the paired test formula is used and the effective residual standard deviation is adjusted for within-subject correlation: sigma_eff = sigma x sqrt(2(1 - rho)). A high rho substantially reduces sigma_eff and thus the minimum detectable logFC.

```{r}
#| label: power-code
#| eval: false

library(pwr)

# Applied per contrast (within sigma estimated from eBayes fit):
eff_sigma <- if (paired && !is.na(within_cor))
  sigma_residual * sqrt(2 * (1 - within_cor)) else
  sigma_residual * sqrt(2)

pw <- pwr.t.test(n = n_subj, d = NULL, sig.level = 0.10, power = 0.80,
                 type = if (paired) "paired" else "two.sample")
min_detectable_logFC <- pw$d * eff_sigma
```

### Power analysis results

```{r}
#| label: show-power
#| eval: true

if (file.exists("03_DEP/c_data/10_power_analysis.csv")) {
  power_df <- read_csv("03_DEP/c_data/10_power_analysis.csv",
                       show_col_types = FALSE)
  power_df |>
    select(contrast, n_subjects, within_cor, effective_sigma,
           min_detectable_d, min_detectable_logFC)
}
```

**Interpretation:** If `min_detectable_logFC` exceeds 0.5 log2 units (approximately a 1.4-fold change), the study can reliably detect only proteins with substantial abundance shifts. This is common in small-n DIA-MS experiments and does not invalidate the analysis --- it means that the detected DEPs, if any, are robust to noise. Paired within-group contrasts benefit most from high rho: at rho = 0.7, the effective sigma is reduced by ~23% relative to the independent-samples assumption.

## Imputation Sensitivity Analysis

### Confirming robustness to missing-value strategy

As a robustness check, an identical limma pipeline is run on the **imputed** version of the matrix (from Stage 2), and the resulting t-statistics are compared per contrast by **Spearman rank correlation**. High concordance (rho > 0.95) confirms that the choice of imputation strategy does not materially alter the protein rankings [@peng2024].

```{r}
#| label: show-imputation-sensitivity
#| eval: true

if (file.exists("03_DEP/c_data/11_imputation_sensitivity.csv")) {
  sens <- read_csv("03_DEP/c_data/11_imputation_sensitivity.csv",
                   show_col_types = FALSE)
  sens
} else {
  cat("Imputation sensitivity file not yet available.\n")
}
```

**Interpretation:**

- **Spearman rho > 0.95**: Excellent concordance. The two approaches rank proteins almost identically; any significance differences are attributable to random variation.
- **Spearman rho 0.85--0.95**: Moderate sensitivity. Proteins that switch significance status deserve closer examination.
- **Spearman rho < 0.85**: Substantial reshaping. The non-imputed results are still reported as primary (per @peng2024), but the discordance should be documented.

## Stage Outputs

All outputs from the DEP pipeline are written to `03_DEP/c_data/` and numbered in pipeline order:

| File | Description |
|:-----|:------------|
| `c_data/01_limma_DAList.rds` | Full `proteoDA` DAList with fitted eBayes model inside |
| `c_data/02_DA_summary.csv` | Significance counts by contrast, direction, and criterion (FDR 0.05, FDR 0.10, Pi < 0.05) |
| `c_data/03_combined_results.csv` | **PRIMARY OUTPUT** --- all 7 contrasts in wide format, one row per protein |
| `c_data/04_per_contrast_results/` | 7 individual CSVs with logFC, t, P.Value, adj.P.Val, pi_score, sig_pi |
| `c_data/05_results.xlsx` | Excel workbook with one sheet per contrast plus DA_Summary |
| `c_data/06_response_profiling_diagnostics.csv` | KS, Fligner-Killeen, Wilcoxon, Cliff's delta for Training and Acute comparisons |
| `c_data/07_response_profiling_training_quantiles.csv` | Quantile comparison of |logFC| for Training_HR vs Training_LR |
| `c_data/08_response_profiling_acute_quantiles.csv` | Quantile comparison of |logFC| for Acute_HR vs Acute_LR |
| `c_data/09_effect_size_bootstrap.csv` | Median |logFC| with 95% BCa CI per contrast (10,000 resamples) |
| `c_data/10_power_analysis.csv` | Minimum detectable logFC at 80% power, alpha = 0.10, per contrast |
| `c_data/11_imputation_sensitivity.csv` | Spearman rho between non-imputed and imputed t-statistics per contrast |
| `c_data/12_DEP_supplementary.xlsx` | Supplementary workbook (13 sheets): methods overview, significance summary, per-contrast results, response profiling, bootstrap CIs, power analysis, imputation sensitivity |

This completes the core proteomics analysis pipeline. The normalized, tested, and annotated results are now ready for biological interpretation, pathway enrichment analysis, and figure generation in Stage 4.

## Session Info

```{r}
#| label: session-info
#| eval: true

sessionInfo()
```

## References

::: {#refs}
:::
