---
title: "F05 prediction — singscore pathway scores"
subtitle: "Responder class and training gains, out of sample: a clean negative"
author: "A_HRvLR_2026"
format:
  html:
    toc: true
    embed-resources: true
    df-print: paged
execute:
  eval: true
  echo: false
  warning: false
  message: false
---

## In plain terms

Can the proteome, summarised as knowledge-based pathway scores, predict a
held-out person? We train on 15 people and predict the 16th, rotating
through everyone, for two questions: which responder group they're in, and
how much they actually gained.

## Method

Each fold is a nested leave-one-subject-out: features are z-scored on the
training people only and that centre and scale are applied to the held-out
person, so nothing about the held-out subject touches the transform.
glmnet and sparse-PLS are tuned by an inner leave-one-out on the training
people alone, and the winning setting is refit on all of them before
scoring the held-out subject. A permutation null shuffles the outcome
across subjects and re-runs the entire nested procedure, giving each cell
its own p-value; Benjamini-Hochberg correction is then applied across the
whole metric grid. n = 16, exploratory.

**singscore is scored per sample**: each sample's pathway score depends
only on that sample's own within-sample gene ranks, never on any other
sample in the cohort. Unlike a cohort-relative summary such as GSVA, the
feature itself carries no trace of the held-out subject, so this family is
genuinely leakage-free — there is no upper-bound caveat to attach to it.

## Results

```{r}
#| label: load-results
pacman::p_load(here, openxlsx, dplyr)

class_metrics <- read.xlsx(
  here(
    "04_Figures", "F05_prediction", "singscore", "responder", "c_data",
    "class_source_data.xlsx"
  ),
  "metrics"
)
cont_metrics <- read.xlsx(
  here(
    "04_Figures", "F05_prediction", "singscore", "continuous", "c_data",
    "cont_source_data.xlsx"
  ),
  "metrics"
)

best_class <- class_metrics |> slice_max(estimate, n = 1, with_ties = FALSE)
best_cont <- cont_metrics |> slice_max(q2, n = 1, with_ties = FALSE)

n_class_cells <- nrow(class_metrics)
n_cont_cells <- nrow(cont_metrics)
n_class_hits <- sum(class_metrics$q_bh < 0.05, na.rm = TRUE)
n_cont_hits <- sum(cont_metrics$q_bh < 0.05, na.rm = TRUE)
```

Across the `r n_class_cells` phase-by-model cells that predict responder
class, the best is `r best_class$model` on the `r best_class$phase` phase:
AUC = `r sprintf("%.2f", best_class$estimate)` (permutation
p = `r sprintf("%.2f", best_class$perm_p)`), but after BH correction across
the grid its q = `r sprintf("%.2f", best_class$q_bh)` — it does not survive.

Across the `r n_cont_cells` outcome-by-phase-by-model cells that predict a
training gain, the best Q² anywhere is
`r sprintf("%.3f", best_cont$q2)` (`r best_cont$outcome`, `r best_cont$phase`
phase, `r best_cont$model`; permutation p = `r sprintf("%.2f", best_cont$perm_p_q2)`),
and its BH q = `r sprintf("%.2f", best_cont$q_bh)`. `r n_class_hits` of
`r n_class_cells` responder-class cells and `r n_cont_hits` of
`r n_cont_cells` continuous cells pass BH < 0.05.

**Nothing survives correction for multiple testing.** One continuous cell
clears the nominal permutation null but not its BH q, and every other cell is
null on both; BH is charged within this leaf, so pooling the two feature
families would only raise the q values further. The proteome, summarised as
per-sample pathway activity, does not predict responder class or training
gains out of sample, and this family carries no leakage to explain the
result away. This agrees with the association null reported elsewhere in
F04 and with the study-wide null: the leaky raw-protein feature space that
an earlier version of this analysis used was dropped rather than reported
alongside a result it could not honestly support.

```{r sessioninfo}
sessionInfo()
```
