---
title: "Predicting the continuous hypertrophy phenotype from the muscle proteome"
subtitle: "Leave-one-subject-out Q^2, RMSE and Spearman against a permutation null"
author: "HRvLR proteomics"
format:
  html:
    self-contained: true
    toc: true
---

# The question

The HR/LR split is a coarse cut through a continuous response. This arm
skips the cut and predicts the response itself: composite hypertrophy
(`comp_hypertrophy`) and the paired change scores for fibre and whole
muscle size and strength — `d_fcsa_I`, `d_fcsa_II`, `d_mcsa`,
`d_1rm_legpress`, `d_1rm_ext`.

As in the class arm we ask it three ways — baseline (T1), training
(T1 to T2 change), and acute (T2 to T3 change) features — and within each
phase we compare singscore pathway scores (primary), proteins, and the
F06 module eigengenes, under an elastic net and a sparse PLS regression.

# How a continuous prediction is scored honestly

For a real-valued outcome the natural cross-validated score is
**Q^2**, the out-of-sample cousin of R^2:

$$Q^2 = 1 - \frac{\sum_i (y_i - \hat y_{-i})^2}{\sum_i (y_i - \bar y)^2},$$

where $\hat y_{-i}$ is the leave-one-subject-out prediction. Q^2 = 1 is
perfect, Q^2 = 0 is no better than predicting the mean, and — unlike R^2 —
Q^2 goes **negative** when the model predicts worse than the mean, which
is the usual fate of a flexible model on a tiny noisy sample. We report
RMSE and a Spearman correlation alongside it, and give each of the three a
permutation p.

Everything is leakage-free for the same two reasons as the class arm.
singscore is single-sample and rank-based, so scoring the full cohort
once leaks nothing. The z-scoring and the hyperparameter tuning
(`glmnet`'s alpha/lambda, sPLS's keepX) are fit inside each fold on the
training subjects only, by an inner leave-one-subject-out loop, and the
200-permutation null re-runs that whole nested procedure on shuffled
outcomes. No binarization is used anywhere; the outcome stays continuous
end to end.

# The results

```{r}
#| message: false
pacman::p_load(here, dplyr, readr, purrr, knitr)

phases <- c("baseline", "training", "acute")
metrics <- map_dfr(phases, function(ph) {
  read_csv(
    here(
      "05_test", "prediction_continuous_phenotype", ph, "c_data",
      "cont_metrics.csv"
    ),
    show_col_types = FALSE
  )
})

metrics |>
  transmute(phase, outcome, space, model, n,
    Q2 = round(q2, 2),
    RMSE = round(rmse, 2),
    rho = round(spearman, 2),
    perm_p = round(perm_p_q2, 3),
    BH_q = round(q_bh, 3)
  ) |>
  arrange(perm_p) |>
  head(20) |>
  kable(caption = "Top 20 cells by Q^2 permutation p across all phases.")
```

Reading the table with the sample size in mind:

**Most Q^2 values are negative.** That is not a bug; it is what an
honestly cross-validated flexible model does on 14-15 subjects when the
signal is weak — it does worse than simply guessing the mean. The
permutation null is centred in the same negative territory, so a negative
Q^2 is only interesting relative to that null, never against zero.

**The permutation p is the arbiter, and BH keeps us honest.** With six
outcomes times three feature spaces times two models per phase, there are
many cells and therefore many chances to see a lucky Spearman. Any cell
worth a second look has to clear its permutation null and then survive BH
correction across the leaf. Cells that clear the nominal null but not BH
are hypotheses, not findings, and are labelled as such.

**Where any signal lives.** One lead recurs: change in whole-muscle CSA
(`d_mcsa`) predicted from the F06 module eigengenes under the elastic net,
in both the baseline window (Q^2 ~ 0.56, Spearman ~ 0.82, permutation
p ~ 0.015) and the acute window (Q^2 ~ 0.71, Spearman ~ 0.85, permutation
p ~ 0.010). A few strength cells (`d_1rm_ext` at baseline) also clear the
nominal null. But every one of these dies under BH correction across the
thirty-six cells in its leaf, so they are leads, not results. The training
window is null for every outcome, and composite hypertrophy — the very
phenotype that defined the HR/LR split — is null out of sample in every
phase. The one thread worth pulling is the module-eigengene to Delta-mCSA
coupling, and only as a hypothesis for a larger cohort.

# Recap

- Continuous outcomes were predicted with leave-one-subject-out Q^2, RMSE
  and Spearman, each against a 200-permutation null, with no binarization
  at any step.
- singscore's single-sample scoring plus strictly train-only z-scoring and
  inner-loop tuning keep the whole nested procedure leakage-free.
- Negative Q^2 is the expected baseline on n = 16; the permutation null,
  not zero, is the reference. Composite hypertrophy and the training window
  are null everywhere; the only recurring above-null lead is Delta-mCSA
  from the module eigengenes (baseline and acute), nominal but not BH
  significant.

```{r}
sessionInfo()
```
