---
title: "03 · Protein against phenotype"
subtitle: "Every protein correlated with the ten adaptation measures"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    df-print: kable
    fig-width: 7
    fig-height: 5
    fig-dpi: 150
    fig-align: center
execute:
  message: false
---

`02_Differential` asks which proteins differ between groups. This asks whether a protein tracks
how much a subject adapted. A contrast compares group means and carries no per-subject outcome,
so it cannot answer that.

Each subject contributes one value per protein: the level at T1, the change over training
(T2 − T1), or the change over the acute bout (T3 − T2). Each is correlated with the ten
phenotypes, all T2 − T1 changes except `volume_load`. Training change is the window that matches
the phenotypes; the baseline level asks whether a starting proteome predicts adaptation, and the
acute change whether the bout response does.

At most 1,574 proteins pass the two-thirds floor per cell, so chance gives up to 79 nominal hits
at p < 0.05. BH within each window and outcome is the reported quantity; `ratio` compares the
nominal count with chance.

```{r setup}
library(here)
library(dplyr)
library(purrr)
library(tibble)
library(ggplot2)

inputs <- c(
  proteins = "01_Preprocess/02_Normalization/c_data/DAList_normalized.rds",
  phenotype = "00_Input/phenotype.csv"
)
paths <- map_chr(inputs, here)
stopifnot(file.exists(paths))
dal <- readRDS(paths[["proteins"]])
phenotype <- readr::read_csv(paths[["phenotype"]], show_col_types = FALSE)
out <- here("02_Differential_Expression", "03_Phenotype", "c_data")
figures <- here("02_Differential_Expression", "03_Phenotype", "b_reports")
for (path in c(out, figures)) dir.create(path, recursive = TRUE, showWarnings = FALSE)
```

## Windows hold 14 or 15 subjects

A subject contributes to a window only when every timepoint it needs was acquired. HR_S29 has no
T1 and HR_S28 no T2 or T3, so the windows hold 14 or 15 subjects.

```{r windows}
meta <- as_tibble(dal$metadata)
by_subject <- function(mat, timepoint) {
  rows <- filter(meta, timepoint == .env$timepoint)
  set_names(as.data.frame(mat[, rows$sample_id, drop = FALSE]), rows$subject) |> as.matrix()
}
change <- function(mat, from, to) {
  a <- by_subject(mat, from)
  b <- by_subject(mat, to)
  both <- intersect(colnames(a), colnames(b))
  b[, both, drop = FALSE] - a[, both, drop = FALSE]
}
windows <- list(
  training = change(dal$data, "T1", "T2"),
  baseline = by_subject(dal$data, "T1"),
  acute = change(dal$data, "T2", "T3")
)
map_int(windows, ncol)
```

## Spearman p comes from the t approximation

Spearman throughout. A protein is tested in a window only when at least two thirds of that
window's subjects observed it: a protein seen in two subjects fits a rank correlation through two
points. The p comes from the t approximation, the one `cor.test(exact = FALSE)` uses, computed
for every protein at once. Above nine observations `cor.test`'s "exact" p is an Edgeworth series,
and in the far tail it returns exactly 0: RPS4X against `d_mcsa`, rho 0.91 on 14 subjects, came
back p = 0 and topped the BH list.

```{r correlate}
outcomes <- c(
  "comp_hypertrophy", "d_fcsa_I", "d_fcsa_II", "d_fcsa_mixed", "d_nfibre_mixed",
  "d_nfibre_I", "d_mcsa", "d_1rm_legpress", "d_1rm_ext", "volume_load"
)
# cor.test(exact = FALSE)'s p, step for step: S statistic, rho back from S, then the t tail.
spearman_p <- function(rho, n) {
  s <- (n^3 - n) * (1 - rho) / 6
  r <- 1 - s / ((n * (n^2 - 1)) / 6)
  t <- r / sqrt((1 - r^2) / (n - 2))
  pmin(2 * if_else(s > (n^3 - n) / 6, pt(t, n - 2), pt(t, n - 2, lower.tail = FALSE)), 1)
}
# cor() re-ranks each protein on the subjects where it was seen.
spearman_by_row <- function(values, outcome) {
  usable <- !is.na(outcome)
  values <- values[, usable, drop = FALSE]
  n <- rowSums(!is.na(values))
  rho <- suppressWarnings(
    cor(t(values), outcome[usable], method = "spearman", use = "pairwise.complete.obs")[, 1]
  )
  rho[n < ceiling(2 * sum(usable) / 3)] <- NA
  tibble(
    uniprot_id = rownames(values), n_subjects = unname(n), rho = unname(rho),
    p = spearman_p(rho, n)
  )
}
outcome_of <- function(name, subjects) {
  set_names(phenotype[[name]], phenotype$subject)[subjects]
}
results <- imap(windows, \(values, window) {
  map(set_names(outcomes), \(name) {
    spearman_by_row(values, outcome_of(name, colnames(values)))
  }) |>
    list_rbind(names_to = "outcome") |>
    mutate(window = window)
}) |>
  list_rbind() |>
  mutate(fdr = p.adjust(p, "BH"), .by = c(window, outcome)) |>
  left_join(tibble(uniprot_id = dal$annotation$uniprot_id, gene = dal$annotation$gene),
    by = "uniprot_id"
  ) |>
  relocate(window, outcome, uniprot_id, gene)
```

The same correlation inside each arm and window, descriptive only: 4 to 8 subjects per correlation.

```{r by-arm}
arm_of <- set_names(phenotype$arm, phenotype$subject)
by_arm <- map(set_names(c("HR", "LR")), \(arm) {
  imap(windows, \(values, window) {
    kept <- values[, arm_of[colnames(values)] == arm, drop = FALSE]
    map(set_names(outcomes), \(name) spearman_by_row(kept, outcome_of(name, colnames(kept)))) |>
      list_rbind(names_to = "outcome") |>
      mutate(window = window, .before = 1)
  }) |>
    list_rbind()
}) |>
  list_rbind(names_to = "arm") |>
  filter(!is.na(p))
```

## Two protein-outcome pairs survive BH

```{r chance}
chance_expectation <- results |>
  filter(!is.na(p)) |>
  summarise(
    n_tested = n(), n_nominal = sum(p < 0.05), n_fdr_05 = sum(fdr < 0.05),
    .by = c(window, outcome)
  ) |>
  transmute(
    analysis = paste0("association: ", window), comparison = outcome, n_tested, n_nominal,
    n_expected = 0.05 * n_tested, ratio = round(n_nominal / n_expected, 2), n_fdr_05
  )
chance_expectation
```

Read `ratio` first. A ratio near 1 is the null count, whatever the smallest p.

```{r figure-chance}
#| fig-width: 8
#| fig-height: 5
chance_figure <- chance_expectation |>
  mutate(
    window = factor(sub("association: ", "", analysis), c("training", "baseline", "acute")),
    outcome = factor(comparison, rev(outcomes))
  ) |>
  ggplot(aes(ratio, outcome, fill = window)) +
  geom_vline(xintercept = 1, linewidth = 0.4, colour = "grey40") +
  geom_col(position = position_dodge(width = 0.75), width = 0.7) +
  scale_fill_manual(
    values = c(training = "#B2182B", baseline = "grey55", acute = "#2166AC"),
    name = NULL
  ) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Nominal hits relative to chance, by outcome and window",
    subtitle = "Spearman per protein, subjects pooled across arms",
    x = "nominal hits / chance expectation", y = NULL,
    caption = paste(
      "Bar length is nominal hits (p < 0.05) over 5% of proteins tested. The vertical line is",
      "chance. Table: c_data/03_phenotype.xlsx, chance_expectation."
    )
  ) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid.major.y = element_blank(), legend.position = "top",
    plot.caption = element_text(size = 7, colour = "grey45", hjust = 0)
  )
chance_figure
```

## Every nominal protein goes in a dot matrix, one per window

Every protein nominal against at least one phenotype, one matrix per window. Fill is Spearman
rho, size is −log10 p, a black ring marks BH < 0.05.

```{r hit-matrices}
label_of <- dal$annotation |>
  mutate(label = if_else(
    is.na(gene) | duplicated(gene) | duplicated(gene, fromLast = TRUE),
    paste0(coalesce(gene, "?"), " (", uniprot_id, ")"), gene
  )) |>
  with(set_names(label, uniprot_id))
hit_pages <- function(data, columns, title, subtitle, table, per_page = 75) {
  ranked <- data |>
    filter(!is.na(p)) |>
    summarise(n_nominal = sum(p < 0.05), best = min(p), .by = label) |>
    filter(n_nominal > 0) |>
    arrange(desc(n_nominal), best)
  page_of <- ceiling(seq_len(nrow(ranked)) / per_page)
  map(seq_len(max(page_of, 0)), \(page) {
    rows <- ranked$label[page_of == page]
    shown <- data |>
      filter(label %in% rows) |>
      mutate(
        label = factor(label, levels = rev(rows)),
        column = factor(column, levels = columns),
        nominal = !is.na(p) & p < 0.05
      )
    ggplot(shown, aes(column, label)) +
      geom_point(data = filter(shown, !nominal), colour = "grey85", size = 0.6) +
      geom_point(
        data = filter(shown, nominal), aes(size = -log10(p), fill = effect),
        shape = 21, colour = "grey30", stroke = 0.2
      ) +
      geom_point(
        data = filter(shown, !is.na(fdr), fdr < 0.05), aes(size = -log10(p)),
        shape = 21, colour = "black", stroke = 0.9, fill = NA
      ) +
      scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B", limits = c(-1, 1)) +
      scale_size_continuous(range = c(1, 4.5), limits = c(-log10(0.05), NA)) +
      scale_x_discrete(drop = FALSE) +
      labs(
        title = sprintf("%s (%d/%d)", title, page, max(page_of)),
        subtitle = sprintf(
          "%s; %d proteins nominal in at least one column; rows %d-%d",
          subtitle, nrow(ranked), min(which(page_of == page)), max(which(page_of == page))
        ),
        x = NULL, y = NULL, fill = "rho", size = "-log10 p",
        caption = paste0(
          "Rows: proteins at nominal p < 0.05 against at least one phenotype, most columns first, ",
          "then by best p. Filled dot: nominal, fill = Spearman rho, size = -log10 p. Black ring: ",
          "BH < 0.05. Grey speck: tested, not nominal. Table: ", table, "."
        )
      ) +
      theme_minimal(base_size = 9) +
      theme(
        axis.text.x = element_text(angle = 35, hjust = 1),
        axis.text.y = element_text(size = if (length(rows) > 30) 5.5 else 8),
        plot.caption = element_text(size = 7, colour = "grey45", hjust = 0)
      )
  })
}
window_titles <- c(
  training = "training change (T2 - T1)", baseline = "level at T1",
  acute = "acute change (T3 - T2)"
)
hits <- imap(window_titles, \(title, window) {
  hit_pages(
    results |>
      filter(window == .env$window) |>
      transmute(label = label_of[uniprot_id], column = outcome, effect = rho, p, fdr),
    outcomes,
    title = paste("Protein association hits,", title),
    subtitle = "Spearman against each phenotype",
    table = "c_data/03_phenotype.xlsx, protein_association"
  )
})
map_int(hits, length)
```

```{r write}
pdf(file.path(figures, "03_phenotype_figures.pdf"), width = 11, height = 8.5)
walk(c(list(chance_figure), list_flatten(unname(hits))), print)
invisible(dev.off())
saveRDS(list(protein_association = results), file.path(out, "phenotype.rds"), compress = "xz")
# by_arm runs to tens of thousands of rows; the workbook keeps the nominal ones.
sheets <- list(
  chance_expectation = chance_expectation, protein_association = arrange(results, fdr),
  protein_by_arm = filter(by_arm, p < 0.05),
  input_manifest = tibble(
    input = names(inputs), path = unname(inputs), md5 = unname(tools::md5sum(paths))
  ),
  package_versions = sessioninfo::package_info("loaded", dependencies = FALSE) |>
    as_tibble() |>
    select(package, version = loadedversion, source)
)
read_me <- tibble(sheet = names(sheets), holds = c(
  "Nominal hits against chance, per window and outcome. Read first.",
  "Spearman rho, p and BH FDR per protein, window and outcome, subjects pooled.",
  "Within-arm correlations with p < 0.05.",
  "Files read, with md5.",
  "Packages loaded at run time."
))
writexl::write_xlsx(c(list(read_me = read_me), sheets), file.path(out, "03_phenotype.xlsx"))
```
