---
title: "02 · Normalization"
subtitle: "Cyclic loess on the filtered DAList"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    df-print: kable
    fig-width: 7
    fig-height: 5
    fig-dpi: 150
    fig-align: center
execute:
  message: false
---

proteoDA compares normalisation methods on the filtered matrix, then applies cyclic loess. The
matrix stays unimputed: limma fits each protein on the samples where it was seen, so the
differential stage reads this file directly.

```{r setup}
library(here)
library(proteoDA)
library(dplyr)
library(tibble)
library(limma)
library(ggplot2)
library(writexl)

inputs <- c(filtered = "01_Preprocess/01_Filtering/c_data/DAList_filtered.rds")
paths <- vapply(inputs, here, character(1))
stopifnot(file.exists(paths))
dal <- readRDS(paths[["filtered"]])
out <- here("01_Preprocess", "02_Normalization", "c_data")
reports <- here("01_Preprocess", "02_Normalization", "b_reports")
for (path in c(out, reports)) dir.create(path, recursive = TRUE, showWarnings = FALSE)
c(proteins = nrow(dal$data), samples = ncol(dal$data))
```

## proteoDA compares the methods, then cyclic loess runs

`write_norm_report()` draws every method proteoDA offers against the grouping; the QC reports
bracket the chosen one. All three PDFs land in `b_reports/`.

```{r normalise}
write_norm_report(dal,
  grouping_column = "group", output_dir = reports,
  filename = "norm_comparison.pdf", overwrite = TRUE
)
write_qc_report(dal,
  color_column = "group", output_dir = reports,
  filename = "qc_pre.pdf", overwrite = TRUE
)
dal <- normalize_data(dal, norm_method = "cycloess")
write_qc_report(dal,
  color_column = "group", output_dir = reports,
  filename = "qc_post.pdf", overwrite = TRUE
)
```

Cyclic loess here is `limma::normalizeCyclicLoess(method = "fast")`. Its span is not the 0.7
in the formals: `adaptive.span = TRUE` replaces it with `chooseLowessSpan(nrow)`, so the span
moves whenever filtering changes the protein count.

## PC1 carries 11.6% of the variance after normalisation

Share of each protein's variance explained by the six group cells (eta squared, 1 minus the
residual over the total sum of squares), and the leading components after normalisation.

```{r structure}
stopifnot(identical(dal$metadata$sample_id, colnames(dal$data)))
cell_fit <- lmFit(dal$data, model.matrix(~group, dal$metadata))
total_ss <- apply(dal$data, 1, \(x) sum((x - mean(x, na.rm = TRUE))^2, na.rm = TRUE))
eta2 <- 1 - cell_fit$sigma^2 * cell_fit$df.residual / total_ss
eta2[rowSums(!is.na(dal$data)) < 4] <- NA
filled <- apply(dal$data, 2, \(x) replace(x, is.na(x), median(x, na.rm = TRUE)))
pca <- prcomp(t(filled), center = TRUE, scale. = TRUE)
components <- tibble(
  component = paste0("PC", 1:4),
  variance_explained = round(summary(pca)$importance[2, 1:4], 3)
)
components
summary(eta2)
```

```{r fig-pca}
#| fig-width: 6
#| fig-height: 4.5
pca_scores <- as_tibble(pca$x[, 1:2], rownames = "sample_id") |>
  left_join(dal$metadata, by = "sample_id")
pca_figure <- ggplot(pca_scores, aes(PC1, PC2, colour = timepoint, shape = arm)) +
  geom_point(size = 2.2) +
  scale_colour_manual(values = c(T1 = "#E69F00", T2 = "#0072B2", T3 = "#009E73")) +
  labs(
    title = "Samples after normalisation",
    subtitle = sprintf("PCA of %d proteins, gaps median-filled for display", nrow(filled)),
    x = sprintf("PC1 (%.1f%%)", 100 * components$variance_explained[1]),
    y = sprintf("PC2 (%.1f%%)", 100 * components$variance_explained[2])
  ) +
  theme_minimal(base_size = 10)
pca_figure
```

## Writes DAList_normalized.rds, the workbook and the figure PDF

```{r write}
saveRDS(dal, file.path(out, "DAList_normalized.rds"), compress = "xz")
pdf(file.path(reports, "02_normalize_figures.pdf"), width = 11, height = 8.5)
print(pca_figure)
invisible(dev.off())
sheets <- list(
  normalized = bind_cols(
    as_tibble(dal$annotation) |> select(uniprot_id, protein, gene, description),
    as_tibble(dal$data)
  ),
  components = components,
  pca_scores = select(pca_scores, sample_id, subject, arm, timepoint, PC1, PC2),
  eta_squared = tibble(uniprot_id = rownames(dal$data), eta2 = unname(eta2)),
  input_manifest = tibble(
    input = names(inputs), path = unname(inputs), md5 = unname(tools::md5sum(paths))
  ),
  package_versions = sessioninfo::package_info("loaded", dependencies = FALSE) |>
    as_tibble() |>
    select(package, version = loadedversion, source)
)
read_me <- tibble(sheet = names(sheets), holds = c(
  "Cyclic-loess log2 abundance, one row per protein.",
  "Variance explained by the first four principal components.",
  "Each sample's PC1 and PC2 score.",
  "Share of each protein's variance explained by the six group cells.",
  "Files read, with md5.",
  "Packages loaded at run time."
))
write_xlsx(c(list(read_me = read_me), sheets), file.path(out, "02_normalize.xlsx"))
```
