---
title: "02 · Normalization"
subtitle: "Cyclic loess on the filtered DAList"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    df-print: kable
    fig-width: 7
    fig-height: 5
    fig-dpi: 150
    fig-align: center
execute:
  message: false
---

proteoDA compares normalisation methods on the filtered matrix, then applies cyclic loess. The
matrix stays unimputed: limma fits each protein on the samples where it was seen, so the
differential stage reads this file directly.

```{r setup}
library(here)
library(proteoDA)
library(dplyr)
library(tibble)
library(ggplot2)
library(writexl)

dal <- readRDS(here("01_Preprocess", "01_Filtering", "c_data", "DAList_filtered.rds"))
out <- here("01_Preprocess", "02_Normalization", "c_data")
reports <- here("01_Preprocess", "02_Normalization", "b_reports")
dir.create(out, recursive = TRUE, showWarnings = FALSE)
c(proteins = nrow(dal$data), samples = ncol(dal$data))
```

## Compare, then normalise

`write_norm_report()` draws every method proteoDA offers against the grouping; the QC reports
bracket the chosen one. All three PDFs land in `b_reports/`.

```{r normalise}
write_norm_report(dal,
  grouping_column = "Group_Time", output_dir = reports,
  filename = "norm_comparison.pdf", overwrite = TRUE
)
write_qc_report(dal,
  color_column = "Group_Time", output_dir = reports,
  filename = "qc_pre.pdf", overwrite = TRUE
)
dal <- normalize_data(dal, norm_method = "cycloess")
write_qc_report(dal,
  color_column = "Group_Time", output_dir = reports,
  filename = "qc_post.pdf", overwrite = TRUE
)
```

Cyclic loess here is `limma::normalizeCyclicLoess(method = "fast")`. Its span is not the 0.7
in the formals: `adaptive.span = TRUE` replaces it with `chooseLowessSpan(nrow)`, so the span
moves whenever filtering changes the protein count.

## What drives the matrix

Share of each protein's variance explained by the six Group_Time cells (eta squared), and the
leading components after normalisation.

```{r structure}
group <- dal$metadata$Group_Time[match(colnames(dal$data), dal$metadata$Col_ID)]
eta2 <- apply(dal$data, 1, \(x) {
  ok <- !is.na(x)
  if (sum(ok) < 4) {
    return(NA_real_)
  }
  fit <- summary(aov(x[ok] ~ group[ok]))[[1]]
  fit[1, 2] / sum(fit[, 2])
})
filled <- apply(dal$data, 2, \(x) replace(x, is.na(x), median(x, na.rm = TRUE)))
pca <- prcomp(t(filled), center = TRUE, scale. = TRUE)
components <- tibble(
  component = paste0("PC", 1:4),
  variance_explained = round(summary(pca)$importance[2, 1:4], 3)
)
components
summary(eta2)
```

```{r fig-pca}
#| fig-width: 6
#| fig-height: 4.5
as_tibble(pca$x[, 1:2], rownames = "Col_ID") |>
  left_join(dal$metadata, by = "Col_ID") |>
  ggplot(aes(PC1, PC2, colour = Timepoint, shape = Group)) +
  geom_point(size = 2.2) +
  scale_colour_manual(values = c(T1 = "#E69F00", T2 = "#0072B2", T3 = "#009E73")) +
  labs(
    title = "Samples after normalisation",
    subtitle = sprintf("PCA of %d proteins, gaps median-filled for display", nrow(filled)),
    x = sprintf("PC1 (%.1f%%)", 100 * components$variance_explained[1]),
    y = sprintf("PC2 (%.1f%%)", 100 * components$variance_explained[2])
  ) +
  theme_minimal(base_size = 10)
```

## Outputs

```{r write}
saveRDS(dal, file.path(out, "DAList_normalized.rds"))
write_xlsx(
  list(
    normalized = bind_cols(
      as_tibble(dal$annotation) |> select(uniprot_id, protein, gene, description),
      as_tibble(dal$data)
    ),
    components = components,
    eta_squared = tibble(uniprot_id = rownames(dal$data), eta2 = unname(eta2))
  ),
  file.path(out, "02_normalize.xlsx")
)
```

```{r session}
sessionInfo()
```
