---
title: "HRvLR 02 — Normalization & Imputation"
subtitle: "Cyclic-loess normalization, the eta-squared scan, and when to fill missing values"
author: "A_HRvLR_2026"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
    embed-resources: true
    df-print: paged
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(here)
knitr::opts_knit$set(root.dir = here::here())
```

## What stage 02 does

Stage 02 takes the filtered `DAList` from stage 01 and makes its samples
comparable. Cyclic loess normalization removes the sample-to-sample intensity
drift that comes from uneven loading, and a per-protein eta-squared scan records
how much of each protein's variance the `Group_Time` cells explain. Stage 02
also fills the missing values three different ways, but those imputed matrices
are for downstream figures and clustering. The differential-expression test in
stage 03 runs on the normalized matrix with its gaps intact.

This page runs that logic live and in memory, the same as
`02_Normalization/a_script/01_run_normalization.R` and the three imputation arms
in `02_Normalization/imputation/a_script/`, and writes nothing. The last section
confirms the in-memory result matches the committed `DAList_normalized.rds`.

## Reads and writes

The stage reads one file, stage 01's `DAList_filtered.rds`. The runnable script
writes `DAList_normalized.rds` and `normalized.csv` (the primary hand-off to
stage 03), `00_report_intermediates.rds` (the eta-squared scan, the
post-normalization PCA, and the stage-01 intermediates, reused by the report and
figure layers), and proteoDA's QC PDFs to `b_reports/`. The three imputation
scripts each read the normalized object and write `DAList_imputed_imp4p.rds`,
`DAList_imputed_mscoreutils.rds`, and `DAList_imputed_missforest.rds` to
`imputation/c_data/`. This document recomputes the objects but saves none of
them.

## Load the filtered `DAList` {#sec-load}

Stage 01 hands off a filtered, un-normalized object that still carries its
missing values. Its `tags` slot records the group filter that produced it.

```{r}
#| label: load
library(proteoDA)
dal <- readRDS(here("01_Filtering", "c_data", "DAList_filtered.rds"))
miss_before <- round(100 * mean(is.na(dal$data)), 1)
dim(dal$data)
miss_before # missing fraction, not yet imputed
dal$tags # provenance so far: the group filter
```

## Normalize with cyclic loess {#sec-normalize}

Samples load onto the instrument slightly differently, so a whole sample can sit
higher or lower for reasons that have nothing to do with biology. That offset
shows up in each sample's median log-intensity and would distort any cross-sample
comparison. `normalize_data(norm_method = "cycloess")` is proteoDA's wrapper
around limma's cyclic loess, which fits smooth curves between pairs of samples to
pull them onto a common scale (limma defaults: `span = 0.7`,
`adaptive.span = FALSE`). It log2-transforms internally, so the normalized matrix
comes back on the log2 scale.

```{r}
#| label: normalize
before <- apply(log2(as.matrix(dal$data)), 2, median, na.rm = TRUE)
dal <- normalize_data(dal, norm_method = "cycloess")
after <- apply(as.matrix(dal$data), 2, median, na.rm = TRUE)

rbind(before = summary(before), after = summary(after)) # spread collapses to a tight band
dal$tags # new 'normalized' + 'norm_method' tags
```

Normalization rescales the observed values but fills nothing in, so the missing
fraction is unchanged:

```{r}
#| label: missing-unchanged
c(
  before_pct = miss_before,
  after_pct = round(100 * mean(is.na(dal$data)), 1)
)
```

The script writes proteoDA's audit PDFs to `b_reports/`. `write_norm_report()`
compares the candidate normalization methods on the filtered matrix before the
choice is committed, and `write_qc_report()` inspects the matrix before and after.
The calls are shown here and run by the script, which keeps the cycloess choice
reviewable without re-implementing the comparison by hand.

```{r}
#| label: qc-reports
#| eval: false
write_norm_report(dal,
  grouping_column = "Group_Time", output_dir = report_dir,
  filename = "norm_comparison.pdf"
) # compares candidate methods (before normalize)
write_qc_report(dal,
  color_column = "Group_Time", output_dir = report_dir,
  filename = "qc_post.pdf"
) # inspects the chosen result
```

This normalized, non-imputed object is the primary hand-off to stage 03. The
imputation section below covers why the test runs on it rather than on a filled
matrix.

## The eta-squared scan {#sec-eta2}

For the figure layer, the script records how much of each protein's variance the
six `Group_Time` cells explain: eta-squared, the between-group sum of squares
over the total. It is a one-way decomposition computed per protein directly on
the normalized values, which are already on the log2 scale, so no further
transform is applied. It feeds the F02 protein-variance panel.

```{r}
#| label: eta2
log_dat <- dal$data
grp_vec <- dal$metadata$Group_Time[match(colnames(log_dat), dal$metadata$Col_ID)]
eta2_vals <- apply(log_dat, 1, function(x) {
  ok <- !is.na(x)
  if (sum(ok) < 4) {
    return(NA_real_)
  }
  xk <- x[ok]
  gk <- grp_vec[ok]
  ss_b <- sum(tapply(xk, gk, length) * (tapply(xk, gk, mean) - mean(xk))^2)
  ss_t <- sum((xk - mean(xk))^2)
  if (ss_t > 0) ss_b / ss_t else NA_real_
})
summary(eta2_vals) # most proteins low; a tail is strongly group-structured
```

The script bundles `eta2_vals`, the post-normalization PCA, and the stage-01
intermediates into `00_report_intermediates.rds` for the reports and figures.

## Imputation for methods that need a complete matrix {#sec-impute}

PCA, clustering, WGCNA, and the rotation tests cannot run on a matrix with gaps.
Three imputation methods fill the normalized matrix, kept side by side so the
choice can be checked against alternatives. missForest is the canonical arm that
the figures and WGCNA read; imp4p and the MsCoreUtils hybrid are comparisons.

None of them feeds the differential-expression test. An imputed value is a
prediction, not a measurement, and it carries none of the measurement error the
real values do. Testing on a filled matrix understates each protein's variance,
shrinks its standard error, and calls more proteins significant than the data
support (Webb-Robertson et al. 2015, PMID 25855118). limma avoids this by fitting
each protein on the samples where it was actually observed, so the test runs on
the non-imputed matrix and imputation stays downstream.

```{r}
#| label: impute-input
mat <- as.matrix(dal$data)
was_missing <- is.na(mat)
c(entries_imputed = sum(was_missing), pct_of_matrix = round(100 * mean(was_missing), 1))
```

### missForest (canonical) {#sec-missforest}

`missForest` is a non-parametric random-forest imputer: each protein's missing
values are predicted from the multivariate structure of the other proteins, so
the matrix is transposed to put proteins in the columns. The fit is stochastic,
so the seed is fixed and the rows are ordered before it runs (Stekhoven and
Buhlmann 2012).

```{r}
#| label: missforest
#| cache: true
library(missForest)
set.seed(42)
ord <- order(rownames(mat))
mf <- missForest(t(mat[ord, ]), maxiter = 10, ntree = 100, verbose = FALSE)
imp <- t(mf$ximp)[rownames(mat), ]
dimnames(imp) <- dimnames(mat)

c(na_remaining = sum(is.na(imp)), OOB_error = round(mf$OOBerror[1], 4))
```

### imp4p (comparison) {#sec-imp4p}

`imp4p` is mechanism-aware. It estimates, per protein, the mix of values missing
at random against values missing because the protein sits below the detection
limit, then imputes each part with a method suited to it (Giai Gianetto et al.):

```{r}
#| label: imp4p
#| cache: true
library(imp4p)
set.seed(42)
cond <- factor(dal$metadata$Group_Time[match(colnames(mat), dal$metadata$Col_ID)])
imp4p_mat <- as.matrix(impute.mi(
  tab = mat, conditions = cond,
  methodMCAR = "mle", methodMNAR = "igcda", progress.bar = FALSE
))
dimnames(imp4p_mat) <- dimnames(mat)
c(na_remaining = sum(is.na(imp4p_mat))) # 0 remaining — complete matrix
```

### MsCoreUtils hybrid (comparison) {#sec-mscoreutils}

The third method splits the proteins first. `imputeLCMD::model.Selector()`
classifies each protein as missing-at-random or not (Lazar et al. 2016), then
`MsCoreUtils::impute_matrix()` routes the MAR features through kNN and the MNAR
features through a left-censored QRILC draw:

```{r}
#| label: mscoreutils
library(MsCoreUtils)
library(imputeLCMD)
set.seed(42)
randna <- as.logical(model.Selector(mat)[[1]]) # TRUE = MAR feature
ms_mat <- impute_matrix(mat, method = "mixed", randna = randna, mar = "knn", mnar = "QRILC")

c(MAR_features = sum(randna), MNAR_features = sum(!randna), na_remaining = sum(is.na(ms_mat)))
```

## Result and verification {#sec-recap}

The runnable script saves the normalized matrix to `02_Normalization/c_data/`;
the three imputation arms save to `02_Normalization/imputation/c_data/`.

| File | Location | Missing values? | Used by |
|---|---|---|---|
| `DAList_normalized.rds` | `c_data/` | kept | the DE test (stage 03), primary |
| `DAList_imputed_missforest.rds` | `imputation/c_data/` | filled | figures, WGCNA (canonical) |
| `DAList_imputed_imp4p.rds` | `imputation/c_data/` | filled | the imputed-DEP comparison |
| `DAList_imputed_mscoreutils.rds` | `imputation/c_data/` | filled | the imputed-DEP comparison |

The in-memory object matches the committed `DAList_normalized.rds`:

```{r}
#| label: verify
committed <- readRDS(here("02_Normalization", "c_data", "DAList_normalized.rds"))
c(
  same_dim = identical(dim(dal$data), dim(committed$data)),
  same_proteins = identical(sort(rownames(dal$data)), sort(rownames(committed$data))),
  same_samples = identical(sort(colnames(dal$data)), sort(colnames(committed$data)))
)
```

Normalization levels the samples without inventing any data, and imputation is
applied only where a method cannot run on an incomplete matrix. Stage 03 fits the
model on the non-imputed normalized matrix.
