---
title: "HRvLR 02 — Normalization & Imputation"
subtitle: "Cyclic-loess normalization, the eta-squared scan, and when to fill missing values"
author: "A_HRvLR_2026"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
    embed-resources: true
    df-print: paged
bibliography: ../../04_Figures/shared/references.bib
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
pacman::p_load(here)
knitr::opts_knit$set(root.dir = here::here())
```

## What stage 02 does

Stage 02 takes the filtered `DAList` from stage 01 and makes its samples
comparable. The stage is built on proteoDA [@thurman2023], which supplies the
`DAList` container, the normalization wrapper, and the QC reports used below.
Cyclic loess normalization [@bolstad2003] removes the sample-to-sample intensity
drift that comes from uneven loading, and a per-protein eta-squared scan records
how much of each protein's variance the `Group_Time` cells explain. Stage 02
also fills the missing values four different ways. Those imputed matrices drive the
downstream figures, the WGCNA modules, and the imputation-sensitivity comparison in
`03_DEP/b_imputed/`. The *reported* differential-expression test in stage 03 runs on the
normalized matrix with its gaps intact.

This page runs that logic live and in memory, the same as
`02_Normalization/a_script/01_run_normalization.R` and the four imputation arms
in `02_Normalization/imputation/a_script/`, and writes nothing. The last section
confirms the in-memory result matches the committed `DAList_normalized.rds`.

## Reads and writes

The stage reads two stage-01 files: `DAList_filtered.rds` (the matrix it
normalizes) and `filtering_intermediates.rds` (passed straight through into the
report bundle). The runnable script writes `DAList_normalized.rds` and
`normalized.csv` (the primary hand-off to stage 03), `00_report_intermediates.rds`
(the eta-squared scan, the post-normalization PCA, and those stage-01
intermediates, reused by the report and figure layers), and proteoDA's QC PDFs to
`b_reports/`. The four imputation
scripts each read the normalized object and write `DAList_imputed_imp4p.rds`,
`DAList_imputed_mscoreutils.rds`, `DAList_imputed_missforest.rds`, and
`DAList_imputed_perseus.rds` to `imputation/c_data/`. This document recomputes the
objects but saves none of them.

## Load the filtered `DAList` {#sec-load}

Stage 01 hands off a filtered, un-normalized object that still carries its
missing values. Its `tags` slot records the group filter that produced it.

```{r}
#| label: load
pacman::p_load(proteoDA)
dal <- readRDS(here("01_Filtering", "c_data", "DAList_filtered.rds"))
miss_before <- round(100 * mean(is.na(dal$data)), 1)
dim(dal$data)
miss_before # missing fraction, not yet imputed
dal$tags # provenance so far: the group filter
```

## Normalize with cyclic loess {#sec-normalize}

Samples load onto the instrument slightly differently, so a whole sample can sit
higher or lower for reasons that have nothing to do with biology. That offset
shows up in each sample's median log-intensity and would distort any cross-sample
comparison. `normalize_data(norm_method = "cycloess")` is proteoDA's wrapper
around limma's cyclic loess [@bolstad2003; @ritchie2015]. proteoDA passes
`method = "fast"`, so each sample is regressed against the row-mean reference rather than
against every other sample pairwise, and the cycle repeats three times. It log2-transforms
internally, so the normalized matrix comes back on the log2 scale.

The span is **0.5075, not 0.7**. proteoDA never passes `span`, and limma defaults
`adaptive.span = TRUE`, which overrides the `span = 0.7` formal with
`chooseLowessSpan(nrow(x))` — at 1,920 proteins that evaluates to 0.5075. The span is
therefore a function of the protein count and moves if the filter moves. Cyclic loess rests on two
assumptions: that most proteins do not change between samples, and that whatever
bias exists varies smoothly with intensity rather than jumping around. Both are
standard for label-free LC-MS and are the same assumptions that motivate
normalizing at all [@karpievitch2012; @arend2025].

```{r}
#| label: normalize
before <- apply(log2(as.matrix(dal$data)), 2, median, na.rm = TRUE)
dal <- normalize_data(dal, norm_method = "cycloess")
after <- apply(as.matrix(dal$data), 2, median, na.rm = TRUE)

rbind(before = summary(before), after = summary(after)) # spread collapses to a tight band
dal$tags # new 'normalized' + 'norm_method' tags
```

Normalization rescales the observed values but fills nothing in, so the missing
fraction is unchanged:

```{r}
#| label: missing-unchanged
c(
  before_pct = miss_before,
  after_pct = round(100 * mean(is.na(dal$data)), 1)
)
```

The script writes proteoDA's audit PDFs to `b_reports/`. `write_norm_report()`
compares the candidate normalization methods on the filtered matrix before the
choice is committed, and `write_qc_report()` inspects the matrix before and after.
The calls are shown here and run by the script, which keeps the cycloess choice
reviewable without re-implementing the comparison by hand.

```{r}
#| label: qc-reports
#| eval: false
write_norm_report(dal,
  grouping_column = "Group_Time", output_dir = report_dir,
  filename = "norm_comparison.pdf"
) # compares candidate methods (before normalize)
write_qc_report(dal,
  color_column = "Group_Time", output_dir = report_dir,
  filename = "qc_post.pdf"
) # inspects the chosen result
```

Be clear about what that does and does not show here. `write_norm_report()` puts all
eight methods proteoDA offers — log2, median, mean, vsn, quantile, cycloess, rlr,
and gi — side by side on the same filtered matrix and scores each one on five
within-group dispersion metrics: pooled coefficient of variation (PCV), pooled
median absolute deviation (PMAD), pooled estimate of variance (PEV), intragroup
correlation (COR), and the density of log2 ratios. A method that levels the samples
without distorting them should shrink the first three, raise the fourth, and centre
the log2-ratio density on zero.

The catch is that the output is a multi-page PDF — `b_reports/norm_comparison.pdf`,
with `qc_pre.pdf` and `qc_post.pdf` for the before and after — and a PDF cannot be
embedded in this page. So the comparison is not reproduced here. The cycloess choice
is asserted on this page, not demonstrated on it, and a reader is being asked to take
it on trust unless they open `norm_comparison.pdf` and read the eight panels
themselves. The call that generates the comparison is directly above; the result of
that comparison is not summarized here, because summarizing a figure nobody has shown
you is how an unchecked choice becomes a fact.

This normalized, non-imputed object is the primary hand-off to stage 03. The
imputation section below covers why the test runs on it rather than on a filled
matrix.

## The eta-squared scan {#sec-eta2}

For the figure layer, the script records how much of each protein's variance the
six `Group_Time` cells explain: eta-squared, the between-group sum of squares
over the total. It is a one-way decomposition computed per protein directly on
the normalized values, which are already on the log2 scale, so no further
transform is applied. It feeds the F02 protein-variance panel.

```{r}
#| label: eta2
log_dat <- dal$data
grp_vec <- dal$metadata$Group_Time[match(colnames(log_dat), dal$metadata$Col_ID)]
eta2_vals <- apply(log_dat, 1, function(x) {
  ok <- !is.na(x)
  if (sum(ok) < 4) {
    return(NA_real_)
  }
  xk <- x[ok]
  gk <- grp_vec[ok]
  ss_b <- sum(tapply(xk, gk, length) * (tapply(xk, gk, mean) - mean(xk))^2)
  ss_t <- sum((xk - mean(xk))^2)
  if (ss_t > 0) ss_b / ss_t else NA_real_
})
summary(eta2_vals) # most proteins low; a tail is strongly group-structured
```

Read those numbers with the sample size in mind. Eta-squared is upward-biased: it is
the ratio of between-group to total sum of squares in the sample, and it never
subtracts the variance the grouping soaks up purely by chance. With six `Group_Time`
cells there are 5 between-group degrees of freedom against roughly 39 residual, so
even if the groups were assigned at random the expected eta-squared is already about
5 / 44 ≈ 0.11. A protein sitting near 0.11 is showing you the null, not structure.
Only the tail well above it is worth reading as group-driven, and even the tail is
a screening signal rather than a test — nothing here is corrected for the thousands
of proteins scanned.

This is why the F02 supplement reports bias-corrected omega-squared instead, which
subtracts the expected chance contribution and so has a null expectation near zero.
The bias is not even constant across proteins: the scan above drops missing values
per protein (`ok <- !is.na(x)`), so a protein observed in fewer samples has fewer
residual degrees of freedom and therefore *more* inflation than a fully observed one.
The gappiest proteins are flattered most. That makes the raw eta-squared usable as a
descriptive spread for the F02 panel, but not as a clean between-protein ranking, and
it is another reason the omega-squared version is the one that carries any claim.

The script bundles `eta2_vals`, the post-normalization PCA, and the stage-01
intermediates into `00_report_intermediates.rds` for the reports and figures.

## Imputation for methods that need a complete matrix {#sec-impute}

PCA, clustering, WGCNA, and the rotation tests cannot run on a matrix with gaps.
Four imputation methods fill the normalized matrix, kept side by side so the
choice can be checked against alternatives. missForest is the canonical arm that
the figures and WGCNA read; imp4p, the MsCoreUtils hybrid, and Perseus are
comparisons.

### Why a gap is not just a gap: MAR and MNAR {#sec-mar-mnar}

Every method below turns on one question, so it is worth answering before the
acronyms start appearing: *why* is this value missing?

**MAR — missing at random.** The gap has nothing to do with the protein's true
abundance. The peptide failed to be identified in that run, the peak was not picked,
the search engine lost it. The protein was there at whatever level it was there at,
and the reason we cannot see it is unrelated to that level. A MAR gap carries no
information about the value it hides, which is what licenses predicting it from other
proteins: the neighbours are as informative as anything else, because the gap itself
tells you nothing.

**MNAR — missing not at random.** The gap exists *because of* the value. The protein
fell below the instrument's detection limit, so the very fact that it is missing tells
you something: it means **low**. An MNAR gap is data. Filling it by prediction from
other proteins throws away the one thing you knew about it and, worse, will typically
fill it with something far too high, because the other proteins have no idea it was
scraping the floor.

Real DIA data is a mixture of the two, skewed toward MNAR — the low-abundance end of
the dynamic range is exactly where dropouts concentrate, and that is not a
coincidence, it is the detection limit doing its job [@lazar2016; @webbrobertson2015].

This is the whole reason the four arms below disagree, and they disagree by design
rather than by accident. missForest treats every gap as MAR and predicts it from the
multivariate structure of the other proteins. The Perseus draw treats every gap as
MNAR and fills it from a downshifted low-intensity Gaussian. Handed the same missing
cell, one asks "what do this protein's neighbours imply it should be?" and the other
asks "how low is low?" — and they answer differently. imp4p and the MsCoreUtils hybrid
sit between them: both try to estimate which gaps are which, then route each kind to a
method suited to it. Keeping all four side by side is how the pipeline avoids
pretending it knows the mixture exactly.

### Why none of this touches the *primary* DE test {#sec-no-impute-de}

No imputed matrix feeds the **primary** differential-expression test, and this is the most
consequential decision in the stage. The canonical stage-03 arm
(`03_DEP/a_non_imputed/`) runs on the normalized matrix with its gaps intact.

Be precise about the scope of that claim, because it is easy to overstate. The imputed
matrices **do** feed a differential-expression test: the four-arm comparison in
`03_DEP/b_imputed/` fits limma to every one of them. `DAList_imputed_missforest.rds` also
feeds the WGCNA modules, F02, F04_association/WGCNA, the concordance figures and the module engine. What the
imputed arms never do is produce the reported result.

That comparison is not decoration. π-hit counts across the five arms run
**267 / 159 / 307 / 511 / 827** — a 5.2-fold swing. Imputation choice is the single largest
lever on the significant-protein count in this pipeline, which is precisely why the primary
arm refuses to pull it.

An imputed value is a prediction, not a measurement. It is a number the model made up,
and — this is the part that bites — it arrives with none of the measurement error a
real observation would carry. A filled cell looks *more certain* to a downstream test
than any cell that was actually measured. Fill a protein's gaps and you have quietly
told the test that you measured it more times, and more cleanly, than you did.

The consequence is mechanical. Testing on a filled matrix understates each protein's
variance, which shrinks its standard error, which inflates its t-statistic, which
calls more proteins significant than the data support. The false positives are not a
side effect; they follow directly from the variance you erased
[@webbrobertson2015; @peng2024]. And the damage is worst precisely where the imputation
was most confident — a protein whose gaps were all filled from a tight distribution
comes out looking like the most reliably measured protein in the experiment.

limma does not need the gaps filled [@ritchie2015]. It fits each protein on the samples
where that protein was actually observed, letting the residual degrees of freedom drop
for proteins with more missingness — which is the honest bookkeeping: a protein seen in
fewer samples *is* known with less confidence, and the model's error bars should say so.
Fitting limma without imputing has been shown to outperform imputing first [@peng2024].
So the test runs on the non-imputed matrix, and imputation stays strictly downstream,
where it serves methods (PCA, WGCNA, clustering) that are descriptive and cannot run on
gaps at all — methods where a wrong fill distorts a picture rather than manufacturing a
p-value.

```{r}
#| label: impute-input
mat <- as.matrix(dal$data)
was_missing <- is.na(mat)
c(entries_imputed = sum(was_missing), pct_of_matrix = round(100 * mean(was_missing), 1))
```

### missForest (canonical) {#sec-missforest}

`missForest` is a non-parametric random-forest imputer: each protein's missing
values are predicted from the multivariate structure of the other proteins, so
the matrix is transposed to put proteins in the columns. The fit is stochastic,
so the seed is fixed and the rows are ordered before it runs [@stekhoven2012].
Note what the previous section implies about this arm: it is a MAR imputer, and
our gaps are skewed MNAR, so it is expected to fill below-detection values too
high. It is the canonical arm because it is well behaved on the descriptive uses
(PCA, WGCNA) it feeds, not because its missingness assumption is the right one.

`maxiter = 10` and `ntree = 100` are the values in the call, and they are
conventional defaults rather than anything tuned on this dataset. `ntree = 100` is
the size of each protein's forest; `maxiter = 10` is a ceiling on the iterations,
not a target, because missForest stops early on its own once the imputed matrix
stops changing between passes. Both trade compute against a small amount of
stability, and neither was selected by comparing outcomes here — which is worth
saying plainly rather than letting the numbers sit there looking chosen.

```{r}
#| label: missforest
#| cache: true
pacman::p_load(missForest)
set.seed(42)
ord <- order(rownames(mat))
mf <- missForest(t(mat[ord, ]), maxiter = 10, ntree = 100, verbose = FALSE)
imp <- t(mf$ximp)[rownames(mat), ]
dimnames(imp) <- dimnames(mat)

c(na_remaining = sum(is.na(imp)), OOB_error = round(mf$OOBerror[1], 4))
```

That `OOB_error` is the one diagnostic missForest gives you for free, and it should
not go past unread. Each tree in a random forest is grown on a bootstrap sample of
the data, which leaves roughly a third of the observations *out of the bag* for that
tree — unseen during its training. Predicting those held-out observations and scoring
the error gives a held-out error estimate without any separate validation split. For
continuous data missForest reports it as a normalized root mean squared error: the
prediction error scaled by the variance of the observed values.

The scale is the useful part. Zero would be perfect prediction. **One** is the number
to anchor on: an OOB error of 1 means the forest predicts a missing value no better
than simply guessing that protein's mean, because the error is as large as the spread
it was trying to explain. Values comfortably below 1 mean the other proteins genuinely
carry information about the missing one; values near or above 1 mean the imputation is
adding noise dressed up as data. In a correlated proteomic matrix a value well under 1
is the expected and reassuring result — proteins covary, so neighbours predict. Read
the printed number against that benchmark rather than against zero.

One caution: a low OOB error says the forest predicts *observed* values well. It says
nothing about whether the values it invented for the below-detection gaps are right,
because there is no ground truth there to score against. A MAR imputer can be
internally excellent and still systematically fill MNAR gaps too high. OOB error will
not catch that; it cannot.

### imp4p (comparison) {#sec-imp4p}

`imp4p` is mechanism-aware. It estimates, per protein, the mix of values missing
at random against values missing because the protein sits below the detection
limit, then imputes each part with a method suited to it [@giaigianetto2016]:

```{r}
#| label: imp4p
#| cache: true
pacman::p_load(imp4p)
set.seed(42)
cond <- factor(dal$metadata$Group_Time[match(colnames(mat), dal$metadata$Col_ID)])
imp4p_mat <- as.matrix(impute.mi(
  tab = mat, conditions = cond,
  methodMCAR = "mle", methodMNAR = "igcda", progress.bar = FALSE
))
dimnames(imp4p_mat) <- dimnames(mat)
c(na_remaining = sum(is.na(imp4p_mat))) # 0 remaining — complete matrix
```

### MsCoreUtils hybrid (comparison) {#sec-mscoreutils}

The third method splits the proteins first. `imputeLCMD::model.Selector()`
classifies each protein as missing-at-random or not [@lazar2016], then
`MsCoreUtils::impute_matrix()` routes the MAR features through kNN and the MNAR
features through a left-censored QRILC draw. This is the hybrid position from the
section above made literal: rather than assuming one mechanism for the whole matrix,
it decides per protein and sends each to the imputer that matches. The `randna` flag
in the code below is that decision — `TRUE` marks a feature judged MAR.

```{r}
#| label: mscoreutils
pacman::p_load(MsCoreUtils, imputeLCMD)
set.seed(42)
randna <- as.logical(model.Selector(mat)[[1]]) # TRUE = MAR feature
ms_mat <- impute_matrix(mat, method = "mixed", randna = randna, mar = "knn", mnar = "QRILC")

c(MAR_features = sum(randna), MNAR_features = sum(!randna), na_remaining = sum(is.na(ms_mat)))
```

### Perseus MNAR (comparison) {#sec-perseus}

`perseus_impute()` reproduces the Perseus default: every missing value is drawn
from a Gaussian shifted down into the low-intensity tail, the pattern expected when
a value is missing because it fell below detection. It is the pure MNAR counterpart
to missForest, which treats every gap as missing at random. The two bracket the
plausible range, which is why both are kept.

The `shift = 1.8` and `width = 0.3` are the MaxQuant/Perseus convention
[@tyanova2016], and they are readable once unpacked. Both are expressed in units of
the observed distribution for that sample. `shift = 1.8` places the centre of the
draw 1.8 standard deviations **below** the observed mean; `width = 0.3` gives the
draw 0.3× the observed standard deviation. Together they say: *low, and tightly
clustered* — which is exactly what a population of below-detection values should look
like. They are squeezed against the floor of the dynamic range, and they do not vary
much, because there is not much room left underneath them to vary in.

These are conventional defaults, not values tuned on this dataset. Nobody swept them
and picked the best; they are what Perseus ships and what the field reports, and they
are used here for comparability with the literature. They are also load-bearing: push
the shift down and every imputed protein looks more depleted, pull it up and the
downshift stops meaning anything. That sensitivity is a reason to treat this arm as a
bracket on the MNAR extreme rather than as a point estimate to trust.

```{r}
#| label: perseus
perseus_dal <- perseus_impute(dal, shift = 1.8, width = 0.3, robust = TRUE, seed = 42)
perseus_mat <- as.matrix(perseus_dal$data)

c(na_remaining = sum(is.na(perseus_mat)), shift = 1.8, width = 0.3) # 0 remaining — complete matrix
```

## Result and verification {#sec-recap}

The runnable script saves the normalized matrix to `02_Normalization/c_data/`;
the four imputation arms save to `02_Normalization/imputation/c_data/`.

| File | Location | Missing values? | Used by |
|---|---|---|---|
| `DAList_normalized.rds` | `c_data/` | kept | the DE test (stage 03), primary |
| `DAList_imputed_missforest.rds` | `imputation/c_data/` | filled | figures, WGCNA (canonical) |
| `DAList_imputed_imp4p.rds` | `imputation/c_data/` | filled | the imputed-DEP comparison |
| `DAList_imputed_mscoreutils.rds` | `imputation/c_data/` | filled | the imputed-DEP comparison |
| `DAList_imputed_perseus.rds` | `imputation/c_data/` | filled | the imputed-DEP comparison |

The in-memory object matches the committed `DAList_normalized.rds`:

```{r}
#| label: verify
committed <- readRDS(here("02_Normalization", "c_data", "DAList_normalized.rds"))
c(
  same_dim = identical(dim(dal$data), dim(committed$data)),
  same_proteins = identical(sort(rownames(dal$data)), sort(rownames(committed$data))),
  same_samples = identical(sort(colnames(dal$data)), sort(colnames(committed$data)))
)
```

Normalization levels the samples without inventing any data, and imputation is
applied only where a method cannot run on an incomplete matrix. Stage 03 fits the
model on the non-imputed normalized matrix.

## Key terms

**DAList** — proteoDA's container [@thurman2023]. It carries the intensity matrix,
the protein annotation, the sample metadata, and a `tags` slot that records what has
been done to it, so an object states its own provenance.

**cycloess (cyclic loess)** — the normalization used here [@bolstad2003]. Under limma's
`method = "fast"`, which is what proteoDA passes, it regresses each sample against the
row-mean reference and pulls the samples onto a common scale, correcting
intensity-dependent bias rather than just shifting a median. The span is not the `0.7`
in limma's signature: `adaptive.span = TRUE` overrides it with `chooseLowessSpan(1920)`
= 0.5075. It rescales
observed values and fills nothing in.

**MAR (missing at random)** — the gap is unrelated to the protein's true abundance:
an identification failure, a lost peak. The gap carries no information about the value
it hides, so predicting it from other proteins is legitimate.

**MNAR (missing not at random)** — the protein is missing *because* it fell below the
detection limit. The gap itself carries information: it means low. Predicting it from
other proteins discards that and fills too high. DIA gaps are a mixture, skewed MNAR
[@lazar2016].

**imputation arm** — one of the four complete matrices produced by filling the gaps a
particular way. They are kept side by side because they encode different, untestable
assumptions about the MAR/MNAR mixture. missForest is canonical for the descriptive
figures and the module work; none of them feeds the *primary* DE test, though all four
feed the sensitivity comparison in `03_DEP/b_imputed/`.

**eta-squared (η²)** — between-group sum of squares over total, per protein: the share
of a protein's variance the six `Group_Time` cells explain. Upward-biased at small n,
with a null expectation near 0.11 here, and more biased for gappier proteins.

**omega-squared (ω²)** — the bias-corrected counterpart, which subtracts the variance
the grouping absorbs by chance and so sits near zero under the null. Used in the F02
supplement for exactly that reason.

**OOB error (out-of-bag)** — random forest's built-in held-out error estimate, scored
on the ~1/3 of observations each tree never saw during training. Reported by missForest
as error normalized by the observed variance, so 1 means "no better than guessing the
mean" and well below 1 means the other proteins genuinely predict the missing one.

## Recap

- Cyclic loess levels the samples and fills nothing in: the missing fraction is
  identical before and after.
- The eight-method comparison behind the cycloess choice lives in
  `b_reports/norm_comparison.pdf` and is *not* reproduced on this page — the choice is
  asserted here, and a reader who wants it demonstrated has to open the PDF.
- MAR and MNAR are different problems. Our gaps are skewed MNAR, which is why the four
  imputation arms disagree: missForest predicts gaps from neighbours, Perseus fills them
  from a downshifted Gaussian, and imp4p and the MsCoreUtils hybrid try to tell which is
  which.
- The parameters (`maxiter = 10`, `ntree = 100`, `shift = 1.8`, `width = 0.3`) are
  conventional defaults, not values tuned on this dataset.
- An imputed value is a prediction carrying no measurement error. Testing on a filled
  matrix shrinks standard errors and manufactures significance, so the DE test in stage
  03 runs on the non-imputed matrix and limma fits each protein where it was actually
  observed [@peng2024].
- The eta-squared scan is a screening description, not a test. Its null expectation is
  already ≈ 0.11, and gappier proteins are flattered more.

```{r sessioninfo}
sessionInfo()
```

## References

::: {#refs}
:::
