---
title: "HRvLR 02 — Normalization & Imputation"
subtitle: "Cyclic-loess normalization, the eta-squared scan, and when to fill missing values"
author: "A_HRvLR_2026"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
    embed-resources: true
    df-print: paged
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(here)
knitr::opts_knit$set(root.dir = here::here())
```

This document reproduces stage 02 of the HRvLR pipeline and shows the result of
each step. It runs the same logic as `02_Normalization/a_script/01_run_normalization.R`
(plus the three imputation arms in `02_Normalization/imputation/a_script/`), in
memory, writing no files. Normalization makes samples comparable; imputation is
applied only for the methods that need a complete matrix.

# Load the filtered `DAList` {#sec-load}

Stage 01 hands off a filtered, un-normalized object that still carries its
missing values:

```{r}
#| label: load
library(proteoDA)
dal <- readRDS(here("01_Filtering", "c_data", "DAList_filtered.rds"))
miss_before <- round(100 * mean(is.na(dal$data)), 1)
dim(dal$data)
miss_before # missing fraction, not yet imputed
dal$tags # provenance so far: the group filter
```

# Normalize with cyclic loess {#sec-normalize}

Samples load onto the instrument slightly differently, so whole samples sit
higher or lower for technical reasons. This shows in each sample's median
log-intensity. `normalize_data(norm_method = "cycloess")` evens it out with a
single native call (limma defaults: `span = 0.7`, `adaptive.span = FALSE`):

```{r}
#| label: normalize
before <- apply(log2(as.matrix(dal$data)), 2, median, na.rm = TRUE)
dal <- normalize_data(dal, norm_method = "cycloess")
after <- apply(as.matrix(dal$data), 2, median, na.rm = TRUE)

rbind(before = summary(before), after = summary(after)) # spread collapses to a tight band
dal$tags # new 'normalized' + 'norm_method' tags
```

Normalization **rescales** but does not fill anything in — the missingness is
unchanged:

```{r}
#| label: missing-unchanged
c(
  before_pct = miss_before,
  after_pct = round(100 * mean(is.na(dal$data)), 1)
)
```

The script writes proteoDA's audit PDFs to `b_reports/`. `write_norm_report()`
compares the candidate normalization methods on the filtered matrix **before** the
choice is committed, and `write_qc_report()` inspects the matrix before and after.
Those calls (shown here, run by the script) make the cycloess choice reviewable
without re-implementing the comparison by hand:

```{r}
#| label: qc-reports
#| eval: false
write_norm_report(dal,
  grouping_column = "Group_Time", output_dir = report_dir,
  filename = "norm_comparison.pdf"
) # compares candidate methods (before normalize)
write_qc_report(dal,
  color_column = "Group_Time", output_dir = report_dir,
  filename = "qc_post.pdf"
) # inspects the chosen result
```

This normalized, **non-imputed** object is the primary hand-off to stage 03 —
`limma` handles missing values per protein, so imputing before the test would
inflate false positives.

# The eta-squared scan {#sec-eta2}

For the figure layer, the script records how much of each protein's variance the
six `Group_Time` cells explain (eta-squared, the between-group sum of squares over
the total). This is a one-way decomposition per protein on the log-intensities,
and it feeds the F02 protein-variance panel:

```{r}
#| label: eta2
log_dat <- dal$data
grp_vec <- dal$metadata$Group_Time[match(colnames(log_dat), dal$metadata$Col_ID)]
eta2_vals <- apply(log_dat, 1, function(x) {
  ok <- !is.na(x)
  if (sum(ok) < 4) {
    return(NA_real_)
  }
  xk <- x[ok]
  gk <- grp_vec[ok]
  ss_b <- sum(tapply(xk, gk, length) * (tapply(xk, gk, mean) - mean(xk))^2)
  ss_t <- sum((xk - mean(xk))^2)
  if (ss_t > 0) ss_b / ss_t else NA_real_
})
summary(eta2_vals) # most proteins low; a tail is strongly group-structured
```

The script bundles `eta2_vals`, the post-normalization PCA, and the stage-01
intermediates into `00_report_intermediates.rds` for the reports and figures.

# Imputation for methods that need a complete matrix {#sec-impute}

PCA, clustering, WGCNA, and the rotation tests need a matrix with no gaps. Three
independent imputation methods are kept so the choice can be checked. Each
operates on the normalized matrix and fills every missing value. **missForest is
the canonical imputed arm** — the figures and WGCNA read it — with imp4p and the
MsCoreUtils hybrid kept as comparisons.

```{r}
#| label: impute-input
mat <- as.matrix(dal$data)
was_missing <- is.na(mat)
c(entries_imputed = sum(was_missing), pct_of_matrix = round(100 * mean(was_missing), 1))
```

## missForest (canonical) {#sec-missforest}

`missForest` is a non-parametric random-forest imputer: each protein's missing
values are predicted from the multivariate structure of the other proteins, so
the matrix is transposed to put proteins in the columns. It is stochastic, so the
seed is fixed and rows are ordered before the fit:

```{r}
#| label: missforest
#| cache: true
library(missForest)
set.seed(42)
ord <- order(rownames(mat))
mf <- missForest(t(mat[ord, ]), maxiter = 10, ntree = 100, verbose = FALSE)
imp <- t(mf$ximp)[rownames(mat), ]
dimnames(imp) <- dimnames(mat)

c(na_remaining = sum(is.na(imp)), OOB_error = round(mf$OOBerror[1], 4))
```

## imp4p (comparison) {#sec-imp4p}

`imp4p` is mechanism-aware: it estimates the per-protein mix of values missing at
random vs missing because the protein is too low to detect, then imputes each
accordingly:

```{r}
#| label: imp4p
#| cache: true
library(imp4p)
set.seed(42)
cond <- factor(dal$metadata$Group_Time[match(colnames(mat), dal$metadata$Col_ID)])
imp4p_mat <- as.matrix(impute.mi(
  tab = mat, conditions = cond,
  methodMCAR = "mle", methodMNAR = "igcda", progress.bar = FALSE
))
dimnames(imp4p_mat) <- dimnames(mat)
c(na_remaining = sum(is.na(imp4p_mat))) # 0 remaining — complete matrix
```

## MsCoreUtils hybrid (comparison) {#sec-mscoreutils}

The third method classifies each protein missing-at-random (MAR) vs not via
`imputeLCMD::model.Selector()`, then routes MAR features to kNN and MNAR features
to a left-censored QRILC draw:

```{r}
#| label: mscoreutils
library(MsCoreUtils)
library(imputeLCMD)
set.seed(42)
randna <- as.logical(model.Selector(mat)[[1]]) # TRUE = MAR feature
ms_mat <- impute_matrix(mat, method = "mixed", randna = randna, mar = "knn", mnar = "QRILC")

c(MAR_features = sum(randna), MNAR_features = sum(!randna), na_remaining = sum(is.na(ms_mat)))
```

# Where stage 02 stands {#sec-recap}

The normalized matrix is saved to `02_Normalization/c_data/`; the three imputation
arms write to `02_Normalization/imputation/c_data/`:

| File | Location | Missing values? | Used by |
|---|---|---|---|
| `DAList_normalized.rds` | `c_data/` | kept | the DE test (stage 03) — **primary** |
| `DAList_imputed_missforest.rds` | `imputation/c_data/` | filled | figures, WGCNA (canonical) |
| `DAList_imputed_imp4p.rds` | `imputation/c_data/` | filled | the imputed-DEP comparison |
| `DAList_imputed_mscoreutils.rds` | `imputation/c_data/` | filled | the imputed-DEP comparison |

```{r}
#| label: verify
committed <- readRDS(here("02_Normalization", "c_data", "DAList_normalized.rds"))
c(
  same_dim = identical(dim(dal$data), dim(committed$data)),
  same_proteins = identical(sort(rownames(dal$data)), sort(rownames(committed$data))),
  same_samples = identical(sort(colnames(dal$data)), sort(colnames(committed$data)))
)
```

Normalization levels the samples but never invents data; imputation is applied
only for methods that cannot run on an incomplete matrix, and never before the DE
test. Stage 03 fits the model on the non-imputed matrix.
