---
title: "01 · Design"
subtitle: "Design formula and the nine contrasts"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    df-print: kable
    fig-width: 7
    fig-height: 5
    fig-dpi: 150
    fig-align: center
execute:
  message: false
---

proteoDA carries the design as a formula on the DAList and reads its terms from metadata column
names. This notebook relabels the metadata, attaches the design and the nine contrasts, and
checks both before anything is fitted.

```{r setup}
library(here)
library(proteoDA)
library(limma)
library(dplyr)
library(tibble)
library(writexl)

dal <- readRDS(here("01_Preprocess", "02_Normalization", "c_data", "DAList_normalized.rds"))
source_meta <- as.data.frame(dal$metadata)
stopifnot(identical(source_meta$Col_ID, colnames(dal$data)))
with(source_meta, table(Group, Timepoint))
```

## Design

Six cell means, one per responder-and-timepoint combination, with subject as a random effect.
HR and LR are different people, so a fixed subject term would absorb the between-arm contrasts;
the random effect keeps them estimable and carries the repeated measures through
`duplicateCorrelation()`.

```{r design}
group_levels <- c("HR_T1", "HR_T2", "HR_T3", "LR_T1", "LR_T2", "LR_T3")
meta <- data.frame(
  sample_id = source_meta$Col_ID,
  responder = factor(source_meta$Group, levels = c("HR", "LR")),
  time = factor(source_meta$Timepoint, levels = c("T1", "T2", "T3")),
  group = factor(source_meta$Group_Time, levels = group_levels),
  subject = source_meta$Subject_ID
)
rownames(meta) <- meta$sample_id
stopifnot(!anyNA(meta$group), !any(is.na(meta$subject) | meta$subject == ""))
dal$metadata <- meta

dal <- add_design(dal, "~ 0 + group + (1 | subject)")
design <- dal$design$design_matrix

# Rank deficiency would make the contrasts inestimable and surface as silent NAs.
# add_contrasts() parses the contrast strings, so every column name must survive make.names().
stopifnot(
  qr(design)$rank == ncol(design),
  identical(make.names(colnames(design)), colnames(design))
)
c(samples = nrow(design), coefficients = ncol(design), residual_df = nrow(design) - ncol(design))
```

## Contrasts

```{r contrasts}
contrast_vector <- c(
  "Training_HR = HR_T2 - HR_T1",
  "Training_LR = LR_T2 - LR_T1",
  "Acute_HR = HR_T3 - HR_T2",
  "Acute_LR = LR_T3 - LR_T2",
  "Baseline_HRvLR = HR_T1 - LR_T1",
  "Trained_HRvLR = HR_T2 - LR_T2",
  "Acute_HRvLR = HR_T3 - LR_T3",
  "Training_Interaction = (HR_T2 - HR_T1) - (LR_T2 - LR_T1)",
  "Acute_Interaction = (HR_T3 - HR_T2) - (LR_T3 - LR_T2)"
)
dal <- add_contrasts(dal, contrasts_vector = contrast_vector)
contrast_names <- colnames(dal$design$contrast_matrix)
dal$design$contrast_matrix
```

```{r roles}
roles <- tibble(
  contrast = contrast_names,
  role = case_when(
    contrast == "Training_Interaction" ~ "primary",
    contrast == "Acute_Interaction" ~ "secondary",
    contrast == "Baseline_HRvLR" ~ "floor",
    TRUE ~ "descriptive"
  )
)
roles
```

`Training_Interaction` is the **primary** comparison: does the training response differ between
responder groups. `Acute_Interaction` asks the same of the acute bout. `Baseline_HRvLR` is the
**floor**. It is not a true negative control: the label was cut from the training outcome, so a
real baseline difference would be predictive. With no predictive signal it is empty, and it is
read as the count a between-arm contrast returns when nothing is there. The six others are
**descriptive**. `Training_Interaction` equals `Trained_HRvLR` minus `Baseline_HRvLR` exactly, so
these are not nine independent questions.

## Within-subject correlation

`duplicateCorrelation()` on the normalised matrix, blocked by subject. `fit_limma_model()`
estimates the same number internally and discards it.

```{r correlation}
strata <- tibble(
  level = "within subject, subject random (unweighted matrix correlation)",
  correlation = duplicateCorrelation(dal$data, design, block = meta$subject)$consensus.correlation
)
mutate(strata, correlation = round(correlation, 3))
```

## Outputs

```{r write}
out <- here("02_Differential_Expression", "01_Design", "c_data")
saveRDS(
  list(dal = dal, contrasts = contrast_vector, roles = roles, floor = "Baseline_HRvLR"),
  file.path(out, "design.rds")
)
write_xlsx(
  list(
    contrasts = tibble(definition = contrast_vector, contrast = contrast_names) |>
      left_join(roles, by = "contrast"),
    correlation_strata = strata
  ),
  file.path(out, "01_design.xlsx")
)
```

```{r session}
sessionInfo()
```
