---
title: "01 · Design"
subtitle: "Design formula and the nine contrasts"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    df-print: kable
    fig-width: 7
    fig-height: 5
    fig-dpi: 150
    fig-align: center
execute:
  message: false
---

proteoDA carries the design as a formula on the DAList and reads its terms from metadata column
names. This notebook sets the factor levels, attaches the design and the nine contrasts, and
checks both before anything is fitted.

```{r setup}
library(here)
library(proteoDA)
library(limma)
library(dplyr)
library(tibble)
library(writexl)

inputs <- c(normalized = "01_Preprocess/02_Normalization/c_data/DAList_normalized.rds")
paths <- vapply(inputs, here, character(1))
stopifnot(file.exists(paths))
dal <- readRDS(paths[["normalized"]])
stopifnot(identical(dal$metadata$sample_id, colnames(dal$data)))
with(dal$metadata, table(arm, timepoint))
```

## Six cell means, subject random

Six cell means, one per arm-and-timepoint combination, with subject as a random effect.
HR and LR are different people, so a fixed subject term would absorb the between-arm contrasts;
the random effect keeps them estimable and carries the repeated measures through
`duplicateCorrelation()`.

```{r design}
group_levels <- c("HR_T1", "HR_T2", "HR_T3", "LR_T1", "LR_T2", "LR_T3")
meta <- dal$metadata |>
  mutate(
    arm = factor(arm, levels = c("HR", "LR")),
    timepoint = factor(timepoint, levels = c("T1", "T2", "T3")),
    group = factor(group, levels = group_levels)
  )
rownames(meta) <- meta$sample_id
stopifnot(!anyNA(meta$group), !any(is.na(meta$subject) | meta$subject == ""))
dal$metadata <- meta

dal <- add_design(dal, "~ 0 + group + (1 | subject)")
design <- dal$design$design_matrix

# Rank deficiency would make the contrasts inestimable and surface as silent NAs.
# add_contrasts() parses the contrast strings, so every column name must survive make.names().
stopifnot(
  qr(design)$rank == ncol(design),
  identical(make.names(colnames(design)), colnames(design))
)
c(samples = nrow(design), coefficients = ncol(design), residual_df = nrow(design) - ncol(design))
```

## Nine contrasts, one primary

```{r contrasts}
contrast_vector <- c(
  "Training_HR = HR_T2 - HR_T1",
  "Training_LR = LR_T2 - LR_T1",
  "Acute_HR = HR_T3 - HR_T2",
  "Acute_LR = LR_T3 - LR_T2",
  "Baseline_HRvLR = HR_T1 - LR_T1",
  "Trained_HRvLR = HR_T2 - LR_T2",
  "Acute_HRvLR = HR_T3 - LR_T3",
  "Training_Interaction = (HR_T2 - HR_T1) - (LR_T2 - LR_T1)",
  "Acute_Interaction = (HR_T3 - HR_T2) - (LR_T3 - LR_T2)"
)
dal <- add_contrasts(dal, contrasts_vector = contrast_vector)
contrast_names <- colnames(dal$design$contrast_matrix)
dal$design$contrast_matrix
```

```{r roles}
roles <- tibble(
  contrast = contrast_names,
  role = case_when(
    contrast == "Training_Interaction" ~ "primary",
    contrast == "Acute_Interaction" ~ "secondary",
    contrast == "Baseline_HRvLR" ~ "floor",
    TRUE ~ "descriptive"
  )
)
roles
```

`Training_Interaction` is the primary comparison: does the training response differ between arms.
`Acute_Interaction` asks the same of the acute bout. `Baseline_HRvLR` is the floor. It is not a
negative control: the label was cut from the training outcome, so a real baseline difference would
be predictive. The six others are descriptive. `Training_Interaction` equals `Trained_HRvLR` minus
`Baseline_HRvLR` exactly.

## Within-subject correlation is 0.189

`duplicateCorrelation()` on the normalised matrix, blocked by subject. `fit_limma_model()`
estimates the same number internally.

```{r correlation}
strata <- tibble(
  level = "within subject, subject random (unweighted matrix correlation)",
  correlation = duplicateCorrelation(dal$data, design, block = meta$subject)$consensus.correlation
)
mutate(strata, correlation = round(correlation, 3))
```

## Writes design.rds and the workbook

```{r write}
out <- here("02_Differential_Expression", "01_Design", "c_data")
dir.create(out, recursive = TRUE, showWarnings = FALSE)
saveRDS(
  list(dal = dal, contrasts = contrast_vector, roles = roles, floor = "Baseline_HRvLR"),
  file.path(out, "design.rds"),
  compress = "xz"
)
sheets <- list(
  contrasts = tibble(definition = contrast_vector, contrast = contrast_names) |>
    left_join(roles, by = "contrast"),
  correlation_strata = strata,
  input_manifest = tibble(
    input = names(inputs), path = unname(inputs), md5 = unname(tools::md5sum(paths))
  ),
  package_versions = sessioninfo::package_info("loaded", dependencies = FALSE) |>
    as_tibble() |>
    select(package, version = loadedversion, source)
)
read_me <- tibble(sheet = names(sheets), holds = c(
  "The nine contrasts, their definitions and roles.",
  "Within-subject correlation from duplicateCorrelation, blocked by subject.",
  "Files read, with md5.",
  "Packages loaded at run time."
))
write_xlsx(c(list(read_me = read_me), sheets), file.path(out, "01_design.xlsx"))
```
