---
title: "F04 association — proteins, snapshot view"
subtitle: "Which single proteins track the training response"
author: "A_HRvLR_2026"
format:
  html:
    toc: true
    embed-resources: true
    df-print: paged
execute:
  eval: true
  echo: false
  warning: false
  message: false
---

## In plain terms

Boil each person down to one number — their protein level at baseline, or
how much it changed after training — and see which proteins line up with
who gained the most muscle.

## Method

Each protein is collapsed to one value per subject in each of three
windows — baseline level (T1), the training change (T2 − T1), and the
acute-bout change (T3 − T2) — and regressed on each continuous training
response (composite hypertrophy, fibre and whole-muscle cross-sectional
area, leg-press and extension 1RM) with `limma::lmFit` and `eBayes`
moderation. Every protein-by-trait-by-phase cell is a separate test, and
p-values are corrected with Benjamini-Hochberg across proteins within each
trait/phase family. This is a snapshot, descriptive, exploratory analysis
at n = 16 subjects: it says which proteins track a person's summary
response, not what happens between the biopsies.

## Results

```{r}
#| label: load-results
pacman::p_load(here, openxlsx, dplyr)

assoc <- read.xlsx(
  here(
    "04_Figures", "F04_association", "limma", "proteins", "c_data",
    "proteins_source_data.xlsx"
  ),
  "protein_association"
)

n_features <- n_distinct(assoc$feature)
n_traits <- n_distinct(assoc$trait)
n_phases <- n_distinct(assoc$phase)
n_tests <- nrow(assoc)
n_hits <- sum(assoc$bh < 0.05, na.rm = TRUE)
min_bh <- min(assoc$bh, na.rm = TRUE)
```

`r n_features` proteins were tested against `r n_traits` training-response
traits across `r n_phases` phases (baseline, training, acute), for
`r n_tests` protein-by-trait-by-phase cells in total. `r n_hits` of those
cells pass BH < 0.05; the smallest BH-adjusted q anywhere in the table is
`r sprintf("%.3f", min_bh)`. No single protein's snapshot value survives
correction for tracking any training response.

```{r sessioninfo}
sessionInfo()
```
